The first thing to say about the code below is there are many lines in
these bits of code that don’t do anything. These are the lines beginning
with the `#` character (read `#` as “hash”). Lines beginning with `#`
are called *comments*. When R sees the `#` at the start of the line, it
realizes that it should ignore everything else on that line, and skips
on the next line. Here’s an example of a comment in R:

```{r}
# R will completely ignore this text.
```

Because R ignores lines beginning with `#`, the text comments is just
for us, the humans that are reading the code. We often use comments in
code to explain what the other parts of the code are doing, to make the
rest of the code easier for humans to read and understand.

Our next task is to use R to simulate a single day of ambulances. We
will represent each ambulance by a random number from 1 through 10. 20
of these numbers represents a simulation of all 20 ambulances available
to the contractor. We will call a simulation of 20 ambulances — one
*trial*.



The core of the program to simulate one trial of ambulances problem
above begins with this command to the computer:

```{r}
# Generate 20 random integers from 1 through 10.
# We will explain the replace=TRUE later.
# Store the 20 numbers with the name "a"
a <- sample(1:10, 20, replace=TRUE)
```

This command orders the computer to randomly generate 20 numbers from
`1` through `10`. Inasmuch as each ambulance has a 1 in 10 chance of
being defective, we decide arbitrarily that a `1` stands for a defective
ambulance, and the other nine numbers (from `2` to `10`) stand for a
non-defective ambulance. The command orders the computer to store the
results of the random drawing in a location in the computer’s memory to
which we give a name such as “a” or “ambulances” or “aardvark” — the
name is up to us.

The next key element in the core of the program is:

```{r}
# Count the number of ones in the random numbers.
# The "a == 1" part identifies all the numbers equal to 1.
# The "sum" part counts how many numbers "a == 1" found.
b <- sum(a == 1)
```

This is a *counting* operation. The command orders the computer to
*count* the number of `1`s among the twenty numbers that are in location
`a` following the random drawing carried out by the `sample` operation.
The result of the `sum` operation will be somewhere between 0 and 20,
the number of ambulances that might be out-of-order on a given day. The
result is then placed in another location in the computer’s memory that
we label `b`.

Now let us place the `sample`and `r py_or_r('count_nonzero', 'sum')`
commands within the entire program that we use to solve this problem,
which is:

```{r}
# Make an array of 1000 numbers to store our 1000 counts.
# These all start off as 0, but we fill them in below as we do 1000 trials.
z <- numeric(1000)
# Repeat one trial of the simulation 1000 times.
for (i in 1:1000) {
    # Everything from the open curly bracket above - {
    # to the close curly bracket below - } is a *block*.
    # We run everything in this block, for each trial.
    # Therefore, R runs all these commands 1000 times.

    # Generate 20 numbers, each between 1 through 10 and put them in array a.
    # Each number will represent an ambulance, and we let 1 represent a
    # defective ambulance.
    a <- sample(1:10, 20, replace=TRUE)
    # Count the number of defective ambulances, and put the result in "b".
    b <- sum(a == 1)
    # Keep track of each trial's results in "counts".
    z[i] <- b
    # End this trial, then go back and repeat the process until all 1000 trials
    # are complete, then proceed.
}

# This part runs after all 1000 trials have finished.

# Determine how many trials resulted in more than 3 ambulances out of order.
k = sum(z > 3)

# Convert to a proportion.
kk = k / 1000

# Print the result.
print(kk)
#> [1] 0.136
```

The `z[i] <- b` statement that follows the `r
py_or_r('np.count_nonzero', 'sum')` *counting* operation simply keeps
track of the results of each trial, placing the number of defective
ambulances that occur in each trial in a location that we usually call
`z`. This is done in each of the 1000 trials that we make, and the
result eventually is an “array” with 1000 numbers in it.

In order to make 1000 repetitions of our experiment — we could have
decided to make ten thousand or some other number of repetitions — we
surround each trial with the `for` *loop*, that has the effect of
repeating the operations for one trial 1000 times. We identify the
operations to repeat with the 

opening — `{` — and closing — `}` — curly brackets.

Since our aim is to count the number of days in which more than 3 (4 or
more) defective ambulances occur, we use another *counting* command
(`sum`) at the end of the 1000 trials. This counting command *counts*
how many times more than 3 defects occurred in the 1000 days recorded in
our `z` array, and we place the result in still another location `k`.
This gives us the total number of days where 4 or more defective
ambulances are seen to occur. Then we divide the number in `k` by 1000,
the number of trials. Thus we obtain an estimate of the chance,
expressed as a probability between 0 and 1, that 4 or more ambulances
will be defective on a given day. And we store that result in a location
that we decide to call `kk`, so that it will be there when the computer
receives the next command to `print` that result on the screen.

Please don’t worry about the details of how each of these commands works
— we will cover those details gradually, over the next few chapters.
But, we hope that you can see in principle, how each of the operations
that the computer carries out are analogous to the operations that you
yourself executed when you solved this problem using a ten-sided spinner
or a random-number table. This is exactly the procedure that we will use
to solve every problem in probability and statistics that we must deal
with. Either we will use a device such as coins or a random number table
as an analogy for the physical process we are interested in (ambulances
becoming defective, in this case), or we will simulate the analogy on
the computer using some R code like the code above.

The R STATS program called “ambulances” may not seem simple to you at
first glance. But please believe us when we say that it is vastly
simpler than the older conventional approach to such problems that has
routinely been taught to students for decades.

## 3.4 How resampling differs from the conventional approach

In the standard approach the student learns to choose and solve a
formula. Doing the algebra and arithmetic is quick and easy. The
difficulty is in choosing the correct formula. Unless you are a
professional mathematician, it may take you quite a while to arrive at
the correct formula—considerable hard thinking, and perhaps some digging
in textbooks. More important than the labor, however, is that you may
come up with the wrong formula, and hence obtain the wrong answer. Most
students who have had a standard course in probability and statistics
are quick to tell you that it is not easy to find the correct formula,
even immediately after finishing a course (or several courses) on the
subject. After leaving school, it is harder still to choose the right
formula. Even many people who have taught statistics at the university
level (including this writer) must look at a book to get the correct
formula for a problem as simple as the ambulances, and then we are not
always sure of the right answer. This is the grave disadvantage of the
standard approach.

In the past few decades, resampling and other Monte Carlo simulation
methods have come to be used extensively in scientific research. But in
contrast to the material in this book, simulation has mostly been used
in situations so complex that mathematical methods have not yet been
developed to handle them. Here are examples of such situations:



1.  For a spaceship that will travel to Mars, calculating the correct
    flight route involves a great many variables, too many to solve with
    formulas. Hence, the Monte Carlo simulation method is used.

2.  The Navy might want to know how long the average ship will have to
    wait for dock facilities. The time of completion varies from ship to
    ship, and the number of ships waiting in line for dock work varies
    over time. This problem can be handled quite easily with the
    experimental simulation method, but formal mathematical analysis
    would be difficult or impossible.

3.  What are the best tactics in baseball? Should one bunt? Should one
    put the best hitter up first, or later? By trying out various
    tactics with dice or random numbers, Earnshaw Cook (in his book
    *Percentage Baseball* ), found that it is best never to bunt, and
    the highest-average hitter should be put up first, in contrast to
    usual practice. Finding this answer would have been much more
    difficult with the analytic method.

4.  Which search pattern will yield the best results for a ship
    searching for a school of fish? Trying out “models” of various
    search patterns with simulation can provide a fast answer.

5.  What strategy in the game of Monopoly will be most likely to win?
    The simulation method systematically plays many games (with a
    computer) testing various strategies to find the best one.

But those five examples are all complex problems. This book and its
earlier editions break new ground by using this method for *simple
rather than complex problems* , especially in statistics rather than
pure probability, and in teaching *beginning rather than advanced*
students to solve problems this way. (Here it is necessary to emphasize
that the resampling method is used to *solve the problems themselves
rather than as a demonstration device to teach the notions found in the
standard conventional approach* . Simulation has been used in elementary
courses in the past, but only to demonstrate the operation of the
analytical mathematical ideas. That is very different than using the
resampling approach to solve statistics problems themselves, as is done
here.)

Once we get rid of the formulas and tables, we can see that statistics
is a matter of *clear thinking, not fancy mathematics* . Then we can get
down to the business of learning how to do that clear statistical
thinking, and putting it to work for you. *The study of probability* is
purely mathematics (though not necessarily formulas) and technique. But
*statistics has to do with meaning* . For example, what is the meaning
of data showing an association just discovered between a type of
behavior and a disease? Of differences in the pay of men and women in
your firm? Issues of causation, acceptability of control, design of
experiments cannot be reduced to technique. This is “philosophy” in the
fullest sense. Probability and statistics calculations are just one
input. Resampling simulation enables us to get past issues of
mathematical technique and focus on the crucial statistical elements of
statistical problems.

If you intend to go on to advanced statistical work, the older standard
method can be learned alongside resampling methods. Your introduction to
the conventional method may thereby be made much more meaningful.



# 4 Resampling with code

Chapter <a href="#resampling-method">3</a> used simulation and
resampling from tables of random numbers, dice and coins. Making random
choices in this way can make it easier to understand the process, but of
course, physical methods of making random outcomes can be slow and
boring.

On the other hand, short computer programs can do a huge number of
resampling trials in a few seconds. The flexibility of a programming
language makes it possible to simulate many different outcomes and
tests.

In this chapter, we begin using the R language to build up tables of
random numbers, and do simple tasks like counting the number of values
in a row, and taking proportions.

With these simple tools, we can simulate many problems in probability
and statistics.

Here is our first R example program. Do not expect to follow all of it
straightaway. For now, read the code below to get an idea of how it
implements the procedure above. We will be coming back to the specifics
later.

The key to reading code is to think about what the computer will do,
when it sees the code.



A simple place to start is the *comment*. A comment is a statement that
the computer will *ignore*. It is text that we put in the program for
our benefit, to explain what is going on to a human reader.

This is an example of a comment:

```{r}
# This is a comment. It doesn't have any effect.
```

The comment starts with a hash character `#`. This character tells the
computer that the rest of the line is a comment, and therefore, that it
can ignore everything on that line that follows the `#`.



To solve the problem, we have to generate some random numbers:



```{r}
a = sample(1:10, 20, replace=TRUE)
```

This uses the `sample` function to generate random integers (counting
numbers) from 1 through 10. The 20 in the command tells R to generate 20
of these numbers. `replace=TRUE` tells R to sample *with replacement*.

For example, the chances of getting a particular result - such as a 3 -
for the first random number - are 1 in 10, or p= 1/10 = 0.1. If we
resample *with replacement* then the chances of getting 3 in the second
number are unchanged, at p=0.1. It is as if we put 10 balls into a
bucket, numbered one through ten, and then selected 20 balls from the
bucket; but after we have selected a ball, we record the number and
*replace* it in the bucket, and shake up the bucket again.

Inasmuch as each ambulance has a 1 in 10 chance of being defective, we
decide arbitrarily that a “1” stands for a defective ambulance, and the
other nine numbers (from “2” to “10”) stand for a not-defective
ambulance. The command above orders the computer to store the results of
the random drawing in a location in the computer’s memory to which we
give a name, in this case `a` (we could have chosen any name such as,
for example, `ambulances`). When we run a statement like the one above,
`a` is a *variable* - the *name* `a` refers to the *value*, which is the
sequence of random numbers the computer created using `sample`.

We can display the value of the variable `a` by using the `print`
function:

```{r}
print(a)
#>  [1] 7 5 6 4 6 9 5 5 8 2 7 5 5 2 6 4 2 3 9 6
```

This shows the 20 random values that we got from `sample`.

The next key element in the core of the program is:

```{r}
b = sum(a == 1)
```

This command can be broken up into two pieces: the first (`a == 1`)
compares each element in `a` to 1, resulting in a True or False value.
We then proceed to count those values, to determine how many of the 20
values in `a` are equal to 1. The result of the count will be somewhere
between 0 and 20, the number of ambulances that might be out-of-order on
a given day. The result is then placed in another location in the
computer’s memory that we label `b`.

```{r}
# Show the value of b
print(b)
#> [1] 0
```

Now let us place the commands to generate the random numbers, and count
how many `1`s we get, within a program that solves this problem:

```{r}
# Make an array that has 400 elements.
# We will use this to store the counts for our 400 repetitions
results <- numeric(400)

# Repeat the simulation 400 times
for (i in 1:400) {

    # The commands between the { above and the } below are the procedure for
    # one trial.
    # The computer runs these commands from first to last, for each trial.

    # Generate 20 numbers, each between "1" and "10," and put them in vector a.
    # Each number will represent an ambulance, and we let 1 represent
    # a defective ambulance.
    a <- sample(1:10, 20, replace=TRUE)

    # Count the number of defective ambulances, and put the result in b.
    b <- sum(a == 1)

    # Keep track of each trial's result in "results".
    results[i] <- b

    # End this trial, then go back and repeat the process until all 400 trials
    # are complete.
}

# Now we have finished the 400 trials.

# Determine how many trials resulted in more than 3 ambulances out of order.
bad_day_count <- sum(results > 3)

# Convert to a proportion.
bad_day_prop <- bad_day_count / 400

# Print the result.
print(bad_day_prop)
#> [1] 0.158
```
