# Cards and pennies


An answer for the following puzzle: “… shuffle a packet of four cards —
two red, two black — and deal them face down in a row. Two cards are
picked at random, say by placing a penny on each. What is the
probability that those two cards are the same color?”

```{r}
# Numbers representing the slips in the hat.
N <- c(1, 1, 2, 2)

# An array in which we will store the result of each trial.
z <- rep('No result yet', 10000)

for (i in 1:10000) {
    # sample, used in this way, has the effect of shuffling the vector
    # into a random order.  See the section linked above for an explanation.
    shuffled <- sample(N)

    A <- shuffled[1]  # The first slip from the shuffled array.
    B <- shuffled[2]  # The second slip from the shuffled array.

    # Set the result of this trial.
    if (A == B) {
        z[i] <- 'Yes'
    } else {
        z[i] <- 'No'
    }
}  # End of the loop.

# How many times did we see "Yes"?
k <- sum(z == 'Yes')

# The proportion.
kk <- k / 10000

message(kk)
```

Now let’s play the game differently, first picking one card and *putting
it back and shuffling* before picking a second card. What are the
results now? You can try it with the cards, but here is another program,
similar to the last, to run that variation.

```{r}
# An array in which we will store the result of each trial.
z <- rep('No result yet', 10000)

for (i in 1:10000) {
    # Shuffle the numbers in N into a random order.
    first_shuffle <- sample(N)
    # Draw a slip of paper.
    A <- first_shuffle[1]  # The first slip.

    # Shuffle again (with all the slips).
    second_shuffle <- sample(N)
    # Draw a slip of paper.
    B <- second_shuffle[1]  # The second slip.

    # Set the result of this trial.
    if (A == B) {
        z[i] <- 'Yes'
    } else {
        z[i] <- 'No'
    }
}  # End of the loop.

# How many times did we see "Yes"?
k <- sum(z == 'Yes')

# The proportion.
kk <- k / 10000

message(kk)
```
