# Fine day and win


This notebook calculates the chances that the Commanders win on a fine
day.

We also go through the logic of the `if` statement, and its associated
`else` clause.

```{r}
# blue means "nice day", yellow means "not nice".
bucket_A <- rep(c('blue', 'yellow'), c(7, 3))
bucket_A
```

Now let us draw a ball at random from bucket_A:

```{r}
a_ball <- sample(bucket_A, size=1)
a_ball
```

How we run our first `if` statement. Running this code will display “The
ball was blue” if the ball was blue, otherwise it will not display
anything:

```{r}
if (a_ball == 'blue') {
    message('The ball was blue')
}
```

Notice that the header line has `if`, followed by an open parenthesis
`(` introducing the *conditional expression* `a_ball == 'blue'`. There
follows close parenthesis `)` to finish the conditional expression. Next
there is a open curly brace `{` to signal the start of the body of the
`if` statement. The *body* of the `if` statement is one or more lines of
code, followed by the close curly brace `}`. Here there is only one
line: `message('The ball was blue')`. R only runs the body of the if
statement if the *condition* is `TRUE`.[^1]

To confirm we see “The ball was blue” if `a_ball` is `'blue'` and
nothing otherwise, we can set `a_ball` and re-run the code:

```{r}
# Set value of a_ball so we know what it is.
a_ball <- 'blue'
```

```{r}
if (a_ball == 'blue') {
    # The conditional statement is True in this case, so the body does run.
    message('The ball was blue')
}
```

```{r}
a_ball <- 'yellow'
```

```{r}
if (a_ball == 'blue') {
    # The conditional statement is False, so the body does not run.
    message('The ball was blue')
}
```

We can add an `else` clause to the `if` statement. Remember the *body*
of the `if` statement runs if the *conditional expression* (here
`a_ball == 'blue')` is `TRUE`. The `else` clause runs if the conditional
statement is `FALSE`. This may be clearer with an example:

```{r}
a_ball <- 'blue'
```

```{r}
if (a_ball == 'blue') {
    # The conditional expression is True in this case, so the body runs.
    message('The ball was blue')
} else {
    # The conditional expression was True, so the else clause does not run.
    message('The ball was not blue')
}
```

Notice that the `else` clause of the `if` statement starts with the end
of the `if` body with the closing curly brace `}`. `else` follows,
followed in turn by the opening curly brace `{` to start the body of the
`else` clause. The body of the `else` clause only runs if the initial
conditional expression is *not* `TRUE`.

```{r}
a_ball <- 'yellow'
```

```{r}
if (a_ball == 'yellow') {
    # The conditional expression was False, so the body does not run.
    message('The ball was blue')
} else {
    # but the else clause does run.
    message('The ball was not blue')
}
```

With this machinery, we can now implement the full logic of step 4
above:

    If you have drawn a blue ball from bucket A:
        Draw a ball from bucket B
        if the ball is green:
            record "yes"
        otherwise:
            record "no".

Here is bucket B. Remember green means “win” (65% of the time) and red
means “lose” (35% of the time). We could call this the “Commanders win
when it is a nice day” bucket:

```{r}
bucket_B <- rep(c('green', 'red'), c(65, 35))
```

The full logic for step 4 is:

Now we have everything we need to run many trials with the same logic.

```{r}
# By default, say we have no result.
result = 'No result'
a_ball <- sample(bucket_A, size=1)
# If you have drawn a blue ball from bucket A:
if (a_ball == 'blue') {
    # Draw a ball at random from bucket B
    b_ball <- sample(bucket_B, size=1)
    # if the ball is green:
    if (b_ball == 'green') {
        # record "yes"
        result <- 'yes'
    # otherwise:
    } else {
        # record "no".
        result <- 'no'
    }
}
# Show what we got in this case.
result
```

```{r}
# The result of each trial.
# To start with, say we have no result for all the trials.
z <- rep('No result', 10000)

# Repeat trial procedure 10000 times
for (i in 1:10000) {
    # draw one "ball" for the weather, store in "a_ball"
    # blue is "nice day", yellow is "not nice"
    a_ball <- sample(bucket_A, size=1)
    if (a_ball == 'blue') {  # nice day
        # if no rain, check on game outcome
        # green is "win" (give nice day), red is "lose" (given nice day).
        b_ball <- sample(bucket_B, size=1)
        if (b_ball == 'green') {  # Commanders win
            # Record result.
            z[i] <- 'yes'
        } else {
            z[i] <- 'no'
        }
    }
    # End of trial, go back to the beginning until done.
}

# Count of the number of times we got "yes".
k <- sum(z == 'yes')
# Show the proportion of *both* fine day *and* wins
kk <- k / 10000
kk
```

The above procedure gives us the probability that it will be a nice day
and the Commanders will win — about 46.1%.

[^1]: In this case, the result of the conditional expression is in fact
    either `TRUE` or `FALSE`. R is more liberal on what it allows in the
    conditional expression; it will take whatever the result is, and
    then force the result into either `TRUE` or `FALSE`, in fact, by
    wrapping the result with the `logical` function, that takes anything
    as input, and returns either `TRUE` or `FALSE`. Therefore, we could
    refer to the result of the conditional expression as something
    “truthy” — that is - something that comes back as `TRUE` or `FALSE`
    from the `logical` function. In the case here, that does not arise,
    because the result is in fact either exactly `TRUE` or exactly
    `FALSE`.
