# Five spades, four girls


This is a compound problem: what are the chances of *both* five or more
spades in one bridge hand, and four girls and a boy in a five-child
family?

“Compound” does not necessarily mean “complicated”. It means that the
problem is a compound of two or more simpler problems.

A natural way to handle such a compound problem is in stages, as we saw
in the archery problem of
<a href="#sec-one-black-archery" class="quarto-xref"><span
class="quarto-unresolved-ref">sec-one-black-archery</span></a>. If a
“success” is achieved in the first stage, go on to the second stage; if
not, don’t go on. More specifically in this example:

- **Step 1.** Use a bridge card deck, and five coins with heads =
  “girl”.
- **Step 2.** Deal a 13-card bridge hand and count the spades. If 5 or
  more spades, record “no” and end the experimental trial. Otherwise,
  continue to step 3.
- **Step 3.** Throw five coins, and count “heads.” If four heads, record
  “yes,” otherwise record “no.”
- **Step 4.** Repeat steps 2 and 3 a thousand times.
- **Step 5.** Compute the proportion of “yes” in step 3. This estimates
  the probability sought.

The R solution to this compound problem is neither long nor difficult.
We tackle it almost as if the two parts of the problem were to be dealt
with separately. We first determine, in a random bridge hand, whether 5
spades or more are dealt, as was done in the problem
<a href="#sec-five-spades-four-clubs" class="quarto-xref"><span
class="quarto-unresolved-ref">sec-five-spades-four-clubs</span></a>.
Then, `if` 5 or more spades are found, we use `sample` to generate a
random family of 5 children. This means that we need not generate
families if 5 or more spades were not dealt to the bridge hand, because
a “success” is only recorded if both conditions are met. After we record
the number of girls in each sample of 5 children, we need only finish
the loop (by `}` and then use `sum` to count the number of samples that
had 4 girls, storing the result in `k`. Since we only drew samples of
children for those trials in which a bridge hand of 5 spades had already
been dealt, `k` will have the number of trials out of 10000 in which
both conditions were met.

```{r}
N <- 10000
trial_results <- numeric(N)

# Deck with 13 spades and 39 other cards
deck <- rep(c('spade', 'others'), c(13, 52 - 13))

for (i in 1:N) {
    # Shuffle deck and draw 13 cards
    hand <- sample(deck, 13)  # replace=FALSE is default

    n_spades <- sum(hand == 'spade')

    if (n_spades >= 5) {
        # Generate a family, zeros for boys, ones for girls
        children <- sample(c('girl', 'boy'), 5, replace=TRUE)
        n_girls <- sum(children == 'girl')
        trial_results[i] <- n_girls
    }
}

k <- sum(trial_results == 4)

kk <- k / N

print(kk)
```

Here is an alternative approach to the same problem, but getting the
result at the end of the loop, by combining Boolean vectors (see
<a href="#sec-combine-booleans" class="quarto-xref"><span
class="quarto-unresolved-ref">sec-combine-booleans</span></a>).

```{r}
N <- 10000
trial_spades <- numeric(N)
trial_girls <- numeric(N)

# Deck with 13 spades and 39 other cards
deck <- rep(c('spade', 'other'), c(13, 39))

for (i in 1:N) {
    # Shuffle deck and draw 13 cards
    hand <- sample(deck, 13)  # replace=FALSE is default
    # Count and store the number of spades.
    n_spades <- sum(hand == 'spade')
    trial_spades[i] <- n_spades

    # Generate a family, zeros for boys, ones for girls
    children <- sample(c('girl', 'boy'), 5, replace=TRUE)
    # Count and store the number of girls.
    n_girls <- sum(children == 'girl')
    trial_girls[i] <- n_girls
}

k <- sum((trial_spades >= 5) & (trial_girls == 4))

kk <- k / N

# Show the result
message(kk)
```
