# Four girls and one boy


What is the probability of selecting four girls and one boy when
selecting five students from any group of twenty-five girls and
twenty-five boys?

```{r}
N <- 10000
trial_results <- numeric(N)

# Constitute the set of 25 girls and 25 boys.
whole_class <- rep(c('girl', 'boy'), c(25, 25))

# Repeat the following steps N times.
for (i in 1:N) {

    # Shuffle the numbers
    shuffled <- sample(whole_class)

    # Take the first 5 numbers, call them c.
    c <- shuffled[1:5]

    # Count how many girls there are, put the result in d.
    d <- sum(c == 'girl')

    # Keep track of each trial result in z.
    trial_results[i] <- d

    # End the experiment, go back and repeat until all 1000 trials are
    # complete.
}

# Count the number of times we got four girls, put the result in k.
k <- sum(trial_results == 4)

# Convert to a proportion.
kk <- k / N

# Print the result.
message(kk)
```

We can also find the probabilities of other outcomes from a histogram of
trial results obtained with the following command:

```{r}
# Do histogram, with one bin for each possible number.
hist(trial_results, breaks=0:max(trial_results), main='# of girls')
```

In the resulting histogram we can see that in 15 percent of the trials,
4 of the 5 selected were girls.

It should be noted that for this problem — as for most other problems —
there are several other resampling procedures that will also do the job
correctly.

In analytic probability theory this problem is worked with a formula for
“combinations.”
