# Four girls then one boy from 25/25


**In this problem, order matters; we are sampling without replacement,
with two outcomes, several of each item.**

What is the probability of getting an ordered series of *four girls and
then one boy* , from a universe of 25 girls and 25 boys? This
illustrates Case 3 above. Clearly we can use the same sampling mechanism
as in the example
<a href="#sec-four-girls-one-boy" class="quarto-xref"><span
class="quarto-unresolved-ref">sec-four-girls-one-boy</span></a>, but now
we record “yes” for a smaller number of composite events.

We record “no” even if a single one boy is chosen but he is chosen 1st,
2nd, 3rd, or 4th, whereas in
<a href="#sec-four-girls-one-boy" class="quarto-xref"><span
class="quarto-unresolved-ref">sec-four-girls-one-boy</span></a>, such
outcomes are recorded as “yes”-es.

- **Step 1.** Generate a class (vector) of length 50, consisting of 25
  strings valued “boy” and 25 strings valued “girl”.
- **Step 2.** Shuffle the class array, and select the first five
  elements.
- **Step 3.** If the first five elements are exactly
  `'girl', 'girl', 'girl',   'girl', 'boy'`, write “yes,” otherwise
  “no.”
- **Step 4.** Repeat steps 2 and 3, say, 10,000 times, and count the
  proportion of “yes” results, which estimates the probability sought.

Let us start the single trial procedure like so:

```{r}
# Constitute the set of 25 girls and 25 boys.
whole_class <- rep(c('girl', 'boy'), c(25, 25))

# Shuffle the class into a random order.
shuffled <- sample(whole_class)
# Take the first 5 class members, call them c.
c <- shuffled[1:5]
# Show the result.
c
```

Our next step (step 3) is to check whether `c` is exactly equal to the
result of interest. The result of interest is:

```{r}
# The result we are looking for - four girls and then a boy.
result_of_interest <- rep(c('girl', 'boy'), c(4, 1 ))
result_of_interest
```

We can then use a vector *comparison* with `==` to do an element by
element (*elementwise*) check, asking whether the corresponding elements
are equal:

```{r}
# A Boolean array, with True where corresponding elements are equal, False
# otherwise.
are_equal <- c == result_of_interest
are_equal
```

We are nearly finished with step 3 — it only remains to check whether
*all* of the elements were equal, by checking whether *all* of the
values in `are_equal` are `TRUE`.

We know that there are 5 elements, so we could check whether there are 5
`TRUE` values with `sum`:

```{r}
# Are there exactly 5 TRUE values in `are_equal`?
sum(are_equal) == 5
```

Another way to ask the same question is by using the `all` function on
`are_equal`. This returns `TRUE` if *all* the elements in `are_equal`
are `TRUE`, and `FALSE` otherwise.

<div class="callout callout-style-default callout-note callout-titled">

<div class="callout-header d-flex align-content-center">

<div class="callout-icon-container">

<i class='callout-icon'></i>

</div>

<div class="callout-title-container flex-fill">

Testing whether all elements of a vector are the same

</div>

</div>

<div class="callout-body-container callout-body">

The `all`, applied to a Boolean vector (as here), checks whether *all*
of the elements in the Boolean vector are `TRUE`. If so, it returns
`TRUE`, otherwise, it returns `FALSE`.

For example:

```{r}
# All elements are TRUE, `all` returns TRUE
all(c(TRUE, TRUE, TRUE, TRUE))
```

```{r}
# At least one element is FALSE, `all` returns FALSE
all(c(TRUE, TRUE, FALSE, TRUE))
```

</div>

</div>

Here is the full procedure for steps 2 and 3 (a single trial):

```{r}
# Shuffle the class into a random order.
shuffled <- sample(whole_class)
# Take the first 5 class members, call them c.
c <- shuffled[1:5]
# For each element, test whether the result is the result of interest.
are_equal <- c == result_of_interest
# Check whether we have the result we are looking for.
is_four_girls_then_one_boy <- all(are_equal)
```

All that remains is to put the single trial procedure into a loop.

```{r}
N <- 10000
trial_results <- numeric(N)

# Repeat the following steps 1000 times.
for (i in 1:N) {

    # Shuffle the class into a random order.
    shuffled <- sample(whole_class)
    # Take the first 5 class members, call them c.
    c <- shuffled[1:5]
    # For each element, test whether the result is the result of interest.
    are_equal <- c == result_of_interest
    # Check whether we have the result we are looking for.
    is_four_girls_then_one_boy <- all(are_equal)

    # Store the result of this trial.
    trial_results[i] <- is_four_girls_then_one_boy

    # End the experiment, go back and repeat until all N trials are
    # complete.
}

# Count the number of times we got four girls then a boy
k <- sum(trial_results)

# Convert to a proportion.
kk <- k / N

# Print the result.
message(kk)
```

This type of problem is conventionally done with a *permutation*
formula.
