# Fruit fly simulation


This notebook uses simulation to test the null hypothesis that it is
equally likely that new fruit files are male or female.

```{r}
# Set the number of trials
n_trials <- 10000

# set the sample size for each trial
sample_size <- 20

# An empty array to store the trials
scores <- numeric(n_trials)

# Do 1000 trials
for (i in 1:n_trials) {
    # Generate 20 simulated fruit flies, where each has an equal chance of being
    # male or female
    a <- sample(c('male', 'female'), size = sample_size, prob = c(0.5, 0.5),
    replace = TRUE)

    # count the number of males in the sample
    b <- sum(a == 'male')

    # store the result of this trial
    scores[i] <- b
}

# Produce a histogram of the trial results
title_of_plot <- paste0("Number of males in", n_trials, " samples of \n", sample_size, " simulated fruit flies")
hist(scores, xlab = 'Number of Males', main = title_of_plot)
```

In the histogram above, we see that in 16 percent of the trials, the
number of males was 14 or more, or 6 or fewer. Or instead of reading the
results from the histogram, we can calculate the result by tacking on
the following commands to the above program:

```{r}
# Determine the number of trials in which we had 14 or more males.
j <- sum(scores >= 14)

# Determine the number of trials in which we had 6 or fewer males.
k <- sum(scores <= 6)

# Add the two results together.
m <- j + k

# Convert to a proportion.
mm <- m/n_trials

# Print the results.
print(mm)
```
