# Another approach to ships with gold and silver


This notebook is a variation on the problem with gold and silver chests
in ships. It shows how we can count and tally the results at the end,
rather than in the trial itself.

Notice that the first part of the code is identical to the first
approach to this problem. There are two key differences — see the
comments for an explanation.

```{r}
# The 3 buckets, each representing two chests on a ship.
# As before.
bucket1 <- c('Gold', 'Gold')  # Chests in first ship.
bucket2 <- c('Gold',  'Silver')  # Chests in second ship.
bucket3 <- c('Silver', 'Silver')  # Chests in third ship.
```

```{r}
# Here is where the difference starts.  We are now going to fill in
# the result for the first chest _and_ the result for the second chest.
#
# Later we will fill in all these values, so the string we put here
# does not matter.

# Whether the first chest was Gold or Silver.
first_chests <- rep('To be announced', 10000)
second_chests <- rep('To be announced', 10000)

for (i in 1:10000) {
    # Select a ship at random from the three ships.
    # As before.
    ship_no <- sample(1:3, size=1)
    # Get the chests from this ship.
    # As before.
    if (ship_no == 1) {
        bucket <- bucket1
    }
    if (ship_no == 2) {
        bucket <- bucket2
    }
    if (ship_no == 3) {
        bucket <- bucket3
    }

    # As before.
    shuffled <- sample(bucket)

    # Here is the big difference - we store the result for the first and second
    # chests.
    first_chests[i] <- shuffled[1]
    second_chests[i] <- shuffled[2]
}  # End loop, go back to beginning.

# We will do the calculation we need in the next cell.  For now
# just display the first 10 values.
ten_first_chests <- first_chests[1:10]
message('The first 10 values of "first_chests:')
```

```{r}
print(ten_first_chests)
```

```{r}
ten_second_chests <- second_chests[1:10]
message('The first 10 values of "second_chests:')
```

```{r}
print(ten_second_chests)
```

In this variant, we recorded the type of first chest for each trial
(“Gold” or “Silver”), and the type of second chest of the second chest
(“Gold” or “Silver”).

**We would like to count the number of times there was “Gold” in the
first chest *and* “Gold” in the second.**

## 10.5 Combining Boolean arrays

We can do the count we need by *combining* the Boolean vectors with the
`&` operator. `&` combines Boolean vectors with a *logical and*.
*Logical and* is a rule for combining two Boolean values, where the rule
is: the result is `TRUE` if the first value is `TRUE` *and* the second
value if `TRUE`.

Here we use the `&` *operator* to combine some Boolean values on the
left and right of the operator:

Above you saw that the `==` operator (as in `== 'Gold'`), when applied
to vectors, asks the question of every element in the vector.

First make the Boolean vectors.

```{r}
ten_first_gold <- ten_first_chests == 'Gold'
message("Ten first == 'Gold'")
```

```{r}
print(ten_first_gold)
```

```{r}
ten_second_gold <- ten_second_chests == 'Gold'
message("Ten second == 'Gold'")
```

```{r}
print(ten_second_gold)
```

Now let us use `&` to combine Boolean vectors:

```{r}
ten_both <- (ten_first_gold & ten_second_gold)
ten_both
```

Notice that R does the comparison *elementwise* — element by element.

You saw that when we did `second_chests == 'Gold'` this had the effect
of asking the `== 'Gold'` question of *each element*, so there will be
one answer per element in `second_chests`. In that case there was a
vector to the *left* of `==` and a single value to the *right*. We were
comparing a vector to a value.

Here we are asking the `&` question of `ten_first_gold` and
`ten_second_gold`. Here there is a vector to the *left* and a vector to
the *right*. We are asking the `&` question 10 times, but the first
question we are asking is:

```{r}
# First question, giving first element of result.
(ten_first_gold[1] & ten_second_gold[1])
```

The second question is:

```{r}
# Second question, giving second element of result.
(ten_first_gold[2] & ten_second_gold[2])
```

and so on. We have ten elements on *each side*, and 10 answers, giving a
vector (`ten_both`) of 10 elements. Each element in `ten_both` is the
answer to the `&` question for the elements at the corresponding
positions in `ten_first_gold` and `ten_second_gold`.

We could also create the Boolean vectors and do the `&` operation all in
one step, like this:



Remember, we wanted the answer to the question: how many trials had
“Gold” in the first chest *and* “Gold” in the second. We can answer that
question for the first 10 trials with `sum`:

```{r}
n_ten_both <- sum(ten_both)
n_ten_both
```

We can answer the same question for *all* the trials, in the same way:

```{r}
first_gold <- first_chests == 'Gold'
second_gold <- second_chests == 'Gold'
n_both_gold <- sum(first_gold & second_gold)
n_both_gold
```

We could also do the same calculation all in one line:

```{r}
n_both_gold <- sum((first_chests == 'Gold') & (second_chests == 'Gold'))
n_both_gold
```

We can then count all the ships where the first chest was gold:

```{r}
n_first_gold <- sum(first_chests == 'Gold')
n_first_gold
```

The final calculation is the proportion of second chests that are gold,
given the first chest was also gold:

```{r}
p_g_given_g <- n_both_gold / n_first_gold
p_g_given_g
```

Of course we won’t get exactly the same results from the two
simulations, in the same way that we won’t get exactly the same results
from any two runs of the same simulation, because of the random values
we are using. But the logic for the two simulations are the same, and we
are doing many trials (10,000), so the results will be very similar.
