# Sampling tools


## 6.2 Samples and labels

Thus far we have used numbers such as 1 and 0 and 10 to represent the
elements we are sampling from. For example, in
<a href="#sec-resampling-two" class="quarto-xref"><span
class="quarto-unresolved-ref">sec-resampling-two</span></a>, we were
simulating the chance of a particular juror being black, given that 26%
of the eligible jurors in the county were black. We used *integers* for
that task, where we started with all the integers from 0 through 99, and
asked R to select values at random from those integers. When R selected
an integer from 0 through 25, we chose to label the resulting simulated
juror as black — there are 26 integers in the range 0 through 25, so
there is a 26% chance that any one integer will be in that range. If the
integer was from 26 through 99, the simulated juror was white (there are
74 integers in the range 26 through 99).

Here is the process of simulating a single juror, adapted from
<a href="#sec-random-zero-through-ninety-nine" class="quarto-xref"><span
class="quarto-unresolved-ref">sec-random-zero-through-ninety-nine</span></a>:

```{r}
# Get 1 random number from 0 through 99
# replace=TRUE is redundant here (why?), but we leave it for consistency.
a <- sample(0:99, 1, replace=TRUE)

# Show the result
a
```

After that, we have to unpack our labeling of 0 through 25 as being
“black” and 26 through 99 as being “white”. We might do that like this:

```{r}
this_juror_is_black <- a < 26
this_juror_is_black
```

This all works as we want it to, but it’s just a little bit difficult to
remember the coding (less than 26 means “black”, greater than 25 means
“white”). We had to use that coding because we committed ourselves to
using random numbers to simulate the outcomes.

However, R can also store bits of text, called *strings*. Values that
are bits of text can be very useful because the text values can be
memorable labels for the entities we are sampling from, in our
simulations.

## 6.3 String values

So far, all the values you have seen in R vectors have been numbers. Now
we get on to values that are bits of text. These are called *strings*.

Here is a single R string value:

```{r}
s <- "Resampling"
s
```

We can see what type of value `v` holds by using the `class` function.

For example, for a number value, you will usually find the `class` is
`numeric`:

```{r}
v <- 10
class(v)
```

What is the `class` of the new bit-of-text value `s`?

```{r}
class(s)
```

The R `character` value is a bit of text, and therefore consists of a
sequence of characters.

As vectors are containers for other things, such as numbers, strings are
containers for characters.

To get the length of a string, use the `nchar` function (Number of
Characters):

```{r}
# Number of characters in s
nchar(s)
```

R has a `substring` function that allows you to select individual
characters or sequences of characters from a string. The arguments to
`substring` are: first — the string; second — the index of the first
character you want to select; and third — the index of the last
character you want to select. For example to select the second character
in the string you would specify 2 as the starting index, and 2 as the
ending index, like this:

```{r}
# Get the second character of the string
second_char <- substring(s, 2, 2)
second_char
```

## 6.4 Strings in vectors

As we can store numbers as elements in vectors, we can also store
strings as vector elements.

```{r}
vector_of_strings = c('Julian', 'Lincoln', 'Simon')
vector_of_strings
```

As for any vector, you can select elements with *indexing*. When you
select an element with a given position (index), you get the *string* at
at that position:

```{r}
# Julian Lincoln Simon's second name
middle_name <- vector_of_strings[2]
middle_name
```

As for numbers, we can compare strings with, for example, the `==`
operator, that asks whether the two strings are equal:

```{r}
middle_name == 'Lincoln'
```

## 6.5 Repeating elements

Now let us go back to the problem of selecting black and white jurors.

We started with the strategy of using numbers 0 through 25 to mean
“black” jurors, and 26 through 99 to mean “white” jurors. We selected
values at random from 0 through 99, and then worked out whether the
number meant a “black” juror (was less than 26) or a “white” juror (was
greater than 25).

It would be good to use strings instead of numbers to identify the
potential jurors. Then we would not have to remember our coding of 0
through 25 and 26 through 99.

If only there was a way to make a vector of 100 strings, where 26 of the
strings were “black” and 74 were “white”. Then we could select randomly
from that array, and it would be immediately obvious that we had a
“black” or “white” juror.

Luckily, of course, we can do that, by using the `rep` function to
construct the vector.

Here is how that works:

```{r}
# The values that we will repeat to fill up the larger array.
juror_types <- c('black', 'white')
# The number of times we want to repeat "black" and "white".
repeat_nos <- c(26, 74)
# Repeat "black" 26 times and "white" 74 times.
jury_pool <- rep(juror_types, repeat_nos)
# Show the result
jury_pool
```

We can use this vector of repeats of strings, to sample from. The result
is easier to grasp, because we are using the string labels, instead of
numbers:

```{r}
# Select one juror at random from the black / white pool.
# replace=TRUE is redundant here, but we leave it for consistency.
one_juror <- sample(jury_pool, 1, replace=TRUE)
one_juror
```

We can select our full jury of 12 jurors, and see the results in a more
obvious form:

```{r}
# Select one juror at random from the black / white pool.
one_jury <- sample(jury_pool, 12, replace=TRUE)
one_jury
```

<div class="callout callout-style-default callout-note callout-titled">

<div class="callout-header d-flex align-content-center">

<div class="callout-icon-container">

<i class='callout-icon'></i>

</div>

<div class="callout-title-container flex-fill">

Using the `size` argument to `sample`

</div>

</div>

<div class="callout-body-container callout-body">

In the code above, we have specified the *size* of the sample we want
(12) with the second argument to `sample`. As you saw in
<a href="#sec-named-arguments" class="quarto-xref"><span
class="quarto-unresolved-ref">sec-named-arguments</span></a>, we can
also give names to the function arguments, in this case, to make it
clearer what we mean by “12” in the code above. In fact, from now on,
that is what we will do; we will specify the *size* of our sample by
using the *name* for the function argument to `sample` — `size` — like
this:

```{r}
# Select one juror at random from the black / white pool.
# Specify the sample size using the "size" named argument.
one_jury <- sample(jury_pool, size=12, replace=TRUE)
one_jury
```

</div>

</div>

We can use `==` on the vector to get `TRUE` values where the juror was
“black” and `FALSE` values otherwise:

```{r}
are_black <- one_jury == 'black'
are_black
```

Finally, we can `sum` to find the number of black jurors
(<a href="#sec-count-with-sum" class="quarto-xref"><span
class="quarto-unresolved-ref">sec-count-with-sum</span></a>):

```{r}
# Number of black jurors in this simulated jury.
n_black <- sum(are_black)
n_black
```

Putting that all together, this is our new procedure to select one jury
and count the number of black jurors:

```{r}
one_jury <- sample(jury_pool, size=12, replace=TRUE)
are_black <- one_jury == 'black'
n_black <- sum(are_black)
n_black
```

Or we can be even more compact by putting several statements together
into one line:

```{r}
# The same as above, but on one line.
n_black = sum(sample(jury_pool, size=12, replace=TRUE) == 'black')
n_black
```

## 6.6 Resampling with and without replacement

Now let us return to the details of Robert Swain’s case, that you first
saw in <a href="#sec-resampling-two" class="quarto-xref"><span
class="quarto-unresolved-ref">sec-resampling-two</span></a>.

We looked at the composition of Robert Swain’s 12-person jury — but in
fact, by law, that does not have to be representative of the eligible
jurors. The 12-person jury is drawn from a jury *panel*, of 100 people,
and this should, in turn, be drawn from the population of all eligible
jurors in the county, consisting, at the time, of “all male citizens in
the community over 21 who are reputed to be honest, intelligent men and
are esteemed for their integrity, good character and sound judgment.”
So, unless there was some bias against black jurors, we might expect the
100-person jury panel to be a plausibly random sample of the eligible
jurors, of whom 26% were black. See [the Supreme Court case
judgement](https://supreme.justia.com/cases/federal/us/380/202) for
details.

In fact, in Robert Swain’s trial, there were 8 black members in the
100-person jury panel. We will leave it to you to adapt the simulation
from <a href="#sec-resampling-two" class="quarto-xref"><span
class="quarto-unresolved-ref">sec-resampling-two</span></a> to ask the
question — is 8% surprising as a random sample from a population with
26% black people?

But we have a different question: given that 8 out of 100 of the jury
panel were black, is it surprising that none of the 12-person jury were
black? As usual, we can answer that question with simulation.

Let’s think about what a single simulated jury selection would look
like.

First we compile a representation of the actual jury panel, using the
tools we have used above.

```{r}
juror_types <- c('black', 'white')
# in fact there were 8 black jurors and 92 white jurors.
panel_nos <- c(8, 92)
jury_panel <- rep(juror_types, panel_nos)
# Show the result
jury_panel
```

Now consider taking a 12-person jury at random from this panel. We
select the first juror at random, so that juror has an 8 out of 100
chance of being black. But when we select the second jury member, the
situation has changed slightly. We can’t select the first juror again,
so our panel is now 99 people. If our first juror was black, then the
chances of selecting another black juror next are not 8 out of 100, but
7 out of 99 — a smaller chance. The problem is, as we shall see in more
detail later, the chances of getting a black juror as the second, and
third and fourth members of the jury depend on whether we selected a
black juror as the first and second and third jury members. At its most
extreme, imagine we had already selected eight jurors, and by some
strange chance, all eight were black. Now our chances of selecting a
black juror as the ninth juror are zero — there are no black jurors left
to select from the panel.

In this case we are selecting jurors from the panel *without
replacement*, meaning, that once we have selected a particular juror, we
cannot select them again, and we do not put them back into the panel
when we select our next juror.

This is the probability equivalent of the situation when you are dealing
a hand of cards. Let’s say someone is dealing you, and you only, a hand
of five cards. You get an ace as your first card. Your chances of
getting an ace as your first card were just the number of aces in the
deck divided by the number of cards — four in 52 – $\frac{4}{52}$. But
for your second card, the probability has changed, because there is one
less ace remaining in the pack, and one less card, so your chances of
getting an ace as your second card are now $\frac{3}{51}$. This is
sampling without replacement — in a normal game, you can’t get the same
card twice. Of course, you could imagine getting a hand where you
sampled *with replacement*. In that case, you’d get a card, you’d write
down what it was, and you’d give the card back to the dealer, who would
*replace* the card in the deck, shuffle again, and give you another
card.

As you can see, the chances change if you are sampling with or without
replacement, and the kind of sampling you do, will dictate how you model
your chances in your simulations.

Because this distinction is so common, and so important, the machinery
you have already seen in `sample` has simple ways for you to select your
sampling type. You have already seen sampling *with replacement*, and it
looks like this:

```{r}
# Take a sample of 12 jurors from the panel *with replacement*
strange_jury <- sample(jury_panel, size=12, replace=TRUE)
strange_jury
```

This is a strange jury, because it can select any member of the jury
pool more than once. Perhaps that juror would have to fill two (or
more!) seats, or run quickly between them. But of course, that is not
how juries are selected. They are selected *without replacement*:

Thus far, we have always done sampling *with replacement*, and, in order
to do that with `sample`, we pass the argument `replace=TRUE`. We do
that because the default for `sample` is `replace=FALSE`, that is, by
default, `sample` does sampling without replacement. If you want to do
sampling without replacement, you can just omit the `replace=TRUE`
argument to `sample`, or you can specify `replace=FALSE` explicitly,
perhaps to remind yourself that this is sampling without replacement.
Whether you omit the `replace` argument, or specify `replace=FALSE`, the
behavior is the same.

```{r}
# Take a sample of 12 jurors from the panel *with replacement*
# replace=FALSE is the default for sample.
ok_jury <- sample(jury_panel, size=12)
ok_jury
```

<div class="callout callout-style-default callout-note callout-titled">

<div class="callout-header d-flex align-content-center">

<div class="callout-icon-container">

<i class='callout-icon'></i>

</div>

<div class="callout-title-container flex-fill">

Comments at the end of lines

</div>

</div>

<div class="callout-body-container callout-body">

You have already seen comment lines. These are lines beginning with `#`,
to signal to R that the rest of the line is text for humans to read, but
R to ignore.

```{r}
# This is a comment.  R ignores this line.
```

You can also put comments at the *end of code lines*, by finishing the
code part of the line, and then putting a `#`, followed by more text.
Again, R will ignore everything after the `#` as a text for humans, but
not for R.

```{r}
message('Hello')  # This is a comment at the end of the line.
```

</div>

</div>

To finish the procedure for simulating a single jury selection, we count
the number of black jurors:

```{r}
n_black <- sum(ok_jury == 'black')  # How many black jurors?
n_black
```

Now we have the procedure for one simulated trial, here is the procedure
for 10000 simulated trials.

```{r}
counts <- numeric(10000)
for (i in 1:10000) {
    # Single trial procedure
    jury <- sample(jury_panel, size=12)  # replace=FALSE is the default.
    n_black <- sum(jury == 'black')  # How many black jurors?
    # Store the result
    counts[i] <- n_black
}
# Number of juries with 0 black jurors.
zero_black <- sum(counts == 0)
# Proportion
p_zero_black <- zero_black / 10000
message(p_zero_black)
```

We have found that, when there are only 8% black jurors in the jury
panel, having no black jurors in the final jury happens about 34% of the
time, even in this case, where the jury is selected completely at random
from the jury panel.

We should look for the main source of bias in the initial selection of
the jury panel, not in the selection of the jury from the panel.

