---
jupyter:
  jupytext:
    metadata_filter:
      notebook:
        additional: all
        excluded:
        - language_info
    text_representation:
      extension: .Rmd
      format_name: rmarkdown
      format_version: '1.0'
      jupytext_version: 0.8.6
  kernelspec:
    display_name: Python 3
    language: python
    name: python3
resampling_with:
    ed2_fname: null
---

```{r setup, include=FALSE}
source("_common.R")
```

# Introducing {{< var lang >}} and the Jupyter notebook {#sec-about-technology}

This chapter introduces you to the technology we will use throughout the book.
By *technology*, we mean two things:

* The {{< var lang >}} programming language, along with some important add-on
  libraries for data analysis.
* The {{< var nb_app >}} notebook system for running and editing
  {{< var lang >}} code in a web interface.

:::{.callout-note}
## Using {{< var lang >}} on the web and on your computer

In this chapter, we concentrate on the technology we use for the interactive
version of the book.  The interactive version allows you to run {{< var lang
>}} code as interactive *notebooks* in your web browser.

Either now, or later, you should also consider running the code on your own
computer — see @sec-running-own-computer.

:::



The chapter introduces [the R language]{.r}[Python and its
libraries]{.python}, and then gives an example to introduce {{< var lang >}}
and the {{< var nb_app >}} Notebook.  If you have not used {{< var lang >}}
before, the example notebook will get you started.  The example also shows how
we will be using notebooks through the rest of the book.

::: python
## Python and its packages

This version of the book uses the *Python* [^python-lang] programming language
to implement resampling algorithms.

Python is a programming language that can be used for many tasks.  It is a
popular language for teaching, but is also used widely in industry and
academia.  It is one of the most widely used programming languages in the
world, and the most popular language for data science.

For many of the initial examples, we will also be using the *NumPy* [^numpy]
package for Python. A *package* is a library of Python code and data.  NumPy
is a package that makes it easier to work with sequences of data values, such
as sequences of numbers.  These are typical in probability and statistics.

Later, we be using the *Matplotlib* [^matplotlib] package. This is the main
Python package with code for producing plots, such as bar charts, histograms,
and scatter plots.  See the rest of the book for more details on these plots.

Still further on in the book, we will use more specialized libraries for data
manipulation and analysis.  *Pandas* [^pandas] is the standard Python package
for loading data files and working with data tables.  *SciPy* [^scipy] is a
package that houses a wide range of numerical routines, including some simple
statistical methods. The *Statsmodels* [^statsmodels] package has code for
many more statistical procedures. We will often find ourselves comparing the
results of our own resampling algorithms to those in SciPy and Statsmodels.
:::

::: r
This version of the book uses the *R* programming language to implement
resampling algorithms.

The current title of the main website for R [^r-lang] is "The R Project for
Statistical Computing", and this is a good summary of how R started, and what R
is particularly good at.  The people who designed R designed it for themselves,
and you and I — those of us who are working to find and understand the patterns
in data.  Over the last 20 years, it has gained wide use for data analysis
across many fields, especially in life sciences, data science and statistics.

[^r-lang]: <https://www.r-project.org/>

Although many people use R as a simple way of exploring data and doing standard
statistical tests, it is a full-fledged programming language.
:::

It is very important that {{< var lang >}} is a *programming language* and not
a set of canned routines for "doing statistics".  It means that we can
explore the *ideas* of probability and statistics using the *language* of {{<
var lang >}} to express those ideas.  It also means that you, and we, and
anyone else in the world, can write new code to share with others, so they can
benefit from our work, understand it, and improve it.  This book is one
example; we have written the {{< var lang >}} code in this book as clearly as
we can to make it easy to follow, and to explain the underlying ideas.  We hope
you will help us by testing what we have done and sending us suggestions for
ways we could improve. Please see the preface for more information about how to
do that.

## The environment

Many of the chapters have sections with code for you to run, and experiment
with. These sections contain *Jupyter notebooks* ^[jupyter-nb]].  Jupyter
notebooks are interactive web pages that allow you to read, write and run {{<
var lang >}} code. We mark the start of each notebook in the text with a note
and link heading like the one you see below. In the web edition of this book,
you can click on the *Download* link in this header to download the section as
a notebook. You can also click on the *Interact* link in this header to open
the notebook on a cloud computer.  This allows you to interact with the
notebook on the cloud computer.  You can run the code, and experiment by making
changes.

In the print version of the book, we point you to the web version, to get the
links.

At the end of this chapter, we explain how to run these notebooks on your own
computer.  In the next section you will see an example notebook; you might want
to run this in the cloud to get started.

## Getting started with the notebook

The next section contains a notebook called "Billie's Bill".  If you are
looking at the web edition, you will see links to interact with this notebook
in the cloud, or download it to your computer.

::: {.notebook name="billies_bill" title="Billie's Bill"}

::: python
The text in this notebook section assumes you have opened the page as an
interactive notebook on the web, or on your own computer (see @sec-running-own-computer).
:::

::: r
The text in this notebook section assumes you have opened the page as an
interactive notebook on the web.  In that case, you are running the notebook in
a version of Jupyter.   We will assume you are using Jupyter in the description
that follows.  You can also run R in Jupyter on your own computer, although we
recommend that you use RStudio to work with R notebook files (see
@sec-running-own-computer).  The procedure for working with notebooks in RStudio is not the same as that for Jupyter.  We will cover RStudio, briefly, in another notebook tutorial after this one.
:::

A notebook can contain blocks of text — like this one — as well as code, and
the results from running the code.

Jupyter Notebooks are made up of *cells*.

::: r
We assume you are running this notebook via the interactive web pages, and
therefore, that you are running the notebook with Jupyter, not RStudio.

RStudio differs a little from Jupyter, because it does not have the idea of a text "cell" — instead it distinguishes between the main body of the notebook, made up of text, and *code chunks* — delimited sections in the notebook that contain code, instead of text.  Jupyter would refer to these code chunks as *code cells*.  We will go into more detail in @sec-running-rstudio.
:::

Jupyter cells can contain text or code.

::: nb-only
This is a cell with text — a text cell.
:::

Notebook text can have formatting, such as links.

For example, this sentence ends with a link to the earlier [second edition of
this book](https://resample.statistics.com/intro-text-online).

If you are in the interactive notebook interface (rather than reading this in
the textbook), you will see the Jupyter menu bar near the top of the page, with
headings "File", "Edit" and so on.

In Jupyter, underneath the File ... menu bar, by default, you may see a row of
icons - the "Toolbar".

In the Jupyter toolbar, you may see icons to run the current cell, among
others.

To move from one cell to the next, you can click the run icon in the toolbar, but it is more efficient to press the Shift key, and press Enter (with Shift still held down).  We will write this as {{< var run_key >}}.

In this, our first notebook, we will be using {{< var lang >}} to solve one of
those difficult and troubling problems in life — working out the bill in a
restaurant.


## The meal in question


Alex and Billie are at a restaurant, getting ready to order. They do not have
much money, so they are calculating the expected bill before they order.

Alex is thinking of having the fish for £10.50, and Billie is leaning towards
the chicken, at £9.25.  First they calculate their combined bill.

Below this text you see a *code* {{< var cell >}}.  It contains the {{< var
lang >}} code to calculate the total bill. Press
[Control-Shift-Enter or Cmd-Shift-Enter (on Mac)]{.r}[Shift-Enter]{.python}
in the {{< var cell >}} below, to see the total.[ (Remember, this is what you should do when running the code in Jupyter, via the interactive web pages.  If you are using RStudio on your own computer, you should use different commands — see @sec-running-rstudio.)]{.r}.

```{python}
10.50 + 9.25
```

```{r}
10.50 + 9.25
```

The contents of the {{< var cell >}} above is {{< var lang >}} code.  As you
would predict, {{< var lang >}} understands numbers like `10.50`, and it
understands `+` between the numbers as an instruction to add the numbers.

When you press {{< var run_key >}}, {{< var lang >}} finds `10.50`, realizes
it is a number, and stores that number somewhere in memory. It does the same
thing for `9.25`, and then it runs the *addition* operation on these two numbers in memory, which gives the number 19.75.

Finally, {{< var lang >}} sends the resulting number (19.75) back to the
notebook for display.  The notebook detects that {{< var lang >}} sent back a
value, and shows it to us.

This is exactly what a calculator would do.


## Comments

Unlike a calculator, we can also put notes next to our calculations, to remind
us what they are for.   One way of doing this is to use a "comment".  You have
already seen comments in the previous chapter.

A comment is some text that the computer will ignore.  In {{< var lang >}},
you can make a comment by starting a line with the `#` (hash) character.  For
example, the next cell is a code cell, but when you run it, it does not show
any result.  In this case, that is because the computer sees the `#` at the
beginning of the line, and then ignores the rest.

```{python}
# This bit of text is for us to read, and the computer to ignore.
```

```{r}
# This bit of text is for us to read, and the computer to ignore.
```

Many of the code cells you see will have comments in them, to explain what the
code is doing.

Practice writing comments for your own code.  It is a very good habit to get
into.  You will find that experienced programmers write many comments on their
code.  They do not do this to show off, but because they have a lot of
experience in reading code, and they know that comments make it much easier to
read and understand code.


## More calculations

Let us continue with the struggle that Alex and Billie are having with their
bill.

They realize that they will also need to pay a tip.

They think it would be reasonable to leave a 15% tip.  Now they need to
multiply their total bill by 0.15, to get the tip.  The bill is about £20, so
they know that the tip will be about £3.

In {{< var lang >}} `*` means multiplication.  This is the equivalent of the
"×" key on a calculator.

What about this, for the correct calculation?

```{python}
# The tip - with a nasty mistake.
10.50 + 9.25 * 0.15
```

```{r}
# The tip - with a nasty mistake.
10.50 + 9.25 * 0.15
```

Oh dear, no, that isn't doing the right calculation.

{{< var lang >}} follows the normal rules of *precedence* with calculations.
These rules tell us to do multiplication before addition.

See <https://en.wikipedia.org/wiki/Order_of_operations> for more detail on the standard rules.

In the case above the rules tell {{< var lang >}} to first calculate `9.25 *
0.15` (to get `1.3875`) and then to add the result to `10.50`, giving
`11.8875`.

We need to tell {{< var lang >}} we want it to do the *addition* and *then*
the multiplication.  We do this with round brackets (parentheses):

:::{#nte-bracket-types .callout-note}
## Three types of brackets in {{< var lang >}}

There are three types of brackets in {{< var lang >}}.

These are:

* *round brackets* or *parentheses*: `()`;
* *square brackets*: `[]`;
* *curly brackets*: `{}`.

Each type of bracket has a different meaning in {{< var lang >}}.  In the
examples, play close to attention to the type of brackets we are using.
:::

```{python}
# The bill plus tip - mistake fixed.
(10.50 + 9.25) * 0.15
```

```{r}
# The bill plus tip - mistake fixed.
(10.50 + 9.25) * 0.15
```

The obvious next step is to calculate the bill *including the tip*.

```{python}
# The bill, including the tip
10.50 + 9.25 + (10.50 + 9.25) * 0.15
```

```{r}
# The bill, including the tip
10.50 + 9.25 + (10.50 + 9.25) * 0.15
```

At this stage we start to feel that we are doing too much typing.  Notice that
we had to type out `10.50 + 9.25` twice there.  That is a little boring, but it
also makes it easier to make mistakes. The more we have to type, the greater
the chance we have to make a mistake.

## Variables {#sec-variables}

To make things simpler, we would like to be able to *store* the result of the
calculation `10.50 + 9.25`, and then re-use this value, to calculate the tip.

This is the role of *variables*.  A *variable* is a value with a name.

Here is a variable:

```{python}
# The cost of Alex's meal.
a = 10.50
```

```{r}
# The cost of Alex's meal.
a <- 10.50
```

`a` is a *name* we give to the value 10.50.  You can read the line above as
"The variable `a` *gets the value* 10.50".  We can also talk of *setting* the
variable.  Here we are *setting* `a` to equal 10.50.

Now, when we use `a` in code, it refers to the value we gave it.  For example,
we can put `a` on a line on its own, and {{< var lang >}} will show us the
*value* of `a`:

```{python}
# The value of a
a
```

```{r}
# The value of a
a
```

We did not have to use the name `a` — we can choose almost any name we like.
For example, we could have chosen `alex_meal` instead:

```{python}
# The cost of Alex's meal.
# alex_meal gets the value 10.50
alex_meal = 10.50
```

```{r}
# The cost of Alex's meal.
# alex_meal gets the value 10.50
alex_meal <- 10.50
```

We often set variables like this, and then display the result, all in the same
{{< var cell >}}.  We do this by first setting the variable, as above, and
then, on the final line of the {{< var cell >}}, we put the variable name on a
line on its own, to ask {{< var lang >}} to show us the value of the variable.
Here we set `billie_meal` to have the value 9.25, and then show the value of
`billie_meal`, all in the same {{< var cell >}}.

```{python}
# The cost of Billie's meal.
billie_meal = 9.25
# Show the value of billies_meal
billie_meal
```

```{r}
# The cost of Alex's meal.
# billie_meal gets the value 10.50
billie_meal <- 10.50
# Show the value of billie_meal
billie_meal
```

Of course, here, we did not learn much, but we often set variable values with
the results of a calculation.  For example:

```{python}
# The cost of both meals, before tip.
bill_before_tip = 10.50 + 9.25
# Show the value of both meals.
bill_before_tip
```

```{r}
# The cost of both meals, before tip.
bill_before_tip <- 10.50 + 9.25
# Show the value of both meals.
bill_before_tip
```

But wait — we can do better than typing in the calculation like this.  We can use the values of our variables, instead of typing in the values again.

```{python}
# The cost of both meals, before tip, using variables.
bill_before_tip = alex_meal + billie_meal
# Show the value of both meals.
bill_before_tip
```

```{r}
# The cost of both meals, before tip, using variables.
bill_before_tip <- alex_meal + billie_meal
# Show the value of both meals.
bill_before_tip
```

We make the calculation clearer by writing the calculation this way — we are
calculating the bill before the tip by adding the cost of Alex's and Billie's
meal — and that's what the code looks like.  But this also allows us to
*change* the variable value, and recalculate.  For example, say Alex decided
to go for the hummus plate, at £7.75.  Now we can tell {{< var lang
>}} that we want `alex_meal` to have the value 7.75 instead of 10.50:

```{python}
# The new cost of Alex's meal.
# alex_meal gets the value 7.75
alex_meal = 7.75
# Show the value of alex_meal
alex_meal
```

```{r}
# The new cost of Alex's meal.
# alex_meal gets the value 7.75
alex_meal = 7.75
# Show the value of alex_meal
alex_meal
```

Notice that `alex_meal` now has a new value.  It was 10.50, but now it is 7.75.
We have *reset* the value of `alex_meal`.   In order to use the new value for
`alex_meal`, we must *recalculate* the bill before tip with *exactly the same
code as before*:

```{python}
# The new cost of both meals, before tip.
bill_before_tip = alex_meal + billie_meal
# Show the value of both meals.
bill_before_tip
```

```{r}
# The new cost of both meals, before tip.
bill_before_tip <- alex_meal + billie_meal
# Show the value of both meals.
bill_before_tip
```

Notice that, now we have rerun this calculation, we have *reset* the value for `bill_before_tip` to the correct value corresponding to the new value for `alex_meal`.

All that remains is to recalculate the bill plus tip, using the new value for
the variable:

```{python}
# The cost of both meals, after tip.
bill_after_tip = bill_before_tip + bill_before_tip * 0.15
# Show the value of both meals, after tip.
bill_after_tip
```

```{r}
# The cost of both meals, after tip.
bill_after_tip = bill_before_tip + bill_before_tip * 0.15
# Show the value of both meals, after tip.
bill_after_tip
```

Now we are using variables with relevant names, the calculation looks right to
our eye.  The code expresses the calculation as we mean it: the bill after tip
is equal to the bill before the tip, plus the bill before the tip times 0.15.

## And so, on

Now you have done some practice with the notebook, and with variables, you are
ready for a new problem in probability and statistics, in the next chapter.

:::
<!---
End of notebook.
-->

## Running the code on your own computer {#sec-running-own-computer}

Many people, including your humble authors, like to be able to run code
examples on their own computers.  This section explains how you can set up to
run the notebooks on your own computer.

Once you have done this setup, you can use the "download" link that you will see for each notebook, to download the notebook to your machine.  From there, you can open the notebook on {{< var nb_app >}}.

:::{#nte-basic-download-links .callout-note}
## Download links and data files

Most of the download links in this book will trigger a download of the notebook
file.  This is a file with extension [`.Rmd`]{.r}[`.ipynb`]{.python}, that you
can open with {{< var nb_app >}}.

Later in the book, you will see examples where the notebook loads a data file.
In that case, the download link for the notebook points to a `.zip` file
containing the notebook and the data file.  Unzip the `.zip` file to get the
notebook and data file, and then open the resulting notebook in {{< var nb_app
>}}.
:::

::: python
You will need to install the Python language on your computer, and then install the following packages:

* NumPy
* Matplotlib - for plots
* SciPy - a collection of modules for scientific computing;
* Pandas - for loading, saving and manipulating data tables;
* Statsmodels - for traditional statistical analysis.
* Jupyter - to run the Jupyter Notebook on your own computer.

One way to install Python and the packages you need, is to install Python from
the Python website [^python-lang]. Then use the *Pip* [^pip] installer to
install the packages you need.

To install the Python packages, first start a terminal application on your
computer. To do this, you can use the Start key, "cmd" in Windows, or the
Command key and space then "Terminal" on Mac.  At the terminal prompt, type the
following command:

```{bash, eval=FALSE}
pip install numpy matplotlib scipy pandas statsmodels jupyter
```

Now you should be able to start the Jupyter notebook application.  See the
[Jupyter documentation for how to start
Jupyter](https://jupyter-notebook.readthedocs.io/en/stable/notebook.html#starting-the-notebook-server).
Open the notebook you downloaded for the chapter; you will now be able to run
the code on your own computer, and experiment by making changes.
:::

You can run any of the code notebooks in this textbook on your own machine by
downloading the notebook, via the download link at the top of each notebook
section, and then opening the resulting notebook in {{< var nb_app >}}.

::: r

### Running notebooks in RStudio {#sec-running-rstudio}

We are use a version of the Jupyter notebook for our interactive (web)
examples.  Although you can use the Jupyter notebook on your own computer, we
recommend you use the RStudio desktop application.[^rstudio] It works in
similar way to Jupyter, but with some variation.  You will find many good
tutorials for RStudio online; at time of writing [this DataCamp
tutorial](https://www.datacamp.com/tutorial/r-studio-tutorial) was a good place
to start.

As you will read on the RStudio download page, linked above, you will need two
software packages on your computer to run RStudio.  These are:

* The base R language.
* The RStudio graphical interface to R.

The base R language gives you the software to run R code and show results. You
can use the base R language on its own, but RStudio gives you a richer
interface to interact with the R language, including the ability to open, edit
and run R notebooks, like the notebook in this chapter.  RStudio uses the base
R language to run R code from the notebook, and show the results.

Install the base R language by going to the main R website at
<https://www.r-project.org>, following the links to the package for your system
(Windows, Mac, or Linux), and install according to the instructions on the
website.

Then install the RStudio interface by going to the RStudio download page at
<https://posit.co/download/rstudio-desktop/>, and navigating to the link to
download RStudio Desktop.  RStudio is an Integrated Development Environment for
R.  The RStudio IDE makes it easier to interact with, and develop, R code. You
only need the free (default) version; it has all the features you will need.
The free version is the only version that we, your humble authors, have used
for this book, and for all our own work and teaching.

See @sec-running-rstudio for a starter tutorial on running notebooks in
RStudio.

Click the Download link below to download this notebook to your computer.

::: {.notebook name="bill_in_rstudio" title="The Bill but in RStudio"}

Notice that the Download link gave you a file ending in `.Rmd`, where "Rmd" is
short for *RMarkdown*.  *RMarkdown* is the name of the notebook format — and it is the native notebook format for RStudio.  The `.Rmd` file is an RMarkdown version of the notebook.

Open the `bill_in_rstudio.Rmd` file in RStudio.

RStudio's idea of a notebook is very similar to Jupyter's — and this is not
a coincidence, the Jupyter notebook was already popular when RStudio came up
with its own take on this idea.

However, there are some differences between Jupyter and RStudio in the way that
they think about notebooks, and in the notebook interface.

To start with the interface, notice that RStudio has its own menu, with "File",
"Edit", etc.   Each notebook has its own tab in the interface.  If you have
just opened this notebook, and no others, you will only have one tab,
corresponding to this notebook.  Depending on your configuration, you may also
have other windows inside the RStudio window, for example, showing the
variables defined in R, and the files in the same directory as the notebook.

Identify the notebook tab — the tab containing this file.   You will notice that the notebook tab has a *notebook toolbar* at the top.  We will come to that soon.

We have started with the interface, but now we return to RStudio's slightly
different concept of a notebook, compared to Jupyter.

For example, Jupyter thinks in terms of *cells*.   All content in a Jupyter
notebook has a containing cell; text cells contain text, and code cells contain
code.   In a Jupyter notebook, this text would be inside a text cell.

RStudio notebook thinks of the notebook differently.  It conceives of the
notebook as being text, by default.   Within the text, there may be *code
chunks*.  These are the equivalent of Jupyter's code cells — they are blocks of
code you can execute within the notebook.  The notebook interface displays any
output from the code, including plots.

For example, the following is a *code chunk* (in RMarkdown terms):

```{r}
# The cost of Alex's meal.
# alex_meal gets the value 10.50
alex_meal <- 10.50
# Display the cost of Alex's meal.
alex_meal
```

Now we return to the interface.   Click inside the code chunk above.  Notice that the code chunk is grey, compared to the usual default white for the rest of the notebook.  At the top left of the code chunk, you will see a *play* icon.  Click this to run the code and see the results.

There are various other ways of running code chunks. For example, the *notebook
toolbar* that you identified about has icons that allow you run the code.  In
particular, there is a "Run" icon that triggers a drop-down menu, with options
for running this and other code chunks.  You will soon find yourself wanting
the keyboard shortcuts to run code-chunks — please do start using these as
early as you can — you will find they make you much more efficient and fluent
in using notebooks in RStudio.  To see the shortcuts, find the main RStudio
menu, and select "Keyboard Shortcuts Help" for more information.

We will leave you with this brief introduction, and point you out into the
interwebs to search for good tutorials on using RStudio.  Books always get out
of date quickly, so we won't risk instant obsolescence by recommending
particular pages or videos here.
:::
<!---
End of notebook.
-->

:::
<!---
End of R segment.
-->
