---
jupyter:
  jupytext:
    metadata_filter:
      notebook:
        additional: all
        excluded:
        - language_info
    text_representation:
      extension: .Rmd
      format_name: rmarkdown
      format_version: '1.0'
      jupytext_version: 0.8.6
  kernelspec:
    display_name: Python 3
    language: python
    name: python3
resampling_with:
    ed2_fname: 03-Afternote-1
---

```{r setup, include=FALSE}
source("_common.R")
```

# Dramatizing the Power of Simulation

A problem in probability is at the heart of every problem in statistics.
Problems in probability often can be very confusing to the intuition.
And formulas often either do not aid the intuition, or lead to the wrong
answer, or both. But simulation can often get you the right answer and
also help you understand why it is correct. Let's dramatize the point
with puzzles:

### The other child

Problem: **If a family has two children and one of them is a boy, what is the
probability that the other will also be a boy? **

Most people — even professional statisticians — often quickly answer
"one half." That is not correct.

The solution is not obvious even after the person has decided to tackle
the problem by simulation. It is not unusual for a person to say: I flip
one coin, and if it is a head (boy), I then flip another coin and see
how often that will be a boy, also, and then actually flip the coin once
and record the outcomes. But that experiment does not resemble the
situation of interest. A proper modeling throws *two* coins, examines to
see if there is a head on *either* , and then examines the *other*.

Or consider table @tbl-otherboy, where we asked the computer to do this
work for us.  The computer has done 50 trials, where one trial is one family of
two children.  It decided the gender of the two children, by choosing at random
from "Boy" or "Girl", and then classified the pair of children as to whether
the other child was a boy. Where both children were girls, it leaves the
classification blank.

The first 50 lines are enough to suggest the correct probability, and
also to make clear the mechanism: Two-girl pairs, a fourth of the cases, are
excluded from the sample. And mixed pairs — which give a "No" answer — are
two-thirds of the remaining pairs, whereas the only pairs that give "Yes"
answers — two boys — are only a third of the remaining pairs. The point of
presenting the puzzle is this: Simulation gets it right very quickly and
easily, whereas the deductive method of mathematical logic can result in much
confusion.


```{python, eval=TRUE, echo=FALSE}
import numpy as np
import pandas as pd

# To make sure we always get the same random table.
np.random.seed(66)

# Choose the first and second children at random.
trials = np.random.choice(['Boy', 'Girl'], size=(50, 2))
# Put into Pandas dataframe to make results easier to manipulate and display.
boys_df = pd.DataFrame(trials, columns=['First', 'Second'])

def classify(row):
    # Classify the row
    n_males = np.sum(row == 'Boy')
    if n_males == 0:
        return ''
    if n_males == 1:
        return 'No'
    return 'Yes'

# Apply classify function to all rows to give result.
boys_df['Other is boy?'] = boys_df.apply(classify , axis=1)
boys_df.insert(0, 'Trial', range(1, 51))
```

```{r, otherboy, eval=TRUE, echo=FALSE}
t = tibble::as_tibble(py$boys_df)
knitr::kable(list(t[1:25,],
                  t[26:50,]),
  booktabs = TRUE,
  caption = '50 simulations of a family with two children
  {#tbl-otherboy}'
)
```

This puzzle illustrates the power of simulation. And it supports by
analogy the general use of the resampling method in probability and
statistics because it reveals the limitations of human deductive
capacity, even among those who are mathematically adept.

Someone might wonder whether formal mathematics can help us with this
problem. Formal (even though not formulaic) analysis can certainly
provide an answer. We can use what is known as the "sample space"
approach which reasons from first principles; here it consists of making
a *list of the possibilities* , and examining the proportion of
"successes" to "failures" in that list.

First we write down the equally-likely ways that two coins can fall:

|   | First coin | Second coin |
| - | ----- | ------ |
| 1 | Heads |  Heads |
| 2 | Heads |  Tails |
| 3 | Tails |  Heads |
| 4 | Tails |  Tails |

Note that it is very easy to make a mistake in writing this list;
great mathematicians have made such mistakes in the past even with
problems as easy as this one (as we will see with the example of
D'Alembert in just a minute).

Now we notice that if we have observed at least one head, the list of
possibilities for the ways the two coins fell shrinks to:

|   | First coin | Second coin |
| - | ----- | ------ |
| 1 | Heads |  Heads |
| 2 | Heads |  Tails |
| 3 | Tails |  Heads |

And now we can see that in only one of three of these possibilities would the
"other" coin be a head. So the probability we ask about is 1/3.

The crucial question is: Does this formal approach make the problem
harder or easier than the simulation approach? That depends on your
mental makeup. In practice, it turns out that the formal method seems to
lead to a higher rate of error if *everyone* employs it than does the
simulation approach. Hence we emphasize the simulation approach here, to
lead to the highest rate of success (and enjoyment). But it is best if
everyone finds the mode that is best for him or her.

Here's a puzzle I call D'Alembert's Misery. Great mathematicians have
blundered on even simple problems in probability because they attacked
them with only logical tools. One famous example was D'Alembert, living
in the 18th Century (1717-1783; story in @schuh1968master, p 165).

D'Alembert asked: What is the chance of throwing at least one head in
two tosses of a coin? He reasoned that there are three cases: tail on
both tosses, tail then head, head on the first toss (no second toss
necessary). Of these three cases, two are successes, and therefore the
probability sought is 2/3, he concluded.

What's the answer?

1.  Toss two coins, and record number of heads.

2.  Repeat (1) a hundred times.

3.  Count number of outcomes with one or two heads (or more easily, the
    number with two tails), and divide by 100 to find the probability
    sought.

Compare D'Alembert's conclusion to the result of your Monte Carlo
experiment performed as above. Some other fascinating puzzles in
probability are found in Chapter XXX.
