---
jupyter:
  jupytext:
    split_at_heading: true
    text_representation:
      extension: .Rmd
      format_name: rmarkdown
      format_version: '1.1'
      jupytext_version: 1.2.4
  kernelspec:
    display_name: Python 3
    language: python
    name: python3
---

```{r setup, include=FALSE}
source("_common.R")
```

# Introduction to functions

This is a very short introduction to functions.

Functions are *recipes* with names [^berkeley-thanks].

[^berkeley-thanks]: Thanks to [the Berkeley
  team](https://www.inferentialthinking.com/chapters/08/Functions_and_Tables)
  for this metaphor.


Let's start with an example where you have seen functions - mathematics.

## cosine

You have used functions in mathematics.  For example, `cosine` is a function in
mathematics.  We often write the `cosine` function as `cos`

We might write something like this, in mathematics:

$$
y = cos(0)
$$

You can read $cos(0)$ as:

> Call the function "cos" with the value 0, and return the result.

Then $y = cos(0)$ means:

> Call the function "cos" with the value 0, and return the result, storing the
> result in "y".

We *pass* a value to the function.  In our case we pass the value $0$.  The
value that we pass is called an *argument*.

The function *returns* a value.  In our case it returns the calculation of the
cosine of $0$.

You may remember that the cosine of 0 is 1.  So, after $y = cos(0)$, we expect
$y$ to equal 1.

Functions in {{< var lang >}} work in a similar way.

## Functions in {{< var lang >}}

::: python
First we load up the `cos` function from a library of functions called `numpy`.

Don't worry about the command to load the function, for now.  We will come back
to that later.  Here is the command you need. Just run it in the notebook.

```{python}
# Get the cos function from the numpy library.
from numpy import cos
```
:::

We *call* the `cos` function, *passing* the value 0, and see what we get:

```{python, opts.label="python_ed"}
cos(0)
```

```{r, opts.label="r_ed"}
cos(0)
```

* `cos` is the *name* of the function;
* `0` is the value of the *argument* passed to the function.
* The value we pass to the function goes between the parentheses `(` and `)`
  [^brackets].
* `1.0` is the value that the function *returns*.  It is the *return value* of
  the function.

[^brackets]: Please be very careful to distinguish between different types of
  brackets in {{< var lang >}}.  The different types of brackets
  are: *parentheses* (round brackets), as in `()`; *square brackets* as in
  `[]`; and *curly brackets* as in `{}`.  These brackets have different
  meanings to {{< var lang >}}; if you use the wrong type of brackets, you
  will either get an error, or you will get a result you didn't expect.
  When we call functions, we use parentheses.

The value we pass to the function is also called the function *argument*.

In our case we pass `0` as the *argument* to the `cos` function.

We can also store the return value in a variable, like this:

```{python, opts.label="python_ed"}
y = cos(0)
# Show the result
y
```

```{r, opts.label="r_ed"}
y <- cos(0)
# Show the result
y
```

Read this code as:

> Pass the value 0 to the `cos` function.  Put the result into the variable
> `y`. Show the contents of variable `y`.

## The recipe analogy

A *function* is like a *recipe*.  When we cook a dish for a meal, we may follow a recipe. At the top of the recipe, we see the *list of ingredients*.  These are the things that the recipe will work on.  The recipe itself gives the procedure to follow to turn the ingredients into a dish.

For example, here is a recipe for a cheese omelet:

> Name: "My favorite cheese omelet"
>
> ## Ingredients
> * Eggs
> * Butter
> * Salt
> * Pepper
> * Cheese
>
> ## Procedure
>
* * Heat pan to medium heat with a little oil.
> * Grate cheese.
> * Crack the eggs into a bowl.
> * Add the pepper and butter.
> * Beat mixture with fork until well mixed.
> * Pour mixture into heated pan.
> * Tilt pan to spread mixture.
> * Add cheese and salt.
> * Fold sides.
> * Cook for about 30 seconds.
>
> ## Result
> An actual cheese omelet

The recipe has a:

* Name ("My favorite cheese omelet");
* List of ingredients (eggs, butter, etc)
* Procedure (heat pan, etc)
* Result (a nice cheese omelet)

Functions work in the same way. `cos`
is a recipe that does a certain task — in this case, it calculates the cosine value corresponding to the value we sent it.

* Name: `cos`
* List of ingredients (the input value to calculate the cosine for, in our
  case, the value of `x`);
* Procedure (do some calculation work to get the corresponding cosine value)
* Result (the corresponding cosine value).

The list of ingredients to the function are called the *arguments*. The *arguments* are the things that we send to the function.

In our case, there is just one argument: the value of `x`, that we want to calculate the cosine for.

The result is called the *return value*. The *return value* is the thing that
the function sends back.

In our case, the return value is the cosine value corresponding to the input
`x` value.

Returning to the recipe analogy, I can give my recipe any name I want.  Above
I called it "My favorite cheese omelet", but could call my recipe "Recipe
number 4", or anything another name that I liked.  Whatever I called it, it
would be the same recipe.  I might prefer a name that describes what the
recipe makes, to help me remember.

Likewise, the name `cos` refers to a procedure above.  The folks who wrote the
{{< var np_or_r >}} functions could have given the function another name, like
`some_function`, but `cos` is a good name, because it helps us remember what
the procedure does.


## Practice with functions

So far we have used the `cos` function from {{< var np_or_r >}}.  Let's try
doing the same thing with the `sin` function:

::: python
```{python}
# Get the sin function from the numpy library.
from numpy import sin
```
:::

```{python}
sin(0)
```

* `sin` - the function name;
* `0` - the value we pass to the function;
* The value goes after the function name, and between parentheses;
* `0.0` - the value that the function returns.

We pass `0` as the *argument* to the `sin` function.


Now your turn.

The `sqrt` function is (you guessed it) the procedure to go from a number (the
argument) to the square root of that number (the return value).

::: python
Try importing the `sqrt` function from NumPy.
```{python}
# Get the sqrt function from the NumPy library.
# Your code here.
```
:::

Call the `sqrt` function with the value `9`; you should see this returns the
value `3.0`.

```{python, opts.label="python_ed"}
# Your code here.
```

```{r, opts.label="r_ed"}
# Your code here.
```
