---
jupyter:
  jupytext:
    metadata_filter:
      notebook:
        additional: all
        excluded:
        - language_info
    text_representation:
      extension: .Rmd
      format_name: rmarkdown
      format_version: '1.0'
      jupytext_version: 0.8.6
  kernelspec:
    display_name: Python 3
    language: python
    name: python3
resampling_with:
    ed2_fname: 02-Intro
    # Also:
    # "04-Afternote-2.Rmd" - incorporated into intro.Rmd
---

```{r setup, include=FALSE}
source("_common.R")
```

# Introduction

## Uses of Probability and Statistics

This chapter introduces you to probability and statistics. First come
examples of the kinds of practical problems that this knowledge can
solve for us. One reason that the term "statistic" often scares and confuses
people is that the term has several sorts of meanings.  We discuss the meanings
of "statistics" in the section "Types of statistics". Then comes a discussion
on the relationship of probabilities to decisions. Following this we talk about
the limitations of probability and statistics. And last is a discussion of why
statistics can be such a difficult subject. Most important, this chapter
describes the types of problems the book will tackle.

At the foundation of sound decision-making lies the ability to make accurate
estimates of the probabilities of future events. Probabilistic problems
confront everyone — a company owner considering whether to expand their
business, to the scientist testing a vaccine, to the individual deciding
whether to buy insurance.

## What kinds of problems shall we solve? {#sec-what-problems}

These are some examples of the kinds of problems that we can handle with the
methods described in this book:

1. *You are a doctor* trying to develop a treatment for COVID19. Currently you
   are working on a medicine labeled AntiAnyVir. You have data from patients to
   whom medicine AntiAnyVir was given. You want to judge on the basis of those
   results whether AntiAnyVir really improves survival or whether it is no
   better than a sugar pill.

2. *You are the campaign manager* for the Republicrat candidate for President
   of the United States. You have the results from a recent poll taken in New
   Hampshire. You want to know the chance that your candidate would win in New
   Hampshire if the election were held today.

3. *You are the manager and part owner* of one of several contractors
   providing ambulances to a hospital. You own 20 ambulances. Based on
   past experience, the chance that any one ambulance will be unfit
   for service on any given day is about one in ten. You want to know
   the chance on a particular day — tomorrow — that three or more
   of them will be out of action.

2. *You are an environmental scientist* monitoring levels of phosphorus
   pollution in a lake.  The phosphorus levels have been fluctuated around a
   relatively low level until recently, but they have been higher in the last
   few years. Does these recent higher levels indicate some important change or
   can we put them down to some chance and ordinary variation from year to
   year?

The core of all these problems, and of the others that we will deal with in
this book, is that you want to know the "chance" or "probability" — different
words for the same idea — that some event will or will not happen, or that
something is true or false. To put it another way, we want to answer questions
about "What is the probability that...?", given the body of information that
you have in hand.

The question "What is the probability that...?" is usually not the ultimate
question that interests us at a given moment.

Eventually, a person wants to use the estimated probability to help make a
*decision* concerning some action one might take. These are the kinds of
decisions, related to the questions about probability stated above, that
ultimately we would like to make:

1. *Should you (the researcher) advise doctors to prescribe medicine
   AntiAnyVir* for COVID19 patients, or, should you (the researcher) continue
   to study AntiAnyVir before releasing it for use? A related matter: should
   you and other research workers feel sufficiently encouraged by the results
   of medicine AntiAnyVir so that you should continue research in this general
   direction rather than turning to some other promising line of research?
   These are just two of the possible decisions that might be influenced by
   the answer to the question about the probability that medicine AntiAnyVir
   is effective in treating COVID19.

2. *Should you advise the Republicrat presidential candidate to go to New
   Hampshire* to campaign? If the poll tells you conclusively that she or he
   will not win in New Hampshire, you might decide that it is not worthwhile
   investing effort to campaign there. Similarly, if the poll tells you
   conclusively that they surely will win in New Hampshire, you probably would
   not want to campaign further there. But if the poll is not conclusive in one
   direction or the other, you might choose to invest the effort to campaign in
   New Hampshire. Analysis of the chances of winning in New Hampshire based on
   the poll data can help you make this decision sensibly.

3. *Should your company buy more ambulances*? Clearly the answer to this
   question is affected by the probability that a given number of your
   ambulances will be out of action on a given day. But of course this
   estimated probability will be only one part of the decision.

4. *Should we search for new causes of phosphorus pollution* as a result of the
   recent measurements from the lake?  If the causes have not changed, and the
   recent higher values were just the result of ordinary variation, our search
   will end up wasting time and money that could have been better spent
   elsewhere.

The kinds of questions to which we wish to find probabilistic and statistical
answers may be found throughout the social, biological and physical sciences;
in business; in politics; in engineering; and in most other forms of human
endeavor.

## Types of statistics {#sec-types-of-statistics}

The term *statistics* sometimes causes confusion and therefore needs
explanation.

Statistics can mean two related things.  It can refer to a certain sort of
*number* — of which more below.  Or it can refer to the field of inquiry that
studies these numbers.

A *statistic* is a *number* that we can calculate from a larger collection of
numbers we are interested in.  For example, table @tbl-yearly-srp has some
yearly measures of "soluble reactive phosphorus" (SRP) from Lough Erne — a
lake in Ireland [@zhou2000lake].

```{python, eval=TRUE, echo=FALSE}
from pathlib import Path

import numpy as np
import pandas as pd
lake = pd.read_csv(Path('data') / 'lough_erne.csv')
yearly_srp = lake.loc[:, ['Year', 'SRP']].copy()
```

```{r, eval=TRUE, echo=FALSE}
ketable(py$yearly_srp,
        caption = "Soluble Reactive Phosphorus in Lough Erne {#tbl-yearly-srp}")
```

We may want to *summarize* this set of SRP measurements. For example, we could
add up all the SRP values to give the *total*.  We could also divide the total
by the number of measurements, to give the *average*.  Or we could measure the
spread of the values by finding the *minimum* and the *maximum* — see table
@tbl-srp-stats). All these numbers are *descriptive statistics*, because
they are summaries that *describe* the collection of SRP measurements.


```{python, eval=TRUE, echo=FALSE}
srp = np.array(yearly_srp['SRP'])
srp_stats = pd.DataFrame({'Descriptive statistics for SRP': pd.Series({
    'Total': np.sum(srp),
    'Mean': np.mean(srp),
    'Minimum': np.min(srp),
    'Maximum': np.max(srp)})})
```

```{r, eval=TRUE, echo=FALSE}
ketable(head(py$srp_stats),
        caption = "Statistics for SRP levels {#tbl-srp-stats}")
```

Descriptive statistics are nothing new to you; you have been using many of them
all your life.

We can calculate other numbers that can be useful for drawing *conclusions* or
*inferences* from a collection of numbers; these are *inferential statistics*.
Inferential statistics are often probability values that give the answer to
questions like "What are the chances that ...".

```{r, eval=TRUE, echo=FALSE}
ysrp <- py$yearly_srp
pre_1990 <- round(mean(ysrp[ysrp$Year < 1990, 'SRP']), 1)
post_1990 <- round(mean(ysrp[ysrp$Year > 1990, 'SRP']), 1)
```

For example, imagine we suspect there was some environmental change in 1990. We
see that the average SRP value before 1990 was `r pre_1990` and the average SRP
value after 1990 was `r post_1990`.  That gives us a difference in the average
of `r post_1990 - pre_1990`. But, could this difference be due to chance
fluctuations from year to year?  Were we just unlucky in getting a few larger
measurements in later years?  We could use methods that you will see in this
book to calculate a *probability* to answer that question.  The probability
value is an *inferential* statistic, because we can use it to draw an
*inference* about the measures.

Inferential statistics use descriptive statistics as their input. Inferential
statistics can be used for two purposes: to aid scientific *understanding* by
estimating the probability that a statement is true or not, and to aid in
making *sound decisions* by estimating which alternative among a range of
possibilities is most desirable.

## Probabilities and decisions

There are two differences between questions about probabilities and
the ultimate decision problems:

1.  Decision problems always involve *evaluation of the
    consequences*  —  that is, taking into account the benefits and
    the costs of the consequences  —  whereas pure questions about
    probabilities are estimated without evaluations of the consequences.

2.  Decision problems often involve a *complex combination* of *sets
    of probabilities and consequences*, together with their
    evaluations. For example: In the case of the contractor's
    ambulances, it is clear that there will be a monetary loss to the
    contractor if she makes a commitment to have 17 ambulances available for
    tomorrow and then cannot produce that many. Furthermore, the contractor
    must take into account the further consequence that there *may* be a loss
    of goodwill for the future if she fails to meet her obligations
    tomorrow  —  and then again there *may not* be any such loss; and if there
    is such loss of goodwill it might be a loss worth \$10,000 *or* \$20,000
    *or* \$30,000. Here the decision problem involves not only the probability
    that there will be fewer than 17 ambulances tomorrow but also the
    immediate monetary loss and the subsequent possible losses of goodwill,
    and the valuation of all these consequences.

Continuing with the decision concerning whether to do more research on medicine
AntiAnyVir: If you do decide to continue research on AntiAnyVir, (a) you may,
or (b) you may not, come up with an important general treatment for viral
infections within, say, the next 3 years. If you do come up with such a general
treatment, of course it will have very great social benefits. Furthermore, (c)
if you decide not to do further research on AntiAnyVir now, you can direct your
time and that of other people to research in other directions, with some chance
that the other research will produce a less-general but nevertheless useful
treatment for some relatively infrequent viral infections. Those three
possibilities have different social benefits. The probability that medicine
AntiAnyVir really has some benefit in treating COVID19, as judged by your prior
research, obviously will influence your decision on whether or not to do more
research on medicine AntiAnyVir. But that judgment about the probability is
only one part of the overall web of consequences and evaluations that must be
taken into account when making your decision whether or not to do further
research on medicine AntiAnyVir.

Why does this book limit itself to the specific probability questions when
ultimately we are interested in decisions? A first reason is division of
labor. The more general aspects of the decision-making process in the face of
uncertainty are treated well in other books. This book's special contribution
is its new approach to the crucial process of estimating the chances that an
event will occur.

Second, the specific elements of the overall decision-making process taught in
this book belong to the interrelated subjects of *probability theory* and
*statistics* . Though probabilistic and statistical theory ultimately is
intended to be part of the general decision-making process, often only the
estimation of probabilities is done systematically, and the rest of the
decision-making process — for example, the decision whether or not to proceed
with further research on medicine AntiAnyVir — is done in informal and
unsystematic fashion. This is regrettable, but the fact that this is standard
practice is an additional reason why the treatment of statistics and
probability in this book is sufficiently complete.

A third reason that this book covers only statistics and not numerical
reasoning about decisions is because most college and university statistics
courses and books are limited to statistics.

## Limitations of probability and statistics

Statistical testing is not equivalent to research, and research is not the same
as statistical testing. Rather, statistical inference is a handmaiden of
research, often but not always necessary in the research process.

A working knowledge of the basic ideas of statistics, especially the
elements of probability, is unsurpassed in its general value to everyone
in a modern society. Statistics and probability help clarify one's
thinking and improve one's capacity to deal with practical problems and
to understand the world. To be efficient, a social scientist or
decision-maker is almost certain to need statistics and probability.

On the other hand, important research and top-notch decision-making have
been done by people with absolutely no formal knowledge of statistics.
And a limited study of statistics sometimes befuddles students into
thinking that statistical principles are guides to research design and
analysis. This mistaken belief only inhibits the exercise of sound
research thinking. Alfred Kinsey long ago put it this way:

> ... no statistical treatment can put validity into generalizations which are
> based on data that were not reasonably accurate and complete to begin with.
> It is unfortunate that academic departments so often offer courses on the
> statistical manipulation of human material to students who have little
> understanding of the problems involved in securing the original data. ...
> When training in these things replaces or at least precedes some of the
> college courses on the mathematical treatment of data, we shall come nearer
> to having a science of human behavior. [@kinsey1948sexual, p 35].

In much — even most — research in social and physical sciences,
statistical testing is not necessary. Where there are large differences between
different sorts of circumstances for example, if a new medicine cures 90
patients out of 100 and the old medicine cures only 10 patients out of 100 — we
do not need refined statistical tests to tell us whether or not the new
medicine really has an effect. And the best research is that which shows large
differences, because it is the large effects that matter. If the researcher
finds that s/he must use refined statistical tests to reveal whether there are
differences, this sometimes means that the differences do not matter much.

To repeat, then, some or even much research — especially in the physical and
biological sciences — does not need the kind of statistical manipulation that
will be described in this book. But most decision problems *do* need the kind
of probabilistic and statistical input that is described in this book.

Another matter: If the raw data are of poor quality, probabilistic and
statistical manipulation cannot be very useful. In the example of the
contractor and her ambulances, if the contractor's estimate that a given
ambulance has a one-in-ten chance of being unfit for service out-of-order on
a given day is very inaccurate, then our calculation of the probability that
three or more ambulances will be out of order on a given day will not be
helpful, and may be misleading. To put it another way, one cannot make bread
without flour, yeast, and water. And good raw data are the flour, yeast and
water necessary to get an accurate estimate of a probability. The most refined
statistical and probabilistic manipulations are useless if the input data are
poor — the result of unrepresentative samples, uncontrolled experiments,
inaccurate measurement, and the host of other ways that information gathering
can go wrong. (See @simon1985basic for a catalog of the obstacles to obtaining
good data.) Therefore, we should constantly direct our attention to ensuring
that the data upon which we base our calculations are the best it is possible
to obtain.

## Why is Statistics Such a Difficult Subject? {#sec-stats-difficult}

Why is statistics such a tough subject for so many people?

"Among mathematicians and statisticians who teach introductory
statistics, there is a tendency to view students who are not skillful in
mathematics as unintelligent," say two of the authors of a
popular introductory text [@mccabe1989instructors, p 2]. As these authors
imply, this view is out-and-out wrong; lack of general intelligence on the
part of students is *not* the root of the problem.

Scan this book and you will find almost no formal mathematics. Yet
nearly every student finds the subject very difficult —  as difficult as
anything taught at universities. The root of the difficulty is that the
*subject matter* is extremely difficult. Let's find out *why* .

It is easy to find out with high precision which movie is playing
tonight at the local cinema; you can look it up on the web or call the cinema
and ask. But consider by contrast how difficult it is to determine with
accuracy:

1. Whether we will save lives by recommending vitamin D supplements for the
   whole population as protection against viral infections.  Some evidence
   suggests that low vitamin D levels predispose to more severe lung
   infections, and that taking supplements can help [@martineau2017vitamin].
   But, how certain can we be of the evidence?  How safe are the supplements?
   Does the benefit, and the risk, differ by ethnicity?
2. What will be the result of more than a hundred million Americans voting for
   president a month from now; the best attempt usually is a sample of 2000
   people, selected in some fashion or another that is far from random, weeks
   before the election, asked questions that are by no means the same as the
   actual voting act, and so on;
3. How men feel about women and vice versa.

The cleverest and wisest people have pondered for thousands of years how to
obtain answers to questions like these, and made little progress. Dealing with
uncertainty was completely outside the scope of the ancient philosophers. It
was not until two or three hundred years ago that people began to make any
progress at all on these sorts of questions, and it was only about one century
ago that we began to have reasonably competent procedures — simply because the
problems are inherently difficult. So it is no wonder that the body of these
methods is difficult.

So: The bad news is that the subject is extremely difficult. The good
news is that you — and that means *you* — can understand it with hard thinking,
even if you have no mathematical background beyond arithmetic and you think
that you have no mathematical capability. That's because the difficulty lies in
such matters as pin-pointing the right question, but not in any difficulties of
mathematical manipulation.
