---
jupyter:
  jupytext:
    metadata_filter:
      notebook:
        additional: all
        excluded:
        - language_info
    text_representation:
      extension: .Rmd
      format_name: rmarkdown
      format_version: '1.0'
      jupytext_version: 0.8.6
  kernelspec:
    display_name: Python 3
    language: python
    name: python3
resampling_with:
    ed2_fname: 12-Chap-8
---

```{r setup, include=FALSE}
source("_common.R")
```

# Probability Theory, Part 4: Estimating Probabilities from Finite Universes {#sec-finite-universes}

## Introduction

The examples in @sec-infinite-universes dealt with *infinite universes* , in
which the probability of a given simple event is unaffected by the outcome of
the previous simple event. But now we move on to finite universes, situations
in which you begin with a *given set of objects* whose number is not enormous —
say, a total of two, or two hundred, or two thousand. If we liken such a
situation to a bucket containing balls of different colors each with a number
on it, we are interested in the probability of drawing various sets of numbered
and colored balls from the bucket on the condition that we *do not replace*
balls after they are drawn.

In the cases addressed in this chapter, it is important to remember that
the single events no longer are independent of each other. A typical
situation in which sampling without replacement occurs is when items are
chosen from a finite universe — for example, when children are selected
randomly from a classroom. If the class has five boys and five girls,
and if you were to choose three girls in a row, then the chance of
selecting a fourth girl on the next choice obviously is lower than the
chance that you would pick a girl on the first selection.

The key to dealing with this type of problem is the same as with earlier
problems: You must choose a simulation procedure that produces simple
events having the same probabilities as the simple events in the actual
problem involving sampling without replacement. That is, you must make
sure that your simulation does not allow duplication of events that have
already occurred. The easiest way to sample without replacement with resampling
techniques is by simply ignoring an outcome if it has already occurred.

Examples @sec-four-girls-one-boy through @sec-five-spades-four-girls deal with
some of the more important sorts of questions one may ask about drawings
without replacement from such an urn. To get an overview, I suggest that you
read over the summaries (in **bold**) introducing examples
@sec-four-girls-one-boy to @sec-five-spades-four-girls before beginning to work
through the examples themselves.

This chapter also revisits the general procedure used in solving problems in
probability and statistics with simulation, here in connection with problems
involving a finite universe. The steps that one follows in simulating the
behavior of a universe of interest are set down in such fashion that one may,
by random drawings, deduce the probability of various events. Having had by now
the experience of working through the problems in @sec-prob-theory-one-b and
@sec-infinite-universes, the reader should have a solid basis to follow the
description of the general procedure which then helps in dealing with specific
problems.

Let us begin by describing some of the major sorts of problems with the
aid of a bucket with six balls.

## Some building-block programs

**Case 1.** Each of six balls is labeled with a number between "1" and
"6." We ask: What is the probability of choosing balls 1, 2, and 3 *in
that order* if we choose three balls without replacement? @fig-success_case1
diagrams the events we consider "success."

```{r fig-success_case1, opts.label='svg_fig', fig.cap="The Event Classified as \"Success\" for Case 1"}
include_svg('diagrams/success_case1.svg')
```

**Case 2.** We begin with the same bucket as in Case 1, but now ask the
probability of choosing balls 1, 2, and 3 *in any order* if we choose
three balls without replacement. @fig-success_case2 diagrams two of the events
we consider success. These possibilities include that which is shown in
@fig-success_case1 above, plus other possibilities.

```{r fig-success_case2, opts.label='svg_fig', fig.cap="An Incomplete List of the Events Classified as \"Success\" for Case 2"}
include_svg('diagrams/success_case2.svg')
```

**Case 3.** The odd-numbered balls "1," "3," and "5," are painted red
and the even-numbered balls "2," "4," and "6" are painted black. What is
the probability of getting a red ball and then a black ball in that
order? Some possibilities are illustrated in @fig-success_case3, which includes
the possibility shown in @fig-success_case1. It also includes *some but not*
all possibilities found in  @fig-success_case2; for example, @fig-success_case2
includes choosing balls 2, 3 and 1 in that order, but @fig-success_case3 does
not.

```{r fig-success_case3, opts.label='svg_fig', fig.cap="An Incomplete List of the Events Classified as \"Success\" for Case 3"}
include_svg('diagrams/success_case3.svg')
```

**Case 4.** What is the probability of getting two red balls and one
black ball in any order?

```{r fig-success_case4, opts.label='svg_fig', fig.cap="An Incomplete List of the Events Classified as \"Success\" for Case 4"}
include_svg('diagrams/success_case4.svg')
```

**Case 5.** Various questions about *matching* may be asked with respect
to the six balls. For example, what is the probability of getting ball 1
on the first draw *or* ball 2 on the second draw *or* ball 3 on the
third draw? (@fig-success_case5) Or, what is the probability of getting all
balls on the draws corresponding to their numbers?

```{r fig-success_case5, opts.label='svg_fig', fig.cap="An Incomplete List of the Events Classified as \"Success\" for Case 5"}
include_svg('diagrams/success_case5.svg')
```

## Problems in finite universes

### Example: four girls and one boy {#sec-four-girls-one-boy}

**What is the probability of selecting four girls and one boy when selecting
five students from any group of twenty-five girls and twenty-five boys?**  This
is an example of sampling without replacement when there are two outcomes and
the order does not matter.

The important difference between this example and the infinite-universe
examples in the prior chapter is that the probability of obtaining a boy
or a girl in a single simple event *differs* from one event to the next
in this example, whereas it stays the same when the sampling is with
replacement. To illustrate, the probability of a girl is .5 (25 out of
50) when the first student is chosen, but the probability of a girl is
either 25/49 or 24/49 when the second student is chosen, depending on
whether a boy or a girl was chosen on the first pick. Or after, say,
three girls and one boy are picked, the probability of getting a girl on
the next choice is (28-3)/(50-4) = 22/46 which is clearly not equal to
.5.

As always, we must create a satisfactory analog to the process whose
probability we want to learn. In this case, we can use a deck of 50
cards, half red and half black, and deal out five cards *without
replacing them* after each card is dealt; this simulates the choice of
five students from among the fifty.

We can no longer use our procedure from before.  If we designated
"1-25" as being girls and "26-50" as being boys and then proceeded to
draw random numbers, the probability of a girl would be the same on
each pick.

At this point, it is important to note that — for this particular problem — we
do not need to distinguish between particular girls (or boys). That is, it does
not matter *which* girl (or boy) is selected in a given trial. Nor did we pay
attention to the *order* in which we selected girls or boys. This is an
instance of Case 4 discussed above. Subsequent problems will deal with
situations where the order of selection, and the particular individuals, do
matter.

Our approach then is to mimic having the class in front of us: an array of 50
strings, half of the entries 'boy' and the other half 'girl'.  We then shuffle
the class (the array), and choose the first N students (strings).

* **Step 1.** Create a list with 50 labels, half 'boy' and half 'girl'.
* **Step 2.** Shuffle the class and select five students. Count whether there
  are four labels equal 'girl'. If so, write "yes," otherwise "no".
* **Step 3.** Repeat step 2, say, 10,000 times, and count the proportion "yes",
  which estimates the probability sought.

The results of a few experimental trials are shown in @tbl-four-girls-one-boy.

+----------------+---------------------------------------+---------------+
| **Experiment** | **Strings Chosen**                    | **Success?**  |
+----------------+---------------------------------------+---------------+
|              1 | 'girl', 'boy', 'boy', 'girl', 'boy'   | No            |
+----------------+---------------------------------------+---------------+
|              2 | 'boy', 'girl', 'girl', 'girl', 'girl' | Yes           |
+----------------+---------------------------------------+---------------+
|              3 | 'girl, 'girl', 'girl', 'boy', 'girl'  | Yes           |
+----------------+---------------------------------------+---------------+

: A few experimental trials of four girls and one boy {#tbl-four-girls-one-boy}

A solution to this problem with {{< var lang >}} is presented below.

::: {.notebook name="finite_universes" title="Experiments in finite universes"}

```{python}
import numpy as np

rnd = np.random.default_rng()
```

```{python}
N = 10000
trial_results = np.zeros(N)

# Constitute the set of 25 girls and 25 boys.
whole_class = np.repeat(['girl', 'boy'], [25, 25])

# Repeat the following steps N times.
for i in range(N):

    # Shuffle the numbers
    shuffled = rnd.permuted(whole_class)

    # Take the first 5 numbers, call them c.
    c = shuffled[:5]

    # Count how many girls there are, put the result in d.
    d = np.sum(c == 'girl')

    # Keep track of each trial result in z.
    trial_results[i] = d

    # End the experiment, go back and repeat until all 1000 trials are
    # complete.

# Count the number of times we got four girls, put the result in k.
k = np.sum(trial_results == 4)

# Convert to a proportion.
kk = k / N

# Print the result.
print(kk)
```

```{r}
N <- 10000
trial_results <- numeric(N)

# Constitute the set of 25 girls and 25 boys.
whole_class <- rep(c('girl', 'boy'), c(25, 25))

# Repeat the following steps N times.
for (i in 1:N) {

    # Shuffle the numbers
    shuffled <- sample(whole_class)

    # Take the first 5 numbers, call them c.
    c <- shuffled[1:5]

    # Count how many girls there are, put the result in d.
    d <- sum(c == 'girl')

    # Keep track of each trial result in z.
    trial_results[i] <- d

    # End the experiment, go back and repeat until all 1000 trials are
    # complete.
}

# Count the number of times we got four girls, put the result in k.
k <- sum(trial_results == 4)

# Convert to a proportion.
kk <- k / N

# Print the result.
message(kk)
```

We can also find the probabilities of other outcomes from a histogram of trial
results obtained with the following command:

```{python}
# Import the plotting package.
import matplotlib.pyplot as plt

# Do histogram, with one bin for each possible number.
plt.hist(trial_results, bins=range(7), align='left', rwidth=0.75)
plt.title('# of girls');
```

```{r}
# Do histogram, with one bin for each possible number.
hist(trial_results, breaks=0:max(trial_results), main='# of girls')
```

```{r, echo=FALSE, eval=TRUE}
trial_results <- get_var('trial_results')  # Might be Python edition.
pct_4 <- round(sum(trial_results == 4) / length(trial_results) * 100)
```

In the resulting histogram we can see that in `r pct_4` percent of the trials,
4 of the 5 selected were girls.

It should be noted that for this problem — as for most other problems — there
are several other resampling procedures that will also do the job correctly.

In analytic probability theory this problem is worked with a formula for
"combinations."

### Example: Five spades and four clubs in a bridge hand {#sec-five-spades-four-clubs}

**This is an example of multiple-outcome sampling without replacement, order does not matter**.

The problem is similar to the example in @sec-four-girls-one-boy, except that
now there are four equally-likely outcomes instead of only two. [A
Python]{.python}[An R]{.r} solution is:

```{python}
# Constitute the deck of 52 cards.
# Repeat the suit names 13 times each, to make a 52 card deck.
deck = np.repeat(['spade', 'club', 'diamond', 'heart'],
                 [13, 13, 13, 13])
# Show the deck
deck
```

```{python}
N = 10000
trial_results = np.zeros(N)

# Repeat the trial N times.
for i in range(N):

    # Shuffle the deck and draw 13 cards.
    hand = rnd.choice(deck, size=13, replace=False)

    # Count the number of spades in "hand", put the result in "n_spades".
    n_spades = np.sum(hand == 'spade')

    # If we have five spades, we'll continue on to count the clubs. If we don't
    # have five spades, the number of clubs is irrelevant — we have not gotten
    # the hand we are interested in.
    if n_spades == 5:
        # Count the clubs, put the result in "n_clubs"
        n_clubs = np.sum(hand == 'club')
        # Keep track of the number of clubs in each trial
        trial_results[i] = n_clubs

    # End one experiment, go back and repeat until all N trials are done.

# Count the number of trials where we got 4 clubs. This is the answer we want -
# the number of hands out of 1000 with 5 spades and 4 clubs. (Recall that we
# only counted the clubs if the hand already had 5 spades.)
n_5_and_4 = np.sum(trial_results == 4)

# Convert to a proportion.
prop_5_and_4 = n_5_and_4 / N

# Print the result
print(prop_5_and_4)
```

```{r}
# Constitute the deck of 52 cards.
# Repeat the suit names 13 times each, to make a 52 card deck.
deck <- rep(c('spade', 'club', 'diamond', 'heart'), c(13, 13, 13, 13))
# Show the deck
deck
```

```{r}
N <- 10000
trial_results <- numeric(N)

# Repeat the trial N times.
for (i in 1:N) {

    # Shuffle the deck and draw 13 cards.
    hand <- sample(deck, 13)  # replace=FALSE is the default.

    # Count the number of spades in "hand", put the result in "n_spades".
    n_spades <- sum(hand == 'spade')

    # If we have five spades, we'll continue on to count the clubs. If we don't
    # have five spades, the number of clubs is irrelevant — we have not gotten
    # the hand we are interested in.
    if (n_spades == 5) {
        # Count the clubs, put the result in "n_clubs"
        n_clubs <- sum(hand == 'club')
        # Keep track of the number of clubs in each trial
        trial_results[i] <- n_clubs
    }

    # End one experiment, go back and repeat until all N trials are done.
}

# Count the number of trials where we got 4 clubs. This is the answer we want -
# the number of hands out of 1000 with 5 spades and 4 clubs. (Recall that we
# only counted the clubs if the hand already had 5 spades.)
n_5_and_4 <- sum(trial_results == 4)

# Convert to a proportion.
prop_5_and_4 <- n_5_and_4 / N

# Print the result
message(prop_5_and_4)
```

### Example: a total of fifteen points in a bridge hand {#sec-fifteen-bridge}

Let us assume that ace counts as 4, king = 3, queen = 2, and jack = 1.

```{python}
# Constitute a deck with 4 jacks (point value 1), 4 queens (value 2), 4
# kings (value 3), 4 aces (value 4), and 36 other cards with no point
# value
whole_deck = np.repeat([1, 2, 3, 4, 0], [4, 4, 4, 4, 36])
whole_deck
```

```{python}
N = 10000
trial_results = np.zeros(N)

# Do N trials.
for i in range(N):
    # Shuffle the deck of cards and draw 13
    hand = rnd.choice(whole_deck, size=13, replace=False)

    # Total the points.
    points = np.sum(hand)

    # Keep score of the result.
    trial_results[i] = points

    # End one experiment, go back and repeat until all N trials are done.
```

```{python}
# Produce a histogram of trial results.
plt.hist(trial_results, bins=range(25), align='left', rwidth=0.75)
plt.title('Points in bridge hands');
```

```{r}
# Constitute a deck with 4 jacks (point value 1), 4 queens (value 2), 4
# kings (value 3), 4 aces (value 4), and 36 other cards with no point
# value
whole_deck <- rep(c(1, 2, 3, 4, 0), c(4, 4, 4, 4, 36))
whole_deck
```

```{r}
N <- 10000
trial_results <- numeric(N)

# Do N trials.
for (i in 1:N) {
    # Shuffle the deck of cards and draw 13
    hand <- sample(whole_deck, size=13)  # replace=FALSE is default.

    # Total the points.
    points <- sum(hand)

    # Keep score of the result.
    trial_results[i] <- points

    # End one experiment, go back and repeat until all N trials are done.
}
```

```{r}
# Produce a histogram of trial results.
hist(trial_results, breaks=0:max(trial_results), main='Points in bridge hands')
```

```{r, echo=FALSE, eval=TRUE}
trial_results <- get_var('trial_results')  # Might be Python edition.
pct_15 <- round(sum(trial_results == 15) / length(trial_results) * 100)
```

From this histogram, we see that in about `r pct_15` percent of our trials we
obtained a total of exactly 15 points. We can also compute this directly:

```{python}
# How many times did we have a hand with fifteen points?
k = np.sum(trial_results == 15)

# Convert to a proportion.
kk = k / N

# Show the result.
kk
```

```{r}
# How many times did we have a hand with fifteen points?
k <- sum(trial_results == 15)

# Convert to a proportion.
kk <- k / N

# Show the result.
kk
```

### Example: Four girls then one boy from 25 girls and 25 boys

**In this problem, order matters; we are sampling without replacement, with two
outcomes, several of each item.**

What is the probability of getting an ordered series of *four girls and then
one boy* , from a universe of 25 girls and 25 boys? This illustrates Case 3
above. Clearly we can use the same sampling mechanism as in the example
@sec-four-girls-one-boy, but now we record "yes" for a smaller number of
composite events.

We record "no" even if a single one boy is chosen but he is chosen 1st, 2nd,
3rd, or 4th, whereas in @sec-four-girls-one-boy, such outcomes are recorded as
"yes"-es.

* **Step 1.** Generate a class ({{< var array >}}) of length 50, consisting of
  25 strings valued "boy" and 25 strings valued "girl".
* **Step 2.** Shuffle the class array, and select the first five
  elements. If those are `['girl', 'girl', 'girl', 'girl', 'boy']`,
  write "yes," otherwise "no."
* **Step 3.** Repeat step 2, say, 10,000 times, and count the proportion of
  "yes" results, which estimates the probability sought.

```{python}
# The result we are looking for - four girls and then a boy.
result_of_interest = np.repeat(['girl', 'boy'], [4, 1])
result_of_interest
```

```{python}
N = 10000
trial_results = np.zeros(N)

# Constitute the set of 25 girls and 25 boys.
whole_class = np.repeat(['girl', 'boy'], [25, 25])

# Repeat the following steps 1000 times.
for i in range(N):

    # Shuffle the numbers
    shuffled = rnd.permuted(whole_class)
    # Take the first 5 class members, call them c.
    c = shuffled[:5]
    # Or: c = rnd.choice(whole_class, replace=False)

    is_four_girls_then_one_boy = np.all(c == result_of_interest)
    # Or:
    # is_four_girls_then_one_boy = (
    #   np.all(c[:4] == 'girl') and c[4] == 'boy'
    # )

    trial_results[i] = is_four_girls_then_one_boy

    # End the experiment, go back and repeat until all N trials are
    # complete.

# Count the number of times we got four girls then a boy
k = np.sum(trial_results)

# Convert to a proportion.
kk = k / N

# Print the result.
print(kk)
```

<!---
Introduce `all`
-->

```{r}
# The result we are looking for - four girls and then a boy.
result_of_interest <- rep(c('girl', 'boy'), c(4, 1 ))
result_of_interest
```

```{r}
N <- 10000
trial_results <- numeric(N)

# Constitute the set of 25 girls and 25 boys.
whole_class <- rep(c('girl', 'boy'), c(25, 25))

# Repeat the following steps 1000 times.
for (i in 1:N) {

    # Shuffle the numbers
    shuffled <- sample(whole_class)
    # Take the first 5 class members, call them c.
    c <- shuffled[1:5]

    is_four_girls_then_one_boy <- all(c == result_of_interest)

    trial_results[i] <- is_four_girls_then_one_boy

    # End the experiment, go back and repeat until all N trials are
    # complete.
}

# Count the number of times we got four girls then a boy
k <- sum(trial_results)

# Convert to a proportion.
kk <- k / N

# Print the result.
message(kk)
```

This type of problem is conventionally done with a *permutation*
formula.

### Example: repeat pairings from random pairing

First put two groups of 10 people into 10 pairs.  Then re-randomize the
pairings.  What is the chance that four or more pairs are the same in the
second random pairing?

**This is a problem in the probability of matching by chance**.

Ten representatives each from two universities, Birmingham and Berkeley, attend
a meeting.  As a social icebreaker, representatives are divided, randomly, into
pairs consisting of one person from each university.

If they held a second round of the icebreaker, with a new random pairing, what
is the chance that four or more pairs will be the same?

In approaching this problem, we start at the point where the first icebreaker
is complete.  We now have to determine what happens after the second round.

* **Step 1.** Let "ace" through "10" of hearts represent the ten
  representatives from Birmingham University.  Let "ace" through "10" of spades
  be their allocated partners (in round one) from Berkeley.
* **Step 2.** Shuffle the hearts and deal them out in a row; shuffle the spades
  and deal in a row just below the hearts.
* **Step 3.** Count the pairs — a pair is one card from the heart row and one
  card from the spade row — that contain the same denomination. If 4 or more
  pairs match, record "yes," otherwise "no."
* **Step 4.** Repeat steps (2) and (3), say, 10,000 times.
* **Step 5.** Count the proportion "yes." This estimates the probability of 4
  or more pairs.

Exercise for the student: Write the steps to do this example with random
numbers.  The {{< var lang >}} solution follows below.

```{python}
N = 10000
trial_results = np.zeros(N)

# Assign numbers to each student, according to their pair, after the first
# icebreaker
birmingham = np.arange(10)
berkeley = np.arange(10)

for i in range(N):
    # Randomly shuffle the students from Berkeley
    shuffled_berkeley = rnd.permuted(berkeley)

    # Randomly shuffle the students from Birmingham
    # (This step is not really necessary — shuffling one array is enough to make the matching random.)
    shuffled_birmingham = rnd.permuted(birmingham)

    # Count in how many cases people landed with the same person as in the
    # first round, and store in trial_results.
    matches = np.sum(shuffled_berkeley == shuffled_birmingham)
    trial_results[i] = matches

# Count the number of times we got 4 or more people assigned to the same person
k = np.sum(trial_results >= 4)

# Convert to a proportion.
kk = k / N

# Print the result.
print(kk)
```

```{r}
N <- 10000
trial_results <- numeric(N)

# Assign numbers to each student, according to their pair, after the first
# icebreaker
birmingham <- 1:10
berkeley <- 1:10

for (i in 1:N) {
    # Randomly shuffle the students from Berkeley
    shuffled_berkeley <- sample(berkeley)

    # Randomly shuffle the students from Birmingham
    # (This step is not really necessary — shuffling one array is enough to make the matching random.)
    shuffled_birmingham <- sample(birmingham)

    # Count in how many cases people landed with the same person as in the
    # first round, and store in trial_results.
    matches <- sum(shuffled_berkeley == shuffled_birmingham)
    trial_results[i] <- matches
}

# Count the number of times we got 4 or more people assigned to the same person
k <- sum(trial_results >= 4)

# Convert to a proportion.
kk <- k / N

# Print the result.
message(kk)
```

We see that in about `r round(get_var('kk') * 100)` percent of the trials did 4
or more couples end up being re-paired with their own partners.  This can also
be seen from the histogram:

```{python}
# Produce a histogram of trial results.
plt.hist(trial_results, bins=range(10), align='left', rwidth=0.75)
plt.title('Same pairs in round two');
```

### Example: Matching Santa Hats

**The welcome staff at a restaurant mix up the hats of a party of six Christmas
Santas. What is the probability that at least one will get their own hat?**.

After a long Christmas day, six Santas meet in the pub to let off steam.
However, as luck would have it, their hosts have mixed up their hats. When the
hats are returned, what is the chance that at least one Santa will get his own
hat back?

First, assign each of the six Santas a number, and place these numbers in an
array. Next, shuffle the array (this represents the mixed-up hats) and compare
to the original. The rest of the problem is the same as the pairs one from
before, except that we are now interested in any trial where at least one ($\ge
1$) Santa received the right hat.

```{python}
N = 10000
trial_results = np.zeros(N, dtype=bool)

# Assign numbers to each owner
owners = np.arange(6)

# Each hat gets the number of their owner
hats = np.arange(6)

for i in range(N):
    # Randomly shuffle the hats and compare to their owners
    shuffled_hats = rnd.permuted(hats)

    # In how many cases did at least one person get their hat back?
    trial_results[i] = np.sum(shuffled_hats == owners) >= 1

# How many times, over all trials, did at least one person get their hat back?
k = np.sum(trial_results)

# Convert to a proportion.
kk = k / N

# Print the result.
print(kk)
```

```{r}
N <- 10000
trial_results <- numeric(N)

# Assign numbers to each owner
owners <- 1:6

# Each hat gets the number of their owner
hats <- 1:6

for (i in 1:N) {
    # Randomly shuffle the hats and compare to their owners
    shuffled_hats <- sample(hats)

    # In how many cases did at least one person get their hat back?
    trial_results[i] <- sum(shuffled_hats == owners) >= 1
}

# How many times, over all trials, did at least one person get their hat back?
k <- sum(trial_results)

# Convert to a proportion.
kk <- k / N

# Print the result.
print(kk)
```

We see that in roughly `r round(get_var('kk') * 100)` percent of the trials at
least one Santa received their own hat back.

### Example: Twenty executives assigned to two divisions of a firm

The top manager wants to spread the talent reasonably evenly, but she does not
want to label particular executives with a quality rating and therefore
considers distributing them with a random selection. She therefore wonders:
What are probabilities of the best ten among the twenty being split among the
divisions in the ratios 5 and 5, 4 and 6, 3 and 7, etc., if their names are
drawn from a hat? One might imagine much the same sort of problem in choosing
two teams for a football or baseball contest.

One may proceed as follows:

1.  Put 10 balls labeled "W" (for "worst") and 10 balls labeled "B"
    (best) in a bucket.
2.  Draw 10 balls without replacement and count the W's.
3.  Repeat (say) 400 times.
4.  Count the number of times each split — 5 W's and 5 B's, 4 and 6, etc. —
    appears in the results.

The problem can be done with {{< var lang >}} as follows:

```{python}
N = 10000
trial_results = np.zeros(N)

managers = np.repeat(['Worst', 'Best'], [10, 10])

for i in range(N):
    chosen = rnd.choice(managers, size=10, replace=False)
    trial_results[i] = np.sum(chosen == 'Best')

plt.hist(trial_results, bins=range(10), align='left', rwidth=0.75)
plt.title('Number of best managers chosen')
```

```{r}
N <- 10000
trial_results <- numeric(N)

managers <- rep(c('Worst', 'Best'), c(10, 10))

for (i in 1:N) {
    chosen <- sample(managers, 10)  # replace=FALSE is the default.
    trial_results[i] <- sum(chosen == 'Best')
}

hist(trial_results, breaks=0:max(trial_results),
     main= 'Number of best managers chosen')
```

### Example: Executives Moving

<!---
**NOTE:** Same as couples example?
-->

A major retail chain moves its store managers from city to city every three
years in order to calculate individuals' knowledge and experience. To make the
procedure seem fair, the new locations are drawn at random. Nevertheless, the
movement is not popular with managers' families. Therefore, to make the system
a bit sporting and to give people some hope of remaining in the same location,
the chain allows managers to draw in the lottery the same posts they are now
in. What are the probabilities that 1, 2, 3 ... will get their present posts
again if the number of managers is 30?

The problem can be solved with the following steps:

1.  Number a set of green balls from "1" to "30" and put them into Bucket A.
    Number a set of red balls from "1" to "30" and then put into Bucket B. For
    greater concreteness one could use 30 little numbered dolls in Bucket A and
    30 little toy houses in Bucket B.
2.  Shuffle Bucket A, and array all its green balls into a row (vector
    A). Array all the red balls from Bucket B into a second row B just
    below row A.
3.  Count how many green balls in row A have the same numbers as the
    red balls just below them, and record that number on a
    scoreboard.
4.  Repeat steps 2 and 3 perhaps 1000 times. Then count in the
    scoreboard the numbers of "0," "1," "2," "3."

### Example: State Liquor Systems Again

Let's end this chapter with the example of state liquor systems that we
first examined in @sec-infinite-universes and which will be discussed again
later in the context of problems in statistics.

Remember that as of 1963, there were 26 U.S. states in whose liquor systems the
retail liquor stores are privately owned ("Private"), and 16 *monopoly* states
where the state government owns the retail liquor stores ("Government").  See
@tbl-whiskey-prices for the prices in the Private and Government states.

We found the average prices were:

```{python, echo=FALSE, eval=TRUE}
import numpy as np
import pandas as pd

price_df = pd.read_csv('data/liquor_prices.csv')
priv_avg, govt_avg = price_df.groupby('state_type').mean().values.ravel()
gov_priv_diff = govt_avg - priv_avg
r_p_a, r_g_a, r_diff = np.round([priv_avg, govt_avg, gov_priv_diff], 2)
```

* Private: \$`r py$r_p_a`;
* Government: \$`r py$r_g_a`;
* Difference (Government - Private): \$`r py$r_diff`.

Let us now consider that all these states' prices constitute one single
*finite* universe. We ask: If these 42 states constitute a universe, and if
they are all shuffled together, how likely is it that if one divides them into
two samples at random (sampling without replacement), containing 16 and 26
observations respectively, the difference in mean prices turns out to be as
great as \$`r py$r_diff` (the difference that was actually observed)?

Again we write each of the forty-two observed state prices on a separate card.
The shuffled deck simulates a situation in which each state has an equal chance
for each price. Repeatedly deal groups of 16 and 26 cards, *without replacing*
the cards as they are chosen, to simulate hypothetical monopoly-state and
private-state samples. In each trial calculate the difference in mean prices.

The steps more systematically:

* **Step A.** Write each of the 42 prices on a card and shuffle.
* **Steps B and C (combined in this case).** i) Draw cards randomly without
  replacement into groups of 16 and 26 cards. Then ii) calculate the mean price
  difference between the groups, and iii) compare the simulation-trial
  difference to the observed mean difference of \$`r py$r_g_a` - \$`r py$r_p_a`
  = \$`r py$r_diff`; if it is as great or greater than \$`r py$r_diff`, write
  "yes," otherwise "no."
* **Step D.** Repeat step B-C a hundred or a thousand times. Calculate the
  proportion "yes," which estimates the probability we seek.

The probability that the postulated universe would produce a difference
between groups as large or larger than observed in 1961 is estimated by
how frequently the mean of the group of randomly-chosen sixteen prices
from the simulated state ownership universe is less than (or equal to)
the mean of the actual sixteen state-ownership prices.

Please notice how the only difference between this treatment of the problem and
the treatment in @sec-infinite-universes is that the drawing in this case is
without replacement whereas in @sec-infinite-universes the drawing is with
replacement.

In @sec-infinite-universes we thought of these states as if they came from a
non-finite universe, which is one possible interpretation in one context. But
one can also reasonably think about them in another context — as if they
constitute the entire universe (aside from those states excluded from the
analysis because of data complexities). If so, one can ask: If these 42 states
constitute a universe, how likely is it that one would choose two samples at
random, containing 16 and 26 observations, that would have prices as different
as \$.49 (the difference that was actually observed)?

### Example: Five or More Spades in One Bridge Hand; Four Girls and a Boy {#sec-five-spades-four-girls}

This is a compound problem: what are the chances of *both* five or more spades
in one bridge hand, and four girls and a boy in a five-child family?

"Compound" does not necessarily mean "complicated". It means that the
problem is a compound of two or more simpler problems.

A natural way to handle such a compound problem is in stages, as we saw
in the archery problem of @sec-one-black-archery. If a "success" is achieved in
the first stage, go on to the second stage; if not, don't go on. More
specifically in this example:

* **Step 1.** Use a bridge card deck, and five coins with heads = "girl".
* **Step 2.** Deal a 13-card bridge hand and count the spades. If 5 or more
  spades, record "no" and end the experimental trial. Otherwise, continue to
  step 3.
* **Step 3.** Throw five coins, and count "heads." If four heads, record "yes,"
  otherwise record "no."
* **Step 4.** Repeat steps 2 and 3 a thousand times.
* **Step 5.** Compute the proportion of "yes" in step 3. This estimates the
  probability sought.

The {{< var lang >}} solution to this compound problem is neither long nor
difficult. We tackle it almost as if the two parts of the problem were to be
dealt with separately. We first determine, in a random bridge hand, whether 5
spades or more are dealt, as was done in the problem
@sec-five-spades-four-clubs. Then, `if` 5 or more spades are found, we use {{<
var sample >}} to generate a random family of 5 children. This means that we
need not generate families if 5 or more spades were not dealt to the bridge
hand, because a "success" is only recorded if both conditions are met. After we
record the number of girls in each sample of 5 children, we need only finish
the loop (by [`}`]{.r}[unindenting the next line]{.python} and then use
{{< var sum >}} to count the number of samples that had 4 girls, storing the
result in `k`. Since we only drew samples of children for those trials in which
a bridge hand of 5 spades had already been dealt, `k` will have the number of
trials out of 10000 in which both conditions were met.

```{python}
N = 10000
trial_results = np.zeros(N)

# Deck with 13 spades and 39 other cards
deck = np.repeat(['spade', 'others'], [13, 52 - 13])

for i in range(N):
    # Shuffle deck and draw 13 cards
    hand = rnd.choice(deck, size=13, replace=False)

    n_spades = np.sum(hand == 'spade')

    if n_spades >= 5:
        # Generate a family, zeros for boys, ones for girls
        children = rnd.choice(['girl', 'boy'], size=5)
        n_girls = np.sum(children == 'girl')
        trial_results[i] = n_girls

k = np.sum(trial_results == 4)

kk = k / N

print(kk)
```

```{r}
N <- 10000
trial_results <- numeric(N)

# Deck with 13 spades and 39 other cards
deck <- rep(c('spade', 'others'), c(13, 52 - 13))

for (i in 1:N) {
    # Shuffle deck and draw 13 cards
    hand <- sample(deck, 13)  # replace=FALSE is default

    n_spades <- sum(hand == 'spade')

    if (n_spades >= 5) {
        # Generate a family, zeros for boys, ones for girls
        children <- sample(c('girl', 'boy'), 5, replace=TRUE)
        n_girls <- sum(children == 'girl')
        trial_results[i] <- n_girls
    }
}

k <- sum(trial_results == 4)

kk <- k / N

print(kk)
```

Here is an alternative approach to the same problem, but getting the result at
the end of the loop, by combining Boolean {{< var array >}}s (see
@sec-combine-booleans).

```{python}
N = 10000
trial_spades = np.zeros(N)
trial_girls = np.zeros(N)

# Deck with 13 spades and 39 other cards
deck = np.repeat(['spade', 'other'], [13, 39])

for i in range(N):
    # Shuffle deck and draw 13 cards
    hand = rnd.choice(deck, 13, replace=False)

    n_spades = np.sum(hand == 'spade')
    trial_spades[i] = n_spades

    # Generate a family, zeros for boys, ones for girls
    children = rnd.choice(['girl', 'boy'], size=5)
    n_girls = np.sum(children == 'girl')
    trial_girls[i] = n_girls

k = np.sum((trial_spades >= 5) & (trial_girls == 4))

kk = k / N

print(kk)
```

```{r}
N <- 10000
trial_spades <- numeric(N)
trial_girls <- numeric(N)

# Deck with 13 spades and 39 other cards
deck <- rep(c('spade', 'other'), c(13, 39))

for (i in 1:N) {
    # Shuffle deck and draw 13 cards
    hand <- sample(deck, 13)  # replace=FALSE is default
    # Count and store the number of spades.
    n_spades <- sum(hand == 'spade')
    trial_spades[i] <- n_spades

    # Generate a family, zeros for boys, ones for girls
    children <- sample(c('girl', 'boy'), 5, replace=TRUE)
    # Count and store the number of girls.
    n_girls <- sum(children == 'girl')
    trial_girls[i] <- n_girls
}

k <- sum((trial_spades >= 5) & (trial_girls == 4))

kk <- k / N

# Show the result
message(kk)
```

:::{.callout-note}
## Speed and readability

The last version is a fraction more expensive, but has the advantage that the
condition we are testing for is summarized on one line.  However, this would
not be a good approach to take if the experiments were not completely
unrelated.
:::

:::

## Summary

This completes the discussion of problems in probability — that is, problems
where we assume that the structure is known. Whereas @sec-infinite-universes
dealt with samples drawn from universes considered *not finite* , this chapter
deals with problems drawn from *finite universes* and therefore you *sample
without replacement*.
