---
jupyter:
  jupytext:
    metadata_filter:
      notebook:
        additional: all
        excluded:
        - language_info
    text_representation:
      extension: .Rmd
      format_name: rmarkdown
      format_version: '1.0'
      jupytext_version: 0.8.6
  kernelspec:
    display_name: Python 3
    language: python
    name: python3
resampling_with:
    ed2_fname: 10-Chap-6
---

```{r setup, include=FALSE}
source("_common.R")
```

# Probability Theory, Part 2: Compound Probability

:::{.callout-warning}
## Draft page partially ported from original PDF

This page is an automated and partial import from the [original second-edition
PDF](https://resample.com/content/text/10-Chap-6.pdf).

We are in the process of updating this page for formatting, and porting any
code from the original [RESAMPLING-STATS
language](http://www.statistics101.net) to Python and R.

Feel free to read this version for the sense, but expect there to be multiple
issues with formatting.

We will remove this warning when the page has adequate formatting, and we have
ported the code.
:::


<!---
We need if statement here.  Introduce here or before.
-->

<!---
Machinery for one pair, two pairs, three of a kind.

np.unique, return_count
-->



## Introduction

In this chapter we will deal with what are usually called "probability
problems" rather than the "statistical inference problems" discussed in
later chapters. The difference is that for probability problems we begin
with a knowledge of the properties of the universe with which we are
working. (See Chapter 4 on the definition of resampling.)

Before we get down to the business of complex probabilistic problems in
this and the next two chapters, let's consider a couple of peculiar
puzzles which do not fit naturally into any chapter in this book, but
which are extremely valuable in showing the power of the Monte Carlo
simulation method.

## Puzzle Problems

**The treasure fleet recovered**

This problem is:

> A Spanish treasure fleet of three ships was sunk at sea off Mexico. One ship
> had a trunk of gold forward and another aft, another ship had a trunk of gold
> forward and a trunk of silver aft, while a third ship had a trunk of silver
> forward and another trunk of silver aft. Divers just found one of the ships
> and a trunk of gold in it. They are now taking bets about whether the other
> trunk found on the same ship will contain silver or gold. What are fair odds?

(This is a restatement of a problem that Joseph Bertrand posed early in
the 19th century.) In the Goldberg variation:

> Three identical boxes each contain two coins. In one box both are pennies, in
> the second both are nickels, and in the third there is one penny and one
> nickel.
>
> A man chooses a box at random and takes out a coin. If the coin is a penny,
> what is the probability that the other coin in the box is also a penny?

These are the logical steps one may distinguish in arriving at a correct
answer with deductive logic (portrayed in @fig-ships-gold-silver).

1.  Postulate three ships — Ship I with two gold chests (G-G), ship II
    with one gold and one silver chest (G-S), and ship III with S-S.
    (Choosing notation might well be considered one or more additional
    steps.)

2.  Assert equal probabilities of each ship being found.

3.  Step 2 implies equal probabilities of being found for each of the
    six chests.

4.  Fact: Diver finds a chest of gold.

5.  Step 4 implies that S-S ship III was not found; hence remove it from
    subsequent analysis.

6.  Three possibilities: 6a) Diver found chest I-Ga, 6b) diver found
    I-Gb, 6c) diver found II-Gc.

    From step 2, the cases a, b, and c in step 6 have equal
    probabilities.

7.  If possibility 6a is the case, then the other trunk is I-Gb; the
    comparable statements for cases 6b and 6c are I-Ga and II-S.

8.  From steps 6 and 7: From equal probabilities of the three cases, and
    no other possible outcome, $P(6a) = 1/3$, $P(6b) = 1/3$, $P(6c) =
    1/3$.

9.  So $P(G) = P(6a) + P(6b)$ = 1/3 + 1/3 = 2/3.

See @fig-ships-gold-silver.

```{r fig-ships-gold-silver, echo=FALSE, fig.align='center', fig.cap="Ships with Gold and Silver"}
include_svg('images/ships_gold_silver.svg')
```

The following simulation arrives at the correct answer:

1.  Construct three buckets containing the numbers "7,7," "7,8," and "8,8"
    respectively.

2.  Choose a bucket at random, and shuffle the numbers in it.

3.  Choose the first element in the chosen bucket's vector (a vector is an
    array or list of numbers). If "8," stop trial and make no further
    record. If "7," continue.

4.  Record the second element in the chosen bucket's vector on the
    scoreboard.

5.  Repeat steps (2 - 5), and calculate the proportion "7's" on a
    scoreboard. (The answer should be about 2/3.)

Here is a computer simulation with {{< var lang >}}:

<!---
We need strings at this point.
-->

::: {.notebook name="gold_silver_ships" title="Ships with gold and silver"}

::: nb-only
In which we solve the problem of gold and silver chests in a discovered ship.
:::

```{python, opts.label="py_ed"}
import random
import numpy as np
```

```{python, opts.label="py_ed"}
# The 3 boxes, where 1 means gold, 0 means silver
ships = np.array([
    ['gold', 'gold'],
    ['gold', 'silver'],
    ['silver', 'silver']
])
# Show what we have.
print(ships)
```


```{python, opts.label="py_ed"}
# Mark trials as not valid to start with.
# Trials where we don't get a gold trunk first will
# keep this 'invalid' marker.
second_trunks = np.repeat(['invalid'], 10000)

for i in range(10000):
    # Select a ship at random from the three ships.
    ship_no = np.random.choice([0, 1, 2])
    # Get the trunks from this ship
    trunks = ships[ship_no, :]
    # Randomize the order of the trunks.
    np.random.shuffle(trunks)
    if trunks[0] == 'gold':  # We found a gold trunk first.
        # Store whether the second trunk was silver or gold.
        second_trunks[i] = trunks[1]

# Number of times we found gold in the second trunk.
n_golds = np.count_nonzero(second_trunks == 'gold')
# Number of times we found silver in the second trunk.
n_silvers = np.count_nonzero(second_trunks == 'silver')
# As a ratio of golds to all second trunks (where the first was gold).
print(n_golds / (n_golds + n_silvers))
```

:::

**The three-door problem**

***Consider the famous problem of the three doors:*** The player faces
three closed containers, one containing a prize and two empty. After the
player chooses, s/he is shown that one of the other two containers is
empty. The player is now given the option of switching from her/his
original choice to the other closed container. Should s/he do so?

***Answer:*** Switching doubles the chances of winning.

When this problem was published in the Sunday newspapers across the
U.S., the thousands of letters — including a good many from Ph.D.s in
mathematics — show that logical mathematical deduction fails badly in
this case. Most people — both laypersons and statisticians — arrive at
the wrong answer.

Simulation, however — and *hands-on* simulation with physical symbols,
rather than computer simulation — is a surefire way of obtaining and
displaying the correct solution. Table 6-1 shows such a simple
simulation with a random-number table. Column 1 represents the container
you choose, column 2 where the prize is. Based on columns 1 and 2,
column 3 indicates the container that the "host" would now open and show
to be empty. Lastly, column 4 scores whether the "switch" or "remain"
strategy would be preferable. A count of the number of winning cases for
"switch" and the "remain" gives the result sought.

Not only is the best choice obvious with this simulation method, but you
are likely to understand quickly why switching is better. No other mode
of explanation or solution brings out this intuition so well. And it is
much the same with other problems in probability and statistics.
Simulation can provide not only answers but also insight into why the
process works as it does. In contrast, formulas frequently produce
obfuscation and confusion for most non-mathematicians.

Table 6-1

**Simple Simulation With a Random-Number Table**

+-----------+-----------+-----------+-----------+-----------+-----------+
| **Random  | **Actual  | **Host    | **Winning |           |           |
| Pick**    | Location* | Opens**   | Strategy* |           |           |
|           | *         |           | *         |           |           |
| Random    |           |           |           |           |           |
| Equiv to  | Random    |           |           |           |           |
| Number    | Equiv to  |           |           |           |           |
| Door      | Number    |           |           |           |           |
|           | Door      |           |           |           |           |
+-----------+-----------+-----------+-----------+-----------+-----------+
| 1 0       | 1         | 0 *1*     | 1         | 2 or 3    | Remain    |
+-----------+-----------+-----------+-----------+-----------+-----------+
| 2 2       | 2         | 2 5       | 2         | 1 or 3    | R         |
+-----------+-----------+-----------+-----------+-----------+-----------+
| 2 4       | 2         | 2 2       | 2         | 1 or 3    | R         |
+-----------+-----------+-----------+-----------+-----------+-----------+
| 4 2       | 1         | 0 *6*     | 3         | 2         | Change    |
+-----------+-----------+-----------+-----------+-----------+-----------+
| 3 7       | 3         | 8 1       | 2         | 1         | C         |
+-----------+-----------+-----------+-----------+-----------+-----------+
| 7 7       | 1         | 1 1       | 1         | 2 or 3    | R         |
+-----------+-----------+-----------+-----------+-----------+-----------+
| 9 9       | 3         | 5 6       | 2         | 1         | C         |
+-----------+-----------+-----------+-----------+-----------+-----------+
| 9 6       | 3         | 0 *5*     | 2         | 1         | C         |
+-----------+-----------+-----------+-----------+-----------+-----------+
| 8 9       | 2         | 6 3       | 3         | 1         | C         |
+-----------+-----------+-----------+-----------+-----------+-----------+
| 8 5       | 2         | 4 3       | 1         | 3         | C         |
+-----------+-----------+-----------+-----------+-----------+-----------+
| 2 8       | 2         | 8 8       | 2         | 1 or 3    | R         |
+-----------+-----------+-----------+-----------+-----------+-----------+
| 6 3       | 3         | 4 8       | 1         | 2         | C         |
+-----------+-----------+-----------+-----------+-----------+-----------+
| 0 *9*     | 3         | 5 2       | 2         | 1         | C         |
+-----------+-----------+-----------+-----------+-----------+-----------+
| 1 0       | 1         | 8 7       | 2         | 3         | C         |
+-----------+-----------+-----------+-----------+-----------+-----------+
| 7 4       | 1         | 7 1       | 1         | 2 or 3    | R         |
+-----------+-----------+-----------+-----------+-----------+-----------+
| 5 1       | 2         | 5 1       | 2         | 1 or 3    | R         |
+-----------+-----------+-----------+-----------+-----------+-----------+
| 0 *2*     | 2         | 5 2       | 2         | 1 or 3    | R         |
+-----------+-----------+-----------+-----------+-----------+-----------+
| 0 *1*     | 1         | 3 3       | 3         | 2         | C         |
+-----------+-----------+-----------+-----------+-----------+-----------+
| 5 2       | 2         | 4 6       | 1         | 3         | C         |
+-----------+-----------+-----------+-----------+-----------+-----------+
| 0 *7*     | 1         | 3 9       | 3         | 2         | C         |
+-----------+-----------+-----------+-----------+-----------+-----------+
| 4 8       | 1         | 8 5       | 2         | 3         | C         |
+-----------+-----------+-----------+-----------+-----------+-----------+

Note: Underlining indicates numbers used. Zeros are omitted; numbers 1,
4, 7 = 1; 2, 5, 8 = 2; 3, 6, 9 = 3

## Examples of basic problems in probability

**A Poker Problem: One Pair (Two of a Kind)**

What is the chance that the first five cards chosen from a deck of 52
(bridge/poker) cards will contain two (and only two) cards of the same
denomination (two 3's for example)? (Please forgive the rather sterile
unrealistic problems in this and the other chapters on probability. They
reflect the literature in the field for 300 years. We'll get more
realistic in the statistics chapters.)

We shall estimate the odds the way that gamblers have estimated gambling
odds for thousands of years. First, check that the deck is not a
pinochle deck and is not missing any cards. (Overlooking such small but
crucial matters often leads to errors in science.) Shuffle thoroughly
until you are satisfied that the cards are randomly distributed. (It is
surprisingly hard to shuffle well.) Then deal five cards, and mark down
whether the hand does or does not contain a pair of the same
denomination. At this point, we must decide whether three of a kind,
four of a kind or two pairs meet our criterion for a pair. Since our
criterion is "two and only two," we decide *not* to count them.

Then replace the five cards in the deck, shuffle, and deal again. Again
mark down whether the hand contains one pair of the same denomination.
Do this many times. Then count the number of hands with one pair, and
figure the proportion (as a percentage) of all hands. In one series of
100 experiments, 44 percent of the hands contained one pair, and
therefore .44 is our estimate (for the time being) of the probability
that one pair will turn up in a poker hand. But we must notice that this
estimate is based on only 100 hands, and therefore might well be fairly
far off the mark (as we shall soon see).

This experimental "resampling" estimation does not require a deck of
cards. For example, one might create a 52-sided die, one side for each
card in the deck, and roll it five times to get a "hand." But note one
important part of the procedure: No single "card" is allowed to come up
twice in the same set of five spins, just as no single card can turn up
twice or more in the same hand. If the same "card" did turn up twice or
more in a dice experiment, one could pretend that the roll had never
taken place; this procedure is necessary to make the dice experiment
analogous to the actual card-dealing situation under investigation.
Otherwise, the results will be slightly in error. This type of sampling
is known as "sampling without replacement," because each card is *not
replaced* in the deck prior to dealing the next card (that is, prior to
the end of the hand).

Table 6-2

**Results of 100 Trials for the Problem "OnePair"**

  ----------- --- --- --- --- --- --- --- --- ---- ------- ----
  Trial 1     2   3   4   5   6   7   8   9   10   11 12   13
  Results Y   Y   N   N   Y   Y   N   N   Y   N    N Y     Y
  ----------- --- --- --- --- --- --- --- --- ---- ------- ----

 Trial 14 15 16 17 18 19 20 21 22 23 24 25 26 Results N Y Y Y Y Y N N Y
N Y N Y

  ----------- ---- ---- ---- ---- ---- ---- ---- ---- ---- ------- ----
  Trial 27    28   29   30   31   32   33   34   35   36   37 38   39
  Results N   Y    N    Y    Y    N    Y    N    N    N    N Y     N
  Trial 40    41   42   43   44   45   46   47   48   49   50      
  Results N   N    N    N    Y    Y    Y    N    N    Y    N       
  ----------- ---- ---- ---- ---- ---- ---- ---- ---- ---- ------- ----

 Subtotal: 23 Yes, 27 No = 46%

  ----------- ---- ---- ---- ---- ---- ---- ---- ---- ---- ----- ---- ----
  Trial 51    52   53   54   55   56   57   58   59   60   61    62   63
  Results N   Y    N    N    Y    N    Y    Y    N    N    N     Y    Y
  Trial 64    65   66   67   68   69   70   71   72   73   74    75   76
  Results Y   N    N    Y    N    N    N    N    Y    N    Y     N    N
  Trial 77    78   79   80   81   82   83   84   85   86   87    88   89
  Results N   N    N    N    Y    N    N    N    Y    Y    N     Y    N
  Trial 90    91   92   93   94   95   96   97   98   99   100        
  Results Y   Y    N    N    Y    Y    Y    Y    N    Y    N          
  ----------- ---- ---- ---- ---- ---- ---- ---- ---- ---- ----- ---- ----

 Subtotal: 21 Yes, 29 No = 42%

Total: 44 Yes, 56 No = 44%

Still another resampling method uses a *random number table* , such as
that which is shown in Table 6-3. Arbitrarily designate the spades as
numbers "01-13," the diamonds as "14-26," the hearts as "27-39," and the
clubs as "40-52." Then proceed across a row (or down a column), writing
down each successive pair of digits, excluding pairs outside "01-52" and
omitting duplication within sets of five numbers. Then translate them
back into cards, and see how many "hands" of five "cards" contain one
pair each. Table 6-4 shows six such hands, of which hands numbered 2, 3
and 6 contain pairs.

Table 6-3

**Table of Random Digits**

  ---- ---- ---- ---- ---- ---- ---- ---- ---- ---- ---- ---- ---- ---- ---- ----
  48   52   78   38   11   90   41   83   43   99   51   55   57   03   83   20
  15   11   84   33   09   24   08   52   42   70   37   16   66   73   15   54
  25   89   70   11   91   65   41   90   88   04   30   72   15   81   34   46
  34   24   66   55   67   79   29   18   36   56   96   95   35   06   05   10
  37   27   58   38   23   84   94   39   99   50   74   80   41   85   98   63
  12   17   04   68   19   98   53   44   16   32   91   01   71   60   19   12
  88   85   44   65   52   01   99   56   72   07   96   39   56   34   86   01
  81   92   77   83   10   58   92   33   63   48   62   66   32   61   59   74
  08   50   15   18   13   45   65   12   32   92   53   82   07   61   71   80
  84   29   90   36   05   95   20   71   17   82   83   38   01   87   74   92
  77   76   46   28   47   15   04   21   04   75   51   83   91   37   14   32
  01   33   90   94   86   10   03   99   95   98   76   97   97   26   45   62
  ---- ---- ---- ---- ---- ---- ---- ---- ---- ---- ---- ---- ---- ---- ---- ----

 Table 6-4

**Six Simulated Trials for the Problem "OnePair"**

**Aces Deuces 3 4 5 6 7 8 9 10 J Q K**

**What the Random Numbers Mean**

  ------------------------ ---- ---- ---- ---- ---- ---------- ---- ---- ---- ---- ---- ---- ----
  Spades                   01   02   03   04   05   06         07   08   09   10   11   12   13
  Diamonds                 14   15   16   17   18   19         20   21   22   23   24   25   26
  Hear ts                  27   28   29   30   31   32         33   34   35   36   37   38   39
  Clubs                    40   41   42   43   44   45         46   47   48   49   50   51   52
  **Simulation Results**                                                                     
  Hand 1:                  48   52   38   11   41   no pairs                                 
  Hand 2:                  15   11   33   09   24   one pair                                 
  Hand 3:                  25   11   41   04   30   one pair                                 
  Hand 4:                  34   24   29   18   36   no pairs                                 
  Hand 5:                  37   27   38   23   39   no pairs                                 
  Hand 6:                  12   17   04   19   44   one pair                                 
  ------------------------ ---- ---- ---- ---- ---- ---------- ---- ---- ---- ---- ---- ---- ----

Now let's do the same job using RESAMPLING STATS on the computer. Let's
name "One Pair" the file which simulates a deck of playing cards and
solves the problem.

Our first task is to simulate a deck of playing cards analogous to the
real cards we used previously. We don't need to simulate all the
features of a deck, but only the features that matter for the problem at
hand. We require a deck with four "1"s, four "2"s, etc., up to four
"13"s. The suits don't matter for our present purposes. Therefore, with
the URN command we join together in a single array the four sets of
thirteen numbers, to represent the 13 denominations.

At this point we have a complete deck in location A. But that "deck" is
in the same order as a new deck of cards. If we do not shuffle the deck,
the results will be predictable. Therefore, we write SHUFFLE a b and
then deal a poker hand by taking the first five cards from the shuffled
hand, using the TAKE statement. Now we must find out if there is one
(and only one) pair; we do this with the MULTIPLES statement — the "2"
in that statement indicates that it is a duplicate, rather than a
singleton or triplicate or quadruplicate that we are testing for — and
we put the result in location D. Next we SCORE in location z how many
pairs there are, the number in each trial being either zero, one, or
two. (The reason we cannot put the result of the MULTIPLES operation
directly into the scorecard vector z is that only the SCORE command
accumulates results from trial to trial rather than over-writing the
result of the past trial with the current one.) And with that we end a
single trial.

With the REPEAT 1000 statement and the END statement, we command the
program to repeat a thousand times the statements in the "loop" between
those two lines. When those 1000 repetitions are over, the computer
moves on to COUNT the number of "1's" in SCOREkeeping vector z, each "1"
indicating a hand with a pair. And we then PRINT to the screen the
result which is found in location k. If we want the *proportion* of the
trials in which a pair occurs, we simply divide the results of the
thousand trials by 1000.


```{python, opts.label="py_ed"}
# Create a bucket (vector) called a with four "1's," four "2's," four "3's,"
# etc., to represent a deck of cards
one_suit = np.arange(1, 14)
print(one_suit)
```

```{python, opts.label="py_ed"}
deck = np.repeat(one_suit, 4)
print(deck)
```

```{python, opts.label="py_ed"}
pairs_per_trial = np.zeros(1000)

# Repeat the following steps 1000 times
for i in range(1000):
    # Shuffle the deck
    np.random.shuffle(deck)

    # Take the first five cards.
    hand = deck[:5]

    # How many pairs?
    # Counts for each card rank.
    counts = np.bincount(hand)
    n_pairs = np.count_nonzero(counts == 2)

    # Keep score of # of pairs
    pairs_per_trial[i] = n_pairs

    # End loop, go back and repeat

# How often was there 1 pair?
n_one_pair = np.count_nonzero(pairs_per_trial == 1)

# Convert to proportion
print(n_one_pair / 1000)
```

Note: The file "onepair" on the Resampling Stats software disk contains
this set of commands.

In one run of the program, the result in kk was .419, so our estimate
would be that the probability of a single pair is .419.

How accurate are these resampling estimates? The accuracy depends on the
*number of hands* we deal — the more hands, the greater the accuracy. If
we were to examine millions of hands, 42 percent would contain a pair
each; that is, the chance of getting a pair in the long run is 42
percent. The estimate of 44 percent based on 100 hands in Table 6-2 is
fairly close to the long-run estimate, though whether or not it is close
*enough* depends on one's needs of course. If you need great accuracy,
deal many more hands.

A note on the "a"s, "b"s, "c"s in the above program, etc.: These
"variables" are called "vectors" in RESAMPLING STATS. A *vector* is an
array of elements that gets filled with numbers as RESAMPLING STATS
conducts its operations. When RESAMPLING STATS completes a trial these
vectors are generally wiped clean except for the "SCORE" vector (here
labeled "z") which keeps track of the result of each trial.

To help keep things straight (though the program does not require it),
we usually use "z" to name the vector that collects all the trial
results, and "k" to denote our overall summary results. Or you could
call it something like "scrbrd."

How many trials (hands) should be made for the estimate? There is no
easy answer.[^how-many-trials] One useful device is to run several (perhaps
ten) equal sized sets of trials, and then examine whether the proportion of
pairs found in the entire group of trials is very different from the
proportions found in the various subgroup sets. If the proportions of pairs in
the various subgroups differ greatly from one another or from the overall
proportion, then keep running additional larger subgroups of trials until the
variation from one subgroup to another is sufficiently small for your
purposes. While such a procedure would be impractical using a deck of cards or
any other physical means, it requires little effort with the computer and
RESAMPLING STATS.

[^how-many-trials]: One simple rule-of-thumb is to quadruple the original
    number. The reason for quadrupling is that four times as many iterations
    (trials) of this resampling procedure give *twice* as much accuracy (as
    measured by the standard deviation, the most frequent measurement of
    accuracy). That is, the error decreases with the square root of the number
    of iterations. If you see that you need *much* more accuracy, then
    *immediately* increase the number of iterations even more than four times
    — perhaps ten or a hundred times.

**Another Introductory Poker Problem**

Which is more likely, a poker hand with two pairs, or a hand with three
of a kind? This is a *comparison* problem, rather than a problem in
*absolute* estimation as was the previous example.

In a series of 100 "hands" that were "dealt" using random numbers, four
hands contained two pairs, and two hands contained three of a kind. Is
it safe to say, on the basis of these 100 hands, that hands with two
pairs are more frequent than hands with three of a kind? To check, we
deal another 300 hands. Among them we see fifteen hands with two pairs
(3.75 percent) and eight hands with three of a kind (2 percent), for a
total of nineteen to ten. Although the difference is not enormous, it is
reasonably clear-cut. Another 400 hands might be advisable, but we shall
not bother.

Earlier I obtained forty-four hands with *one* pair each out of 100
hands, which makes it quite plain that *one* pair is more frequent than
*either* two pairs or three-of-a-kind. Obviously, we need *more* hands
to compare the odds in favor of two pairs with the odds in favor of
three-of-a-kind than to compare those for one pair with those for either
two pairs or three-of-a-kind. Why? Because the difference in odds
between one pair, and either two pairs or three-of-a-kind, is much
greater than the difference in odds between two pairs and
three-of-a-kind. This observation leads to a general rule: The closer
the odds between two events, the *more trials* are needed to determine
which has the higher odds.

Again it is interesting to compare the odds with the formulaic
mathematical computations, which are 1 in 21 (4.75 percent) for a hand
containing two pairs and 1 in 47 (2.1 percent) for a hand containing
three-of-a-kind — not too far from the estimates of .0375 and .02
derived from simulation.

To handle the problem with the aid of the computer, we simply need to
estimate the proportion of hands having triplicates and the proportion
of hands with two pairs, and compare those estimates.

To estimate the hands with three-of-a-kind, we can use a program just
like "One Pair" earlier, except instructing the MUL- TIPLES statement to
search for triplicates instead of duplicates. The program, then, is:

```{python, opts.label="py_ed"}
triples_per_trial = np.zeros(1000)

# Repeat the following steps 1000 times
for i in range(1000):
    # Shuffle the deck
    np.random.shuffle(deck)

    # Take the first five cards.
    hand = deck[:5]

    # How many triples?
    counts = np.bincount(hand)
    n_triples = np.count_nonzero(counts == 3)

    # Keep score of # of triples
    triples_per_trial[i] = n_triples

    # End loop, go back and repeat

# How often was there 1 pair?
n_triples = np.count_nonzero(triples_per_trial == 1)

# Convert to proportion
print(n_triples / 1000)
```

Note: The file "3kind" on the Resampling Stats software disk contains
this set of commands.

To estimate the probability of getting a two-pair hand, we revert to the
original program (counting pairs), except that we examine all the
results in SCOREkeeping vector z for hands in which we had *two* pairs,
instead of *one* .

```{python, opts.label="py_ed"}
pairs_per_trial = np.zeros(1000)

# Repeat the following steps 1000 times
for i in range(1000):
    # Shuffle the deck
    np.random.shuffle(deck)

    # Take the first five cards.
    hand = deck[:5]

    # How many pairs?
    # Counts for each card rank.
    counts = np.bincount(hand)
    n_pairs = np.count_nonzero(counts == 2)

    # Keep score of # of pairs
    pairs_per_trial[i] = n_pairs

    # End loop, go back and repeat

# How often were there 2 pairs?
n_two_pairs = np.count_nonzero(pairs_per_trial == 2)

# Convert to proportion
print(n_two_pairs / 1000)
```

Note: The file "2pair" on the Resampling Stats software disk contains
this set of commands.

For efficiency (though efficiency really is not important here because
the computer performs its operations so cheaply) we could develop both
estimates in a single program by simply generating 1000 hands, and count
the number with three-of-a-kind and the number with two pairs.

Before we leave the poker problems, we note a difficulty with Monte
Carlo simulation. The probability of a royal flush is so low (about one
in half a million) that it would take much computer time to compute. On
the other hand, considerable inaccuracy is of little matter. Should one
care whether the probability of a royal flush is 1/100,000 or 1/500,000?

## The concepts of replacement and non-replacement

In the poker example above, we *did not replace* the first card we drew.
If we were to replace the card, it would leave the probability the same
before the second pick as before the first pick. That is, the
conditional probability remains the same. *If we replace, conditions do
not change.* But if we do not replace the item drawn, the probability
changes from one moment to the next. (Perhaps refresh your mind with the
examples in the discussion of conditional probability in Chapter 5.)

If we sample with replacement, the sample drawings remain

*independent* of each other — a topic addressed in Chapter 5.

In many cases, a key decision in modeling the situation in which we are
interested is whether to sample with or without replacement. The choice
must depend on the characteristics of the situation.

There is a close connection between the lack of finiteness of the
concept of universe in a given situation, and sampling with replacement.
That is, when the universe (population) we have in mind is not small, or
has no conceptual bounds at all, then the probability of each successive
observation remains the same, and this is modeled by sampling with
replacement. ("Not finite" is a less expansive term than "infinite,"
though one might regard them as synonymous.)

Chapter 7 discusses problems whose appropriate concept of a universe is
finite, whereas Chapter 8 discusses problems whose appropriate concept
of a universe is not finite. This general procedure will be discussed
several times, with examples included.
