# Resampling with code

Chapter \@ref(resampling-method) used simulation and resampling from
tables of random numbers, dice, and coins.  Making random choices in this way
can make it easier to understand the process, but of course, physical methods
of making random outcomes can be slow and boring.

We saw that short computer programs can do a huge number of resampling
trials in a few seconds.  The flexibility of a programming language makes it
possible to simulate many different outcomes and tests.

Programs can build up tables of random numbers, and do simple
tasks like counting the number of values in a row, and taking
proportions.  With these simple tools, we can simulate many problems
in probability and statistics.

In this chapter, we will again model a problem using `r book_lang`,
but this time we will explain all the constructs used in detail.

By the end of the chapter, you will be able to ...


## A treatment for Burkitt's lymphoma

In 2006, researchers came up with a new treatment for Burkitt's
lymphoma—a cancer of the lymphatic system.  In developed countries,
the cure rate for Burkitt's lymphoma is about 90%.  The new treatment
was applied to 17 patients, and all 17 patients were cured.

The first question that comes to mind is: how likely is that to happen
by chance?  In other words, is what we saw something special that
warrants further investigation, or is it pretty common for 17
randomly chosen patients to all be cured using a 90% effective
treatment?

Before we start, we import NumPy and ask for its random number generator:

```{python}
import numpy as np

# Ask for NumPy's default random number generator and name
# it `rnd`---short for "random"
rnd = np.random.default_rng()
```

Next, we need to come up with a physical model.  To simulate whether a
given patient is cured or not by a 90% effective treatment, we throw a
ten sided die:

```{python}
outcomes = ['cured', 'cured', 'cured',
            'cured', 'cured', 'cured',
            'cured', 'cured', 'cured',
            'not cured']
single_patient_outcome = rnd.choice(outcomes)
```

Here, we take all possible outcomes and then ask NumPy to pick one at random.
An alternative would be to ask for specific outcomes with given probabilities:

```python
single_patient_outcome = choice(['cured', 'not cured'], p=[0.9, 0.1])
```


Let us simulate a single trial, where seventeen patients are treated
with the 90% effective treatment:

```{python}
N = 17

cured = np.zeros(N, dtype=bool)
for i in range(N):
    outcome = rnd.choice(['cured', 'not cured'], p=[0.9, 0.1])
    cured[i] = outcome

all_cured = np.all(cured == 'cured')
```

Here, for each of the 17 patients, we draw 'cured' (90% probable) or 'not cured' (10% probable) from our outcomes.
We could just as easily draw all 17 at the same time:

```{python}
cured = rnd.choice(['cured', 'not cured'], p=[0.9, 0.1], size=17)
all_cured = np.all(cured == 'cured')
```

The above resembles one trial where 17 patients were treated.  We would like to count, over many trials, how often *all* patients were cured:

```{python}
T = 10000

# Array to store trial outcomes in
z = np.zeros(T, dtype=bool)

for i in range(T):
    cured = rnd.choice(['cured', 'not cured'], p=[0.9, 0.1], size=17)
    all_cured = np.all(cured == 'cured')

    z[i] = all_cured

# Determine in how many trials all patients were cured
k = np.count_nonzero(z)

# Convert to a proportion
kk = k / T

print(kk)
```

From this experiment, we see that there is roughly a one-in-six chance that all 17 patients are cured when using a 90% effective treatment.

---

Here is our first `r book_lang` example program.  Do not expect to
follow all of it straightaway.  For now, read the code below to get an
idea of how it implements the procedure above.  We will be coming back
to the specifics later.

The key to reading code is to think about what the computer will do, when it sees the code.

`r begin_nb('resampling_with_code')`

A simple place to start is the *comment*.  A comment is a statement that the
computer will *ignore*.   It is text that we put in the program for our
benefit, to explain what is going on to a human reader.

This is an example of a comment:

```{python, opts.label="py_ed"}
# This is a comment. It doesn't have any effect.
```

```{r, opts.label="r_ed"}
# This is a comment. It doesn't have any effect.
```

The comment starts with a hash character `#`.  This character tells the
computer that the rest of the line is a comment, and therefore, that it can
ignore everything on that line that follows the `#`.

`r begin_ispy()`

In the next bit of code, we load the *function* `randint` that
produces random numbers.

The code has four lines.  The first and third are comments; the
computer ignores them. The second line loads the NumPy library that we
use in nearly all our examples, and the last line loads the function
to generate random numbers.  After we run this code, we have a
function called `randint`, and a library called `np`.

```{python, opts.label="py_ed"}
# Import the numpy library
import numpy as np
# Get the function to generate random numbers.
from numpy.random import randint
```

`r end_ispy()`

To solve the problem, we have to generate some random numbers:

`r begin_ispy()`

```{python, opts.label="py_ed"}
a = randint(1, 11, 20)
```

This uses the `randint` *function* to generate random integers (counting numbers) from 1 up to, *but not including*, 11.  Therefore, this command will generate random numbers from 1 through 10.  The 20 in the command tells the function to generate 20 of these numbers.

`r end_ispy()`

`r begin_isr()`

```{r, opts.label="r_ed"}
a = sample(1:10, 20, replace=TRUE)
```

This uses the `sample` function to generate random integers (counting numbers)
from 1 through 10.  The 20 in the command tells R to generate 20 of these
numbers.  `replace=TRUE` tells R to sample *with replacement*.

For example, the chances of getting a particular result - such as a 3 - for the
first random number - are 1 in 10, or p= 1/10 = 0.1.  If we resample *with
replacement* then the chances of getting 3 in the second number are unchanged,
at p=0.1.  It is as if we put 10 balls into a bucket, numbered one through ten,
and then selected 20 balls from the bucket; but after we have selected a ball,
we record the number and *replace* it in the bucket, and shake up the bucket
again.

`r end_isr()`

Inasmuch as each ambulance has a 1 in 10 chance of being
defective, we decide arbitrarily that a "1" stands for a defective ambulance,
and the other nine numbers (from "2" to "10") stand for a not-defective
ambulance. The command above orders the computer to store the results of the random
drawing in a location in the computer's memory to which we give a name, in this case `a` (we could have chosen any name such as, for example, `ambulances`).  When we run a statement like the one above, `a` is
a *variable* - the *name* `a` refers to the *value*, which is the sequence of
random numbers the computer created using `r py_or_r("randint", "sample",
backticks=TRUE)`.

We can display the value of the variable `a` by using the `print` function:

```{python, opts.label="py_ed"}
print(a)
```

```{r, opts.label="r_ed"}
print(a)
```

This shows the 20 random values that we got from
`r py_or_r("randint", "sample")`.


The next key element in the core of the program is:

```{python, opts.label="py_ed"}
b = np.count_nonzero(a == 1)
```

```{r, opts.label="r_ed"}
b = sum(a == 1)
```

This command can be broken up into two pieces: the first (`a == 1`)
compares each element in `a` to 1, resulting in a True or False
value. We then proceed to count those values, to determine how many of
the 20 values in `a` are equal to 1. The result of the count will be
somewhere between 0 and 20, the number of ambulances that might be
out-of-order on a given day. The result is then placed in another
location in the computer's memory that we label `b`.

```{python, opts.label="py_ed"}
# Show the value of b
print(b)
```

```{r, opts.label="r_ed"}
# Show the value of b
print(b)
```

Now let us place the commands to generate the random numbers, and count how
many `1`s we get, within a program that solves this problem:

```{python, opts.label="py_ed"}
# Make an array that has 400 elements.
# We will use this to store the counts for our 400 repetitions
results = np.zeros(400)

# Repeat the simulation 400 times
for i in np.arange(400):

    # The indented commands are the procedure for one trial.
    # The computer runs these commands from first to last, for each trial.

    # Generate 20 numbers, each between "1" and "10," and put them in vector a.
    # Each number will represent an ambulance, and we let 1 represent
    # a defective ambulance.
    a = randint(1, 11, 20)

    # Count the number of defective ambulances, and put the result in b.
    b = np.count_nonzero(a == 1)

    # Keep track of each trial's result in "results".
    results[i] = b

    # End this trial, then go back and repeat the process until all 400 trials
    # are complete.

# Now we have finished the 400 trials.

# Determine how many trials resulted in more than 3 ambulances out of order.
bad_day_count = np.count_nonzero(results > 3)

# Convert to a proportion.
bad_day_prop = bad_day_count / 400

# Print the result.
print(bad_day_prop)
```

```{r, opts.label="r_ed"}
# Make an array that has 400 elements.
# We will use this to store the counts for our 400 repetitions
results <- numeric(400)

# Repeat the simulation 400 times
for (i in 1:400) {

    # The commands between the { above and the } below are the procedure for
    # one trial.
    # The computer runs these commands from first to last, for each trial.

    # Generate 20 numbers, each between "1" and "10," and put them in vector a.
    # Each number will represent an ambulance, and we let 1 represent
    # a defective ambulance.
    a <- sample(1:10, 20, replace=TRUE)

    # Count the number of defective ambulances, and put the result in b.
    b <- sum(a == 1)

    # Keep track of each trial's result in "results".
    results[i] <- b

    # End this trial, then go back and repeat the process until all 400 trials
    # are complete.
}

# Now we have finished the 400 trials.

# Determine how many trials resulted in more than 3 ambulances out of order.
bad_day_count <- sum(results > 3)

# Convert to a proportion.
bad_day_prop <- bad_day_count / 400

# Print the result.
print(bad_day_prop)
```

`r end_nb()`

The
`r py_or_r("results[i] = b", "results[i] <- b")`
statement that follows the
`r py_or_r("b = np.count_nonzero(a == 1)", "b = sum(a == 1)")`
operation keeps track of the results of each trial, placing the number
of defective ambulances that occur in each trial in a location that we
call "results". This is done in each of the 400 trials that we make, and the
result eventually is a "vector" with 400 numbers in it.  A vector is just
a sequence of numbers.

In order to make 400 repetitions of our experiment---we could have
decided to make a thousand or some other number of repetitions---we put
`r py_or_r("for i in np.arange(400):", "for (i in 1:400) {")`
before the statements that generate the random numbers, count how many of these numbers are 1, representing a defective ambulance, and store this result
of a single trial. Then we complete each repetition "loop" by `r begin_ispy()`
bringing the indentation of the statements back to the left margin
`r end_ispy()` `r begin_isr()`
adding a closing `}` bracket.
`r end_isr()`

Since our aim is to count the number of days in which more than 3 (4 or more)
defective ambulances occur, we use the
`r py_or_r("bad_day_count = np.count_nonzero(results > 3)", "bad_day_count = sum(results > 3)")`
command to count how many times in the 400 days recorded in our results vector
at the end of the 400 trials more than 3 defects occurred, and we place the
result in still another location "bad_day_count".

This gives us the total number of days where 4 or more defective ambulances are
seen to occur. Then we divide the number in "bad_day_count" by 400, the number
of trials. Thus we obtain an estimate of the chance, expressed as a probability
between 0 and 1, that 4 or more ambulances will be defective on a given day.
And we store that result in a location that we decide to call "bad_day_prop" so
that it will be there when the computer receives the next command to `print`
that result on the screen.

Can you see how each of the operations that the computer carries out are
analogous to the operations that you yourself executed when you solved this
problem using a labeled coins or a random-number table? This is exactly the
procedure that we will use to solve every problem in probability and statistics
that we must deal with. Either we will use a device such as coins or a random
number table as an analogy for the physical process we are interested in
(ambulances becoming defective, in this case), or we will simulate the analogy
on the computer using the `r book_lang` program above.

The program above may not seem simple at first glance, but we think
you will find, over the course of this book, that these programs
become much simpler to understand than the older conventional approach
to such problems that has routinely been taught to students for
decades.
