---
jupyter:
  jupytext:
    metadata_filter:
      notebook:
        additional: all
        excluded:
        - language_info
    text_representation:
      extension: .Rmd
      format_name: rmarkdown
      format_version: '1.0'
      jupytext_version: 0.8.6
  kernelspec:
    display_name: Python 3
    language: python
    name: python3
resampling_with:
    ed2_fname: null
---

```{r setup, include=FALSE}
source("_common.R")
```

# More resampling with code

@sec-resampling-code introduced a problem in probability, that was also a problem in statistics.  We asked how surprised we should be at the results of a trial of a new cancer treatment regime.

Here we study another urgent problem in the real world - racial bias and the
death penalty.

## A question of life and death

This example comes from the excellent Berkeley introduction to data science
[@adhikari2021data8].

Robert Swain was a young Black man who was sentenced to death in the early
60s.  Swain's trial was held in Talladega County, Alabama. At the time, 26% of
the eligible jurors in that county were Black, but every member of Swain's
jury was White.  Swain and his legal team appealed to the Alabama Supreme
Court, and then to the [US Supreme
Court](https://en.wikipedia.org/wiki/Swain_v._Alabama), arguing
that Black jurors were very rarely selected for juries in Talladega county, even though they made up about a quarter of the eligible pool of jurors.
The US Supreme Court rejected this argument, in a 6 to 3 opinion, writing that
"The overall percentage disparity has been small and reflects no studied
attempt to include or exclude a specified number of Negros.".

Swain's team presented a lot of evidence on bias in jury selection, but here we will look at the obvious and apparently surprising fact that Swain's jury was entirely White.
The Supreme Court decided that the "disparity" between selection of White and Black jurors "has been small" — but how would they, and how would we, make a rational decision about whether this disparity *really* was "small"?

## A small disparity and a hypothetical world

To answer the question that the Supreme Court asked, we return to the method we used in the last chapter.

Let us imagine a hypothetical world, in which each individual Black or White
person had an equal chance of being selected for the jury.  Call this world Hypothetical County, Alabama.

Just as in 1960's Talladega County, 26% of eligible jurors in Hypothetical
County are Black.  Hypothetical County jury selection has no bias against
Black people, so we expect around 26% of the jury to be Black. 0.26 * 12 =
3.12, so we expect that, on average, just over 3 out of 12 jurors in
a Hypothetical County jury will be Black.  But, if we select each juror at
random from the population, that means that, sometimes, by chance, we will have
fewer than 3 Black jurors, and sometimes will have more than 3 Black jurors.
And, by chance, sometimes we will have no Black jurors.  But, if the jurors
really are selected at random, how often would we expect this to happen — that
there are no Black jurors?  We would like to estimate the *probability* that
we will get no Black jurors.  If that probability is small, then we have some
evidence that the disparity in selection between Black and White jurors, was
not "small".

::: {.question}
What is the probability of an *all White* jury being randomly selected out of a population having 26% Black people?
:::

## Designing the experiment

Before we start, we need to figure out three things:

1. What do we mean by one trial?
2. What is the outcome of interest from the trial?
3. How do we simulate one trial?

We then take **three steps** to calculate the desired probability:

1. *Repeat* the simulated trial procedure N times.
2. *Count* M, the number of trials with an outcome that matches the outcome we
   are interested in.
3. Calculate the *proportion*, M/N.
   This is an estimate of the probability in question.

For this problem, our task is made a little easier by the fact that our *trial* (in the resampling sense) is a simulated *trial* (in the legal sense). One trial requires 12 simulated jurors, each labeled by race (White or Black).

The outcome we are interested in is the number of Black jurors.

Now comes the harder part.  How do we simulate one trial?

### One trial

One trial requires 12 jurors, and we are interested only in the race of each juror.
In Hypothetical County, where selection by race is entirely random, each juror
has a 26% chance of being Black.

We need a way of simulating a 26% chance.

One way of doing this is by getting a random number from 0 through 99
(inclusive).  There are 100 number in the range 0 through 99 (inclusive).

::: {.callout-note}
#### Pick a number from 1 through 5

Ranges can be confusing in normal speech because it is not always clear
whether they include their beginning and end.  For example, if someone says
"pick a number between 1 and 5", do they mean *all* the numbers, including the
first and last (any of 1 or 2 or 3 or 4 or 5)?  Or do they mean only the
numbers that are *between* 1 and 5 (so 2 or 3 or 4).  Or do they mean all the
numbers up to, but not including 5 (so 1 or 2 or 3 or 4).

To avoid this confusion, we will nearly always use "from" and "through" in
ranges, meaning that we do include both the start and the end number.  For
example, if we say "pick a number from 1 through 5" we mean one of 1 or 2 or 3
or 4 or 5.
:::

We will arbitrarily say that the juror is White if the random number is in the
range from 0 through 73.  74 of the 100 numbers are in this range, so
the juror has a 74/100 = 74% chance of getting the label "White".  We will say
the juror is Black if the random number is in the range 74 though 99.  There
are 26 such numbers, so the juror has a 26% chance of getting the label
"Black".

Next we need a way of getting a random number in the range 0 through 99. This
is an easy job for the computer, but if we had to do this with a physical
device, we could get a single number by throwing *two* 10-sided dice, say a
blue die and a green die.  The face of the blue die will be the 10s digit, and
the green face will be the ones digit.  So, if the blue die comes up with 8 and
the green die has 4, then the random number is 84.

We could then simulate 12 jurors by repeating this process 12 times, each time
writing down "White" if the number is from 0 through 74, and "Black"
otherwise.  The trial outcome is the number of times we wrote "Black" for these
12 simulated jurors.

### Using code to simulate a trial

We use the same logic to simulate a trial with the computer. A little code
makes the job easier, because we can ask {{< var lang >}} to give us 12 random
numbers from 0 through 99, and to count how many of these numbers are in the
range from 75 through 99.  Numbers in the range from 75 through 99 correspond
to Black jurors.

:::: {.callout-note}
#### Ranges in {{< var lang >}}

We said above that ranges can be confusing, and we use "from 1 through 5" to avoid this confusion.

::: {.python}
Python and NumPy also have the concept of a range of integers (positive or
negative whole numbers).  Here is a NumPy range:

```{python, opts.label="py_ed"}
# Load the NumPy library.
import numpy as np

# Make a range from 0 up to (not including) 6
some_numbers = np.arange(0, 6)

# Show the result
some_numbers
```

Notice that `np.arange(0, 6)` means "from 0 through 5" — the range *does not
include* the last value.  Another way of putting this is that `np.arange(0,
6)` means "from 0 up to, but not including, 6".

This takes a little getting used to, but we think you will find it starts to
come naturally after a while.
:::

::: {.r}
R also has the concept of a range of integers (positive or negative whole
numbers).  Here is an R range:

```{r, opts.label="r_ed"}
# Make a range from 0 through 5
some_numbers <- 0:5

# Show the result
some_numbers
```

Notice that `0:5` means exactly "from 0 through 5" — the range includes all
the integers starting at 0 and including 5.
:::

::::

#### Random numbers from 0 through 99

We can now use {{< var np_or_r >}} and the random number function from the last chapter to get 12 random numbers from 0 through 99.

```{python, opts.label="py_ed"}
# Ask NumPy for a random number generator.
rnd = np.random.default_rng()

# Get 12 random numbers from 0 through 99
# (from 0, up to, but not including 100).
a = rnd.integers(0, 100, size=12)

# Show the result
a
```

```{r, opts.label="r_ed"}
# Get 12 random numbers from 0 through 99
a <- sample(0:99, 12, replace=TRUE)

# Show the result
a
```

#### Counting the jurors

We use *comparison* and [`np.sum`]{.python}[`sum`]{.r} to count how
many numbers are are greater than 74, and therefore, in the range from 75
through 99:

```{python, opts.label="py_ed"}
# How many numbers are greater than 74?
b = np.sum(a > 74)
# Show the result
b
```

```{r, opts.label="r_ed"}
# How many numbers are greater than 74?
b <- sum(a > 74)
# Show the result
b
```

#### A single simulated trial

We assemble the pieces from the last few sections to make a {{< var cell >}} that simulates a single trial:

```{python, opts.label="py_ed"}
rnd = np.random.default_rng()

# Get 12 random numbers from 0 through 99
a = rnd.integers(0, 100, size=12)

# How many numbers are greater than 74?
b = np.sum(a > 74)

# Show the result
b
```

```{r, opts.label="r_ed"}
# Get 12 random numbers from 0 through 99
a <- sample(0:99, 12, replace=TRUE)
# How many are greater than 74?
b <- sum(a > 74)
# Show the result
b
```

## Three simulation steps

Now we come back to the details of how we:

1. Repeat the simulated trial many times;
2. record the results for each trial;
3. calculate the required proportion as an estimate of the probability
   we seek.

Repeating the trial many times is the job of the `for` loop, and we will come
to that soon.

In order to record the results, we will store each trial result in {{< var
an_array >}}.

:::{.callout-note}
### More on {{< var array >}}s

Since we will be working with {{< var array >}}s a lot, it is worth
knowing more about them.

A [NumPy array]{.python}[vector]{.r} is a *container* that stores many
elements of the same type.  You have already seen, in @sec-resampling-method,
how we can create {{< var an_array >}} from a sequence of
numbers using the [`np.array`]{.python}[`c()`]{.r} function.

```{python, opts.label="py_ed"}
# Make an array of numbers, store with the name "some_numbers".
some_numbers = np.array([0, 1, 2, 3, 4, 5, 6, 7, 8, 9])
# Show the value of "some_numbers"
some_numbers
```

```{r, opts.label="r_ed"}
# Make a vector of numbers, store with the name "some_numbers".
some_numbers <- c(0, 1, 2, 3, 4, 5, 6, 7, 8, 9)
# Show the value of "some_numbers"
some_numbers
```

Another way that we can create {{< var array >}}s is to use the
[`np.zeros`]{.python}[`numeric`]{.r} function to make a new array where all
the elements are 0.

```{python, opts.label="py_ed"}
# Make a new array containing 5 zeros.
# store with the name "z".
z = np.zeros(5)
# Show the value of "z"
z
```

```{r, opts.label="r_ed"}
# Make a new vector containing 5 zeros.
z <- numeric(5)
# Show the value of "z"
z
```

Notice the argument `5` to the [`np.zeros`]{.python}[`integer`]{.r} function.
This tells the function how many zeros we want in the {{< var array >}}
that the function will return.

The are various useful things we can do with this {{< var array >}} container.
One is to ask how many elements there are in the {{< var array >}} container.
We can use the [`len`]{.python}[`length`]{.r} function to calculate the number of elements in {{< var an_array >}}:

```{python, opts.label="py_ed"}
# Show the number of elements in "z"
len(z)
```

```{r, opts.label="r_ed"}
# Show the number of elements in "z"
length(z)
```

Another thing we can do is *set* the value for a particular element in the
{{< var array >}}.  To do this, we use square brackets following the
{{< var array >}} value, on the left hand side of the equals sign,
like this:

```{python, opts.label="py_ed"}
# Set the value of the *first* element in the array.
z[0] = 99
# Show the new contents of the array.
z
```

```{r, opts.label="r_ed"}
# Set the value of the first element in the vector.
z[1] = 99
# Show the new contents of the vector.
z
```

Read the first line of code as "the element at position
[0]{.python}[1]{.r} gets a value of 99".

::: python
Notice that the position number of the first element in the array is 0, and the
position number of the second element is 1.  Think of the position as an
*offset* from the beginning of the array.  The first element is at the
beginning of the array, and so it is at offset (position) 0.  This can be a
little difficult to get used to at first, but you fill find that thinking of
the positions of offsets in this way soon starts to come naturally, and later
you will also find that it helps you to avoid some common mistakes when using positions for getting and setting values.
:::

For practice, let us also set the value of the third element in the {{< var array >}}:

```{python, opts.label="py_ed"}
# Set the value of the *third* element in the array.
z[2] = 99
# Show the new contents of the array.
z
```

```{r, opts.label="r_ed"}
# Set the value of the third element in the vector.
z[3] = 99
# Show the new contents of the vector.
z
```

We can also *get* the value of the element at a given position, using the same
square-bracket notation:

```{python, opts.label="py_ed"}
# Get the value of the *first* element in the array.
# Store the value with name "v"
v = z[0]
# Show the value we got
v
```

```{r, opts.label="r_ed"}
# Get the value of the *first* element in the array.
# Store the value with name "v"
v = z[0]
# Show the value we got
v
```

Read the first code line here as "v gets the value at position 0 in the {{< var array >}}.

Using square brackets to get and set element values is called *indexing* into
the {{< var array >}}.
:::

### Repeating trials

As a preview, let us now imagine that we want to do 100 simulated trials of
Robert Swain's jury.  This will give us 100 counts.  We will want to store the
count for each trial.

In order to do this, we make {{< var an_array >}} to hold the 100 counts.  Call this {{< var array >}} `z`.

```{python, opts.label="py_ed"}
# An array to hold the 100 count values.
z = np.zeros(100)
```

```{r, opts.label="r_ed"}
# A vector to hold the 100 count values.
z <- numeric(100)
```

Now we could run a single trial to get a single simulated count.  Here we just
repeat the code {{< var cell >}} you saw above.  Notice that we can get a
different result each time we run this code, because the numbers in `a` are
*random* choices from `some_numbers`, and different random numbers will give
different counts.

```{python, opts.label="py_ed"}
rnd = np.random.default_rng()
# Get 12 random numbers from 0 through 99
a = rnd.integers(0, 100, 12)
# How many numbers are greater than 74?
b = np.sum(a > 74)
# Show the result
b
```

```{r, opts.label="r_ed"}
# Get 12 random numbers from 0 through 99
a <- sample(0:99, 12, replace=TRUE)
# How many are greater than 74?
b <- sum(a == 9)
# Show the result
b
```

Now we have the result of a single trial, we can store it as the first number
in the `z` {{< var array >}}:

```{python, opts.label="py_ed"}
# Store the single trial count as the first value in the "z" array.
z[0] = b
# Show all the values in the "z" array.
z
```

```{r, opts.label="r_ed"}
# Store the single trial count as the first value in the "z" vector.
z[1] <- b
# Show all the values in the "z" vector.
z
```

Of course we could just keep doing this: run the {{< var cell >}} corresponding
to a trial, above, to get a new count, and then store it at the next position
in the `z` {{< var array >}}. For example, we could simulate the first three
trials with:

```{python, opts.label="py_ed"}
# First trial
a = rnd.integers(0, 100, 12)
b = np.sum(a > 74)
# Store the result at the first position in z
# Remember, the first position is offset 0.
z[0] = b
# Second trial
a = rnd.integers(0, 100, 12)
b = np.sum(a > 74)
# Store the result at the second position in z
z[1] = b
# Third trial
a = rnd.integers(0, 100, 12)
b = np.sum(a > 74)
# Store the result at the third position in z
z[2] = b

# And so on ...
```


```{r, opts.label="r_ed"}
# First trial
a <- sample(0:99, 12, replace=TRUE)
b <- sum(a == 9)
# Store the result at the first position in z
z[1] <- b

# Second trial
a <- sample(0:99, 12, replace=TRUE)
b <- sum(a == 9)
# Store the result at the second position in z
z[2] <- b

# Third trial
a <- sample(0:99, 12, replace=TRUE)
b <- sum(a == 9)
# Store the result at the third position in z
z[3] <- b

# And so on ...
```

This would get terribly long and boring to type for 100 trials.  One task that
computer code is very good at, is repeating the same procedure many times.
For example, {{< var lang >}} can do this using a `for` loop. You have already
seen a preview of the `for` loop in @sec-resampling-method. Here we dive into
`for` loops in more depth.

:::{.callout-box}
### For-loops in {{< var lang >}}

<!-- Matthew wants to switch this to integers instead of strings; S needs to find out why -->

Imagine a factory robot in front of a conveyor belt along which several items have been placed.
It services each of these items, one by one, using the same procedure, until its work is done.

Here is an example:

```{python, opts.label="py_ed"}
for item in ['shoe', 'saddle', 'jacket']:
    print('I am stitching a', item)
```

```{r}
items <- c('shoe', 'saddle', 'jacket')
for (item in items) {
  message('I am stitching a ', item)
}
```

Here, the robot is working its way through three items: a shoe, a saddle, and a jacket.
Every time that a new object arrives in front of it, that object is assigned the label `item`.
The [indented block of code under]{.python}[block of code inside]{.r} the `for`
is executed for each one, and in this code block `label` can be accessed. You
will also notice us using a new function: [`print`]{.python}[`message`]{.r}.
This function receives several arguments, and then prints them out to screen.

The code above can be unpacked into the following equivalent program:

```{python, opts.label="py_ed"}
item = 'shoe'
print('I am stitching a', item)

item = 'saddle'
print('I am stitching a', item)

item = 'jacket'
print('I am stitching a', item)
```

```{r}
item = 'shoe'
message('I am stitching a ', item)

item = 'saddle'
message('I am stitching a ', item)

item = 'jacket'
message('I am stitching a ', item)
```

We used the name `item` here, but any other name would do.
Thus, the following is equivalent:

```{python, opts.label="py_ed"}
for package in ['shoe', 'saddle', 'jacket']:
    print('I am stitching a', package)
```

```{r}
items <- c('shoe', 'saddle', 'jacket')
for (package in items) {
  message('I am stitching a ', package)
}
```

It is also common to operate on numbers in a for loop:

```{python, opts.label="py_ed"}
for i in [0, 1, 2, 3, 4]:
    print('The square of', i, 'is', i * i)
```

```{r}
numbers = c(0, 1, 2, 3, 4)
for (i in numbers) {
  message('The square of ', i, ' is ', i * i)
}
```

When we want to count to a hundred, it becomes laborious to type out the
numbers. For that, {{< var lang> }} has a shorthand [called `range`—very similar to NumPy's `arange`]{.python}:

```{python, opts.label="py_ed"}
for i in range(5):
    print('The square of', i, ' is ', i * i)
```

```{r}
for (i in 0:4) {
  message('The square of ', i, ' is ', i * i)
}
```

:::

### Putting it all together

We found that we could construct the code for a single trial.  Here is the code for a single trial, again:

```{python, opts.label="py_ed"}
rnd = np.random.default_rng()
# Get 12 random numbers from 0 through 99
a = rnd.integers(0, 100, 12)
# How many numbers are greater than 74?
b = np.sum(a > 74)
# Show the result
b
```

```{r, opts.label="r_ed"}
# Get 12 random numbers from 0 through 99
a <- sample(0:99, 12, replace=TRUE)
# How many are greater than 74?
b <- sum(a == 9)
# Show the result
b
```

Then we found that we could use {{< var array >}}s to store the results
of these trials, and that we could use `for` loops to do many repeats of steps
like those for a single trial.

Now we can put these parts together to do 100 simulated trials:

```{python, opts.label="py_ed"}
# Procedure for 100 simulated trials.

# An array to store the counts for each trial.
z = np.zeros(100)

# Repeat the trial procedure 100 times.
for i in np.arange(0, 100):
    rnd = np.random.default_rng()
    # Get 12 random numbers from 0 through 99
    zero_through_99 = np.arange(0, 100)
    a = rnd.choice(zero_through_99, 12)
    # How many numbers are greater than 74?
    b = np.sum(a > 74)
    # Store the result at the next position in the "z" array.
    z[i] = b
    # Now go back and do the next trial until finished.
# Show the result of all 100 trials.
z
```

```{r, opts.label="r_ed"}
# Procedure for 100 simulated trials.

# An array to store the counts for each trial.
z <- numeric(100)

# Repeat the trial procedure 100 times.
for (i in 1:100) {
    # Get 12 random numbers from 0 through 99
    a <- sample(0:99, 12, replace=TRUE)
    # How many are greater than 74?
    b <- sum(a == 9)
    # Store the result at the next position in the "z" array.
    z[i] = b
    # Now go back and do the next trial until finished.
}
# Show the result of all 100 trials.
z
```

Finally, we need to count how many of the trials in `z` ended up with all-White juries.  These are the trials with a `z` value of 0.

To do this, we can ask {{< var an_array >}} which elements match a certain condition.  E.g.:

```{python}
x = np.array([0, 1, 2, 3])
mask = x < 2
print(mask)

# Select elements that match True values in mask
print(x[mask])
```

```{r}
x <- c(0, 1, 2, 3)
mask = x < 2
print(mask)
print(x[mask])
```

We now use that same technique to ask, of *each of the 100 counts*, whether the {{< var array >}} `z` is equal to 0, like this:

```{python, opts.label="py_ed"}
# Is the value of z equal to 0?
all_white = z == 0
# Show the result of the comparison.
all_white
```

```{r, opts.label="r_ed"}
# Is the value of z equal to 0?
all_white <- z == 0
# Show the result of the comparison.
all_white
```

Now we need to get the number of {{< var true >}} values in `all_white`, to find
how many simulated trials gave all-white juries.

```{python, opts.label="py_ed"}
# Count the number of True values in "all_white"
# This is the same as the number of values in "z" that are equal to 0.
n_all_white = np.sum(all_white)
# Show the result of the comparison.
n_all_white
```

```{r, opts.label="r_ed"}
# Count the number of True values in "all_white"
# This is the same as the number of values in "z" that are equal to 0.
n_all_white = sum(all_white)
# Show the result of the comparison.
n_all_white
```

`n_all_white` is the number of simulated trials for which all the jury members
were White. It only remains to get the proportion of trials for which this was
true, and to do this, we divide by the number of trials.

```{python, opts.label="py_ed"}
# Proportion of trials where all jury members were White.
p = n_all_white / 100
# Show the result
p
```

```{r, opts.label="r_ed"}
# Proportion of trials where all jury members were White.
p <- n_all_white / 100
# Show the result
p
```

From this simulation experiment, we see that there is roughly a 2% chance that
a jury selected randomly from the population, which was 26% Black, would have
no Black jurors.
