---
jupyter:
  jupytext:
    metadata_filter:
      notebook:
        additional: all
        excluded:
        - language_info
    text_representation:
      extension: .Rmd
      format_name: rmarkdown
      format_version: '1.0'
      jupytext_version: 0.8.6
  kernelspec:
    display_name: Python 3
    language: python
    name: python3
resampling_with:
    ed2_fname: null
---

```{r setup, include=FALSE}
source("_common.R")
```

# More resampling with code {#sec-resampling-two}

@sec-resampling-code introduced a problem in probability, that was also a
problem in statistics.  We asked how surprised we should be at the results of a
trial of a new cancer treatment regime.

Here we study another urgent problem in the real world - racial bias and the
death penalty.

## A question of life and death

This example comes from the excellent Berkeley introduction to data science
[@adhikari2021data8].

Robert Swain was a young black man who was sentenced to death in the early
60s.  Swain's trial was held in Talladega County, Alabama. At the time, 26% of
the eligible jurors in that county were black, but every member of Swain's
jury was white.  Swain and his legal team appealed to the Alabama Supreme
Court, and then to the [US Supreme
Court](https://en.wikipedia.org/wiki/Swain_v._Alabama), arguing that there was
racial bias in the jury selection.  They noted that there had been no black
jurors in Talladega county since 1950, even though they made up about a
quarter of the eligible pool of jurors. The US Supreme Court rejected this
argument, in a 6 to 3 opinion, writing that "The overall percentage disparity
has been small and reflects no studied attempt to include or exclude a
specified number of Negros.".

Swain's team presented a variety of evidence on bias in jury selection, but
here we will look at the obvious and apparently surprising fact that Swain's
jury was entirely white. The Supreme Court decided that the "disparity"
between selection of white and black jurors "has been small".  In this
particular case the disparity is between 0% of selected black jurors and 26%
of eligible black jurors.  How would the Supreme Court judges, and how would
we, make a rational decision about whether this particular disparity really
was "small"?

You might be wondering about the result of this decision for Robert Swain.  In
fact his death sentence was invalidated by a [later, unrelated
decision](https://en.wikipedia.org/wiki/Furman_v._Georgia) and he served
a long prison sentence instead. In 1986, the Supreme Court overturned the
precedent set by Swain's case, in [Batson v. Kentucky, 476 U.S.
79](https://supreme.justia.com/cases/federal/us/476/79).

## A small disparity and a hypothetical world

To answer the question that the Supreme Court asked, we return to the method we used in the last chapter.

Let us imagine a hypothetical world, in which each individual black or white
person had an equal chance of being selected for the jury.  Call this world Hypothetical County, Alabama.

Just as in 1960's Talladega County, 26% of eligible jurors in Hypothetical
County are black.  Hypothetical County jury selection has no bias against
black people, so we expect around 26% of the jury to be black.  26% of 12 is
3.12 (0.26 * 12 = 3.12), so we expect that, on average, just over 3 out of 12
jurors in a jury from Hypothetical County will be black.  But, if we select
each juror at random from the population, that means that, sometimes, by
chance, we will have fewer than 3 black jurors, and sometimes will have more
than 3 black jurors. And, by chance, sometimes we will have no black jurors.
But, if the jurors really are selected at random, how often would we expect
this to happen — that there are no black jurors?  We would like to estimate
the *probability* that we will get no black jurors.  If that probability is
small, then we have some evidence that the disparity in selection between
black and white jurors, was not "small".

::: {.question}
What is the probability of an *all white* jury being randomly selected out of a population having 26% black people?
:::

## Designing the experiment

Before we start, we need to figure out three things:

1. What do we mean by one trial?
2. What is the outcome of interest from the trial?
3. How do we simulate one trial?

We then take **three steps** to calculate the desired probability:

1. *Repeat* the simulated trial procedure N times.
2. *Count* M, the number of trials with an outcome that matches the outcome we
   are interested in.
3. Calculate the *proportion*, M/N.
   This is an estimate of the probability in question.

For this problem, our task is made a little easier by the fact that our *trial* (in the resampling sense) is a simulated *trial* (in the legal sense). One trial requires 12 simulated jurors, each labeled by race (white or black).

The outcome we are interested in is the number of black jurors.

Now comes the harder part.  How do we simulate one trial?

### One trial

One trial requires 12 jurors, and we are interested only in the race of each juror.
In Hypothetical County, where selection by race is entirely random, each juror
has a 26% chance of being black.

We need a way of simulating a 26% chance.

One way of doing this is by getting a random number from 0 through 99
(inclusive).  There are 100 numbers in the range 0 through 99 (inclusive).

We will arbitrarily say that the juror is white if the random number is in the
range from 0 through 73.  74 of the 100 numbers are in this range, so
the juror has a 74/100 = 74% chance of getting the label "white".  We will say
the juror is black if the random number is in the range 74 though 99.  There
are 26 such numbers, so the juror has a 26% chance of getting the label
"black".

Next we need a way of getting a random number in the range 0 through 99. This
is an easy job for the computer, but if we had to do this with a physical
device, we could get a single number by throwing *two* 10-sided dice, say a
blue die and a green die.  The face of the blue die will be the 10s digit, and
the green face will be the ones digit.  So, if the blue die comes up with 8 and
the green die has 4, then the random number is 84.

We could then simulate 12 jurors by repeating this process 12 times, each time
writing down "white" if the number is from 0 through 74, and "black"
otherwise.  The trial outcome is the number of times we wrote "black" for these
12 simulated jurors.

::: {.notebook name="life_and_death" title="Simulating juries"}

### Using code to simulate a trial

We use the same logic to simulate a trial with the computer. A little code
makes the job easier, because we can ask {{< var lang >}} to give us 12 random
numbers from 0 through 99, and to count how many of these numbers are in the
range from 75 through 99.  Numbers in the range from 75 through 99 correspond
to black jurors.

### Random numbers from 0 through 99 {#sec-random-zero-through-ninety-nine}

We can now use {{< var np_or_r >}} and
[the random number functions]{.python}[`sample`]{.r}
from the last chapter to get 12 random numbers from 0 through 99.

```{python}
# Import the Numpy library, rename as "np"
import numpy as np

# Ask Numpy for a random number generator.
rnd = np.random.default_rng()

# All the integers from 0 up to, but not including 100.
zero_thru_99 = np.arange(100)

# Get 12 random numbers from 0 through 99
a = rnd.choice(zero_thru_99, size=12)

# Show the result
a
```

```{r}
# Get 12 random numbers from 0 through 99
a <- sample(0:99, size=12, replace=TRUE)

# Show the result
a
```

#### Counting the jurors

We use *comparison* and [`np.sum`]{.python}[`sum`]{.r} to count how
many numbers are greater than 74, and therefore, in the range from 75 through
99:

```{python}
# How many numbers are greater than 74?
b = np.sum(a > 74)
# Show the result
b
```

```{r}
# How many numbers are greater than 74?
b <- sum(a > 74)
# Show the result
b
```

#### A single simulated trial

We assemble the pieces from the last few sections to make a {{< var cell >}}
that simulates a single trial:

```{python}
rnd = np.random.default_rng()
zero_thru_99 = np.arange(100)

# Get 12 random numbers from 0 through 99
a = rnd.choice(zero_thru_99, size=12)

# How many numbers are greater than 74?
b = np.sum(a > 74)

# Show the result
b
```

```{r}
# Get 12 random numbers from 0 through 99
a <- sample(0:99, size=12, replace=TRUE)
# How many are greater than 74?
b <- sum(a > 74)
# Show the result
b
```

## Three simulation steps

Now we come back to the details of how we:

1. Repeat the simulated trial many times;
2. record the results for each trial;
3. calculate the required proportion as an estimate of the probability we seek.

Repeating the trial many times is the job of the `for` loop, and we will come
to that soon.

In order to record the results, we will store each trial result in {{< var
an_array >}}.

### More on {{< var array >}}s

Since we will be working with {{< var array >}}s a lot, it is worth knowing
more about them.

A [Numpy array]{.python}[vector]{.r} is a *container* that stores many
elements of the same type.  You have already seen, in @sec-resampling-method,
how we can create {{< var an_array >}} from a sequence of
numbers using the [`np.array`]{.python}[`c()`]{.r} function.

```{python}
# Make an array of numbers, store with the name "some_numbers".
some_numbers = np.array([0, 1, 2, 3, 4, 5, 6, 7, 8, 9])
# Show the value of "some_numbers"
some_numbers
```

```{r}
# Make a vector of numbers, store with the name "some_numbers".
some_numbers <- c(0, 1, 2, 3, 4, 5, 6, 7, 8, 9)
# Show the value of "some_numbers"
some_numbers
```

Another way that we can create {{< var array >}}s is to use the
[`np.zeros`]{.python}[`numeric`]{.r} function to make a new array where all the
elements are 0.

```{python}
# Make a new array containing 5 zeros.
# store with the name "z".
z = np.zeros(5)
# Show the value of "z"
z
```

```{r}
# Make a new vector containing 5 zeros.
z <- numeric(5)
# Show the value of "z"
z
```

Notice the argument `5` to the [`np.zeros`]{.python}[`numeric`]{.r} function.
This tells the function how many zeros we want in the {{< var array >}} that
the function will return.

## [Array]{.python}[Vector]{.r} length {#sec-array-length}

The are various useful things we can do with this {{< var array >}} container.
One is to ask how many elements there are in the {{< var array >}} container.
We can use the [`len`]{.python}[`length`]{.r} function to calculate the number
of elements in {{< var an_array >}}:

```{python}
# Show the number of elements in "z"
len(z)
```

```{r}
# Show the number of elements in "z"
length(z)
```

## Indexing into {{< var array >}}s with integers {#sec-array-indexing}

Another thing we can do with {{< var array >}}s is *set* the value for
a particular element.  To do this, we use square brackets following the {{< var
array >}} value, on the left hand side of the equals sign, like this:

```{python}
# Set the value of the *first* element in the array.
z[0] = 99
# Show the new contents of the array.
z
```

```{r}
# Set the value of the first element in the vector.
z[1] = 99
# Show the new contents of the vector.
z
```

Read the first line of code as "the element at position
[0]{.python}[1]{.r} gets a value of 99".

::: python
Notice that the position number of the first element in the array is 0, and the
position number of the second element is 1.  Think of the position as an
*offset* from the beginning of the array.  The first element is at the
beginning of the array, and so it is at offset (position) 0.  This can be a
little difficult to get used to at first, but you will find that thinking of
the positions of offsets in this way soon starts to come naturally, and later
you will also find that it helps you to avoid some common mistakes when using
positions for getting and setting values.
:::

For practice, let us also set the value of the third element in the {{< var
array >}}:

```{python}
# Set the value of the *third* element in the array.
z[2] = 99
# Show the new contents of the array.
z
```

```{r}
# Set the value of the third element in the vector.
z[3] <- 99
# Show the new contents of the vector.
z
```

Read the first code line above as as "set the value at position
[2]{.python}[3]{.r}
in the {{< var array >}} to have the value 99".

We can also *get* the value of the element at a given position, using the same
square-bracket notation:

```{python}
# Get the value of the *first* element in the array.
# Store the value with name "v"
v = z[0]
# Show the value we got
v
```

```{r}
# Get the value of the *first* element in the array.
# Store the value with name "v"
v = z[1]
# Show the value we got
v
```

Read the first code line here as "v gets the value at position
[0]{.python}[1]{.r}
in the {{< var array >}}".

Using square brackets to get and set element values is called *indexing* into
the {{< var array >}}.

### Repeating trials

As a preview, let us now imagine that we want to do 50 simulated trials of
Robert Swain's jury in Hypothetical County.  We will want to store the count
for each trial, to give 50 counts.

In order to do this, we make {{< var an_array >}} to hold the 50 counts.
Call this {{< var array >}} `z`.

```{python}
# An array to hold the 50 count values.
z = np.zeros(50)
```

```{r}
# A vector to hold the 50 count values.
z <- numeric(50)
```

We could run a single trial to get a single simulated count.  Here we just
repeat the code {{< var cell >}} you saw above.  Notice that we can get a
different result each time we run this code, because the numbers in `a` are
*random* choices from the range 0 through 99, and different random numbers will
give different counts.

```{python}
rnd = np.random.default_rng()
zero_thru_99 = np.arange(100)
# Get 12 random numbers from 0 through 99
a = rnd.choice(zero_thru_99, size=12)
# How many numbers are greater than 74?
b = np.sum(a > 74)
# Show the result
b
```

```{r}
# Get 12 random numbers from 0 through 99
a <- sample(0:99, size=12, replace=TRUE)
# How many are greater than 74?
b <- sum(a == 9)
# Show the result
b
```

Now we have the result of a single trial, we can store it as the first number
in the `z` {{< var array >}}:

```{python}
# Store the single trial count as the first value in the "z" array.
z[0] = b
# Show all the values in the "z" array.
z
```

```{r}
# Store the single trial count as the first value in the "z" vector.
z[1] <- b
# Show all the values in the "z" vector.
z
```

Of course we could just keep doing this: run the {{< var cell >}} corresponding
to a trial, above, to get a new count, and then store it at the next position
in the `z` {{< var array >}}. For example, we could store the counts for the
first three trials with:

```{python}
# First trial
a = rnd.choice(zero_thru_99, size=12)
b = np.sum(a > 74)
# Store the result at the first position in z
# Remember, the first position is offset 0.
z[0] = b
# Second trial
a = rnd.choice(zero_thru_99, size=12)
b = np.sum(a > 74)
# Store the result at the second position in z
z[1] = b
# Third trial
a = rnd.choice(zero_thru_99, size=12)
b = np.sum(a > 74)
# Store the result at the third position in z
z[2] = b

# And so on ...
```

```{r}
# First trial
a <- sample(0:99, size=12, replace=TRUE)
b <- sum(a == 9)
# Store the result at the first position in z
z[1] <- b

# Second trial
a <- sample(0:99, size=12, replace=TRUE)
b <- sum(a == 9)
# Store the result at the second position in z
z[2] <- b

# Third trial
a <- sample(0:99, size=12, replace=TRUE)
b <- sum(a == 9)
# Store the result at the third position in z
z[3] <- b

# And so on ...
```

This would get terribly long and boring to type for 50 trials.  Luckily
computer code is very good at repeating the same procedure many times. For
example, {{< var lang >}} can do this using a `for` loop. You have already seen
a preview of the `for` loop in @sec-resampling-method and
@sec-resampling-code. Here we dive into `for` loops in more depth.

### For-loops in {{< var lang >}} {#sec-for-loops}

A for-loop is a way of asking {{< var lang >}} to:

* Take a sequence of things, one by one, and
* Do the same task on each one.

We often use this idea when we are trying to explain a repeating procedure.
For example, imagine we wanted to explain what the supermarket checkout person
does for the items in your shopping basket.  You might say that they do this:

> For each item of shopping in your basket, they take the item off the conveyor
> belt, scan it, and put it on the other side of the till.

You could also break this description up into bullet points with indentation,
to say the same thing:

* For each item from your shopping basket, they:
  * Take the item off the conveyor belt.
  * Scan the item.
  * Put it on the other side of the till.

Notice the logic; the checkout person is repeating the same procedure for each
of a series of items.

This is the logic of the `for` loop in {{< var lang >}}.  The procedure that
{{< var lang >}} repeats is called the *body of the for loop*.  In the example
of the checkout person above, the repeating procedure is:

  * Take the item off the conveyor belt.
  * Scan the item.
  * Put it on the other side of the till.

Now imagine we wanted to use {{< var lang >}} to print out the year of birth
for each of the authors for the third edition of this book:

Author                Year of birth
------                ----
Julian Lincoln Simon  1932
Matthew Brett         1964
Stéfan van der Walt   1980

We want to see this output:

```
Author birth year is 1932
Author birth year is 1964
Author birth year is 1980
```

Of course, we could just ask {{< var lang >}} to print out these exact lines,
like this:

```{python}
print('Author birth year is 1932')
print('Author birth year is 1964')
print('Author birth year is 1980')
```

```{r}
message('Author birth year is 1932')
message('Author birth year is 1964')
message('Author birth year is 1980')
```

We might instead notice that we are repeating the same procedure for each of
the three birth years, and decide to do the same thing using a `for` loop:

```{python}
author_birth_years = np.array([1932, 1964, 1980])

# For each birth year
for birth_year in author_birth_years:
    # Repeat this procedure ...
    print('Author birth year is', birth_year)
```

```{r}
author_birth_years <- c(1932, 1964, 1980)

# For each birth year
for (birth_year in author_birth_years) {
    # Repeat this procedure ...
    message('Author birth year is ', birth_year)
}
```

The `for` loop starts with a line where we tell it what items we want to repeat
the procedure for:

::: python

```
for birth_year in author_birth_years:
```

This *initial line* of the `for` loop ends with a colon.

The next thing in the `for` loop is the procedure Python should follow for each
item.  Python knows that the following lines are the procedure it should
repeat, because the lines are *indented*. The *indented* lines are the *body of
the for loop*.
:::

::: r

```
for (birth_year in author_birth_years) {
```

This *initial line* of the `for` loop ends with an *opening curly brace* `{`.
The opening curly brace tells R that what follows, up until the matching
closing curly brace `}`, is the procedure R should follow for each item.   The
lines between the opening `{` and closing `}` curly braces* are the *body of
the for loop*.
:::

The initial line of the `for` loop above tells {{< var lang >}} that it should
take *each item* in `author_birth_years`, one by one — first 1932, then 1964,
then 1980.  For each of these numbers it will:

* Put the number into the variable `birth_year`, then
* Run the [indented]{.python} code[ between the curly braces]{.r}.

Just as the person at the supermarket checkout takes each item in turn, for
each iteration (repeat) of the `for` loop, `birth_year` gets a new value from
the sequence in `author_birth_years`. `birth_year` is called the *loop
variable*, because it is the variable that gets a new value each time we begin
a new iteration of the `for` loop procedure.  As for any variable in {{< var
lang >}}, we can call our loop variable anything we like. We used `birth_year`
here, but we could have used `y` or `year` or some other name.

::: r

Notice that R insists we put parentheses (round brackets) around: the loop
variable; `in`; and the sequence that will fill the loop variable — like this:

```
for (birth_year in author_birth_years) {
```

Do not forget these round brackets — R insists on them.
:::

Now you know what the `for` loop is doing, you can see that the `for` loop
above is equivalent to the following code:

```{python}
birth_year = 1932  # Set the loop variable to contain the first value.
print('Author birth year is', birth_year)  # Use it.
birth_year = 1964  # Set the loop variable to contain the next value.
print('Author birth year is', birth_year)  # Use the second value.
birth_year = 1980
print('Author birth year is', birth_year)
```

```{r}
birth_year <- 1932  # Set the loop variable to contain the first value.
message('Author birth year is ', birth_year)  # Use the first value.
birth_year <- 1964  # Set the loop variable to contain the next value.
message('Author birth year is ', birth_year)  # Use the second value.
birth_year <- 1980
message('Author birth year is ', birth_year)
```

Writing the steps in the `for` loop out like this is called *unrolling* the
loop.  It can be a useful exercise to do this when you come across a `for`
loop, in order to work through the logic of the loop. For example, you may want
to write out the unrolled equivalent of the first couple of iterations, to see
what the loop variable will be, and what will happen in the body of the loop.

We often use `for` loops with ranges (see @sec-ranges).  Here we use a loop to
print out the numbers [0 through 3]{.python}[1 through 4]{.r}:

```{python}
for n in np.arange(4):
    print('The loop variable n is', n)
```

```{r}
for (n in 1:4) {
    message('The loop variable n is ', n)
}
```

Notice that the range ended at [(the number before)]{.python} 4, and that means
we repeat the loop body 4 times.  We can also use the loop variable value from
the range as an *index*, to get or set the first, second, etc values from {{<
var an_array >}}.

For example, maybe we would like to show the author position *and* the author
year of birth.

Remember our author birth years:

```{python}
author_birth_years
```

```{r}
author_birth_years
```

We can get (for example) the second author birth year with:

```{python}
author_birth_years[1]
```

::: python
Remember, for Python, the first element is position 0, so the second element is
position 1.
:::

```{r}
author_birth_years[2]
```

Using the combination of looping over a range, and {{< var array >}} indexing,
we can print out the author position *and* the author birth year:

```{python}
for n in np.arange(3):
    year = author_birth_years[n]
    print('Birth year of author position', n, 'is', year)
```

::: python
Again, remember Python considers 0 as the first position.
:::

```{r}
for (n in 1:3) {
    year <- author_birth_years[n]
    message('Birth year of author position ', n, ' is ', year)
}
```

Just for practice, let us unroll the three iterations through this `for`
loop, to remind ourselves what the code is doing:

```{python}
# Unrolling the for loop.
n = 0
year = author_birth_years[n]  # Will be 1932
print('Birth year of author position', n, 'is', year)
n = 1
year = author_birth_years[n]  # Will be 1964
print('Birth year of author position', n, 'is', year)
n = 2
year = author_birth_years[n]  # Will be 1980
print('Birth year of author position', n, 'is', year)
```

```{r}
# Unrolling the for loop.
n <- 1
year <- author_birth_years[n]  # Will be 1932
message('Birth year of author position ', n, ' is ', year)
n <- 2
year <- author_birth_years[n]  # Will be 1964
message('Birth year of author position ', n, ' is ', year)
n <- 3
year <- author_birth_years[n]  # Will be 1980
message('Birth year of author position ', n, ' is ', year)
```

::: python

### `range` in Python `for` loops {#sec-for-range}

So far we have used `np.arange` to give us the sequence of integers that we
feed into the `for` loop.  But — as you saw in @sec-python-range — we can also
get a range of numbers from Python's `range` function.  `range` is a common
and useful alternative way to provide a range of numbers to a `for` loop.

You have just seen how we would use `np.arange` to send the numbers 0, 1, 2,
and 3 to a `for` loop, in the example above, repeated here:

```{python}
for n in np.arange(3):
    year = author_birth_years[n]
    print('Birth year of author position', n, 'is', year)
```

We could also use `range` instead of `np.arange` to do the same task:

```{python}
for n in range(3):
    year = author_birth_years[n]
    print('Birth year of author position', n, 'is', year)
```

In fact, you will see this pattern throughout the book, where we use `for`
statements like `for value in range(10000):` to ask Python to put each number
in the range 0 up to (not including) 10000 into the variable `value`, and then
do something in the body of the loop.  Just to be clear, we could always, and
almost as easily, write `for value in np.arange(10000):` to do the same task.
However, we generally prefer `range` in our Python `for` loops, because it is
just a little less typing (without the `np.a` of `np.arange`), and because it
is a more common pattern in standard Python code.[^range-efficiency]

[^range-efficiency]: Actually, there is a reason why many Python programmers
   prefer `range` to `np.arange` in the headers for their `for` loops. `range`
   is a very efficient container, in that it doesn't need to take up all the
   memory required to create the full array, it just needs to keep track of the
   number to give you next. For example, consider
   `for i in np.arange(10000000):` — in this case Python has to make an array
   with 10,000,000 elements, and then, from that array, it passes each value
   one by one to the `for` loop. On the other hand,
   `for i in range(10000000):` will do the job just as well, passing the same
   sequence of 0 through 9,999,999 to `i`, one by one, but `range(10000000)`
   never has to make the whole 10,000,000 element array — it just needs to
   keep track of which number to give up next.  Therefore `range` is very
   quick, and very efficient in memory.  This doesn't have any great practical
   impact for the arrays we are using here, typically of 10,0000 elements or
   so, but it can be important for larger arrays.

:::

### Putting it all together

Here is the code we worked out above, to implement a single trial:

```{python}
rnd = np.random.default_rng()
zero_thru_99 = np.arange(100)
# Get 12 random numbers from 0 through 99
a = rnd.choice(zero_thru_99, size=12)
# How many numbers are greater than 74?
b = np.sum(a > 74)
# Show the result
b
```

```{r}
# Get 12 random numbers from 0 through 99
a <- sample(0:99, size=12, replace=TRUE)
# How many are greater than 74?
b <- sum(a == 9)
# Show the result
b
```

We found that we could use {{< var array >}}s to store the results of these
trials, and that we could use `for` loops to repeat the same procedure many
times.

Now we can put these parts together to do 50 simulated trials:

```{python}
# Procedure for 50 simulated trials.

# The Numpy random number generator.
rnd = np.random.default_rng()

# All the numbers from 0 through 99.
zero_through_99 = np.arange(100)

# An array to store the counts for each trial.
z = np.zeros(50)

# Repeat the trial procedure 50 times.
for i in np.arange(50):
    # Get 12 random numbers from 0 through 99
    a = rnd.choice(zero_through_99, size=12)
    # How many numbers are greater than 74?
    b = np.sum(a > 74)
    # Store the result at the next position in the "z" array.
    z[i] = b
    # Now go back and do the next trial until finished.
# Show the result of all 50 trials.
z
```

```{r}
# Procedure for 50 simulated trials.

# A vector to store the counts for each trial.
z <- numeric(50)

# Repeat the trial procedure 50 times.
for (i in 1:50) {
    # Get 12 random numbers from 0 through 99
    a <- sample(0:99, size=12, replace=TRUE)
    # How many are greater than 74?
    b <- sum(a > 74)
    # Store the result at the next position in the "z" vector.
    z[i] = b
    # Now go back and do the next trial until finished.
}
# Show the result of all 50 trials.
z
```

Finally, we need to count how many of the trials in `z` ended up with all-white
juries.  These are the trials with a `z` (count) value of 0.

To do this, we can ask {{< var an_array >}} which elements match a certain
condition.  E.g.:

```{python}
x = np.array([2, 1, 3, 0])
y = x < 2
# Show the result
y
```

```{r}
x <- c(2, 1, 3, 0)
y = x < 2
# Show the result
y
```

We now use that same technique to ask, of *each of the 50 counts*, whether the
{{< var array >}} `z` is equal to 0, like this:

```{python}
# Is the value of z equal to 0?
all_white = z == 0
# Show the result of the comparison.
all_white
```

```{r}
# Is the value of z equal to 0?
all_white <- z == 0
# Show the result of the comparison.
all_white
```

We need to get the number of {{< var true >}} values in `all_white`, to find
how many simulated trials gave all-white juries.

```{python}
# Count the number of True values in "all_white"
# This is the same as the number of values in "z" that are equal to 0.
n_all_white = np.sum(all_white)
# Show the result of the comparison.
n_all_white
```

```{r}
# Count the number of True values in "all_white"
# This is the same as the number of values in "z" that are equal to 0.
n_all_white = sum(all_white)
# Show the result of the comparison.
n_all_white
```

`n_all_white` is the number of simulated trials for which all the jury members
were white. It only remains to get the proportion of trials for which this was
true, and to do this, we divide by the number of trials.

```{python}
# Proportion of trials where all jury members were white.
p = n_all_white / 50
# Show the result
p
```

```{r}
# Proportion of trials where all jury members were white.
p <- n_all_white / 50
# Show the result
p
```

From this initial simulation, it seems there is around a
`r round(get_var('p') * 100, 1)`% chance that a jury selected randomly from
the population, which was 26% black, would have no black jurors.

## Many many trials

Our experiment above is only 50 simulated trials.  The higher the number of
trials, the more confident we can be of our estimate for `p` — the proportion
of trials where we get an all-white jury.

It is no extra trouble for us to tell the computer to do a very large number
of trials.  For example, we might want to run 10,000 trials instead of 50.
All we have to do is to run the loop 10,000 times instead of 50 times. The
computer has to do more work, but it is more than up to the job.

Here is exactly the same code we ran above, but collected into one {{< var
cell >}}, and using 10,000 trials instead of 50.  We have left out the
comments, to make the code more compact.

```{python}
# Full simulation procedure, with 10,000 trials.
rnd = np.random.default_rng()
zero_through_99 = np.arange(100)
# 10,000 trials.
z = np.zeros(10000)
for i in np.arange(10000):
    a = rnd.choice(zero_through_99, size=12)
    b = np.sum(a > 74)
    z[i] = b
all_white = z == 0
n_all_white = sum(all_white)
p = n_all_white / 10000
p
```

```{r}
# Full simulation procedure, with 10,000 trials.
z <- numeric(10000)
for (i in 1:10000) {
    a <- sample(0:99, size=12, replace=TRUE)
    b <- sum(a > 74)
    z[i] = b
}
all_white <- z == 0
n_all_white <- sum(all_white)
p <- n_all_white / 10000
p
```

We now have a new, more accurate estimate of the proportion of Hypothetical
County juries that are all white. The proportion is
`r round(get_var('p'), 3)`, and so
`r round(get_var('p') * 100, 1)`%.

This proportion means that, for any one jury from Hypothetical County, there
is a less than one in 20 chance that the jury would be all white.

As we will see in more detail later, we might consider using the results from
this experiment in Hypothetical County, to reflect on the result we saw in the
real Talladega County.  We might conclude, for example, that there was likely
some systematic difference between Hypothetical County and Talledega County.
Maybe the difference was that there was, in fact, some bias in the jury
selection in Talledega county, and that the Supreme Court was wrong to reject
this.  You will hear more of this line of reasoning later in the book.

<!---
End of notebook.
-->
:::

## Conclusion

In this chapter we studied a real life-and-death question, on racial bias and
the death penalty.   We continued our exploration of the ways we can use
probability, and resampling, to draw conclusions about real events.  Along the
way, we went into more detail on {{< var array >}}s in {{< var lang >}}, and
`for` loops; two basic tools in resampling.

In the next chapter, we will work through some more problems in probability,
to show how we can use resampling, to answer questions about chance.   We will
add some more tools for writing code in {{< var lang >}}, to make your
programs easier to write, read, and understand.
