---
resampling_with:
    ed2_fname: null
jupyter:
  jupytext:
    notebook_metadata_filter: all,-language_info
    split_at_heading: true
    text_representation:
      extension: .Rmd
      format_name: rmarkdown
      format_version: '1.2'
      jupytext_version: 1.14.6
  kernelspec:
    display_name: Python 3 (ipykernel)
    language: python
    name: python3
---

```{r setup, include=FALSE}
source("_common.R")
```

# Tools for samples and sampling {#sec-sampling-tools}

## Introduction

Now you have some experience with {{< var lang >}}, probabilities and
resampling, it is time to introduce some useful tools for our experiments and
programs.

::: {.notebook name="sampling_tools" title="Sampling tools"}

## Samples and labels


Thus far we have used numbers such as 1 and 0 and 10 to represent the elements
we are sampling from.  For example, in @sec-resampling-two, we were simulating
the chance of a particular juror being black, given that 26% of the eligible
jurors in the county were black.   We used *integers* for that task, where we
started with all the integers from 0 through 99, and asked {{< var np_or_r >}}
to select values at random from those integers.  When {{< var np_or_r >}}
selected an integer from 0 through 25, we chose to label the resulting
simulated juror as black — there are 26 integers in the range 0 through 25, so
there is a 26% chance that any one integer will be in that range.  If the
integer was from 26 through 99, the simulated juror was white (there are 74
integers in the range 26 through 99).

Here is the process of simulating a single juror, adapted from
@sec-random-zero-through-ninety-nine:

```{python opts.label="py_ed"}
import numpy as np
# Ask NumPy for a random number generator.
rnd = np.random.default_rng()

# All the integers from 0 up to, but not including 100.
zero_thru_99 = np.arange(100)

# Get one random numbers from 0 through 99
a = rnd.choice(zero_thru_99)

# Show the result
a
```

```{r opts.label="r_ed"}
# Get 1 random number from 0 through 99
# replace=TRUE is redundant here (why?), but we leave it for consistency.
a <- sample(0:99, 1, replace=TRUE)

# Show the result
a
```

After that, we have to unpack our labeling of 0 through 25 as being "black"
and 26 through 99 as being "white".  We might do that like this:

```{python opts.label="py_ed"}
this_juror_is_black = a < 26
this_juror_is_black
```

```{r opts.label="r_ed"}
this_juror_is_black <- a < 26
this_juror_is_black
```

This all works as we want it to, but it's just a little bit difficult to
remember the coding (less than 26 means "black", greater than 25 means
"white").  We had to use that coding because we committed ourselves to using
random numbers to simulate the outcomes.

However, {{< var lang >}} can also store bits of text, called
*strings*.  Values that are bits of text can be very useful because the
text values can be memorable labels for the entities we are sampling
from, in our simulations.

::: python

Before we get to strings, let us consider the different types of value we have
seen so far.

## Types of values in Python

You have already come across the idea that Python values can be integers (positive or negative whole numbers), like this:

```{python opts.label="py_ed"}
v = 10
v
```

Here the variable `v` holds the value.  We can see what type of value `v` holds
by using the `type` function:

```{python opts.label="py_ed"}
type(v)
```

As you may have noticed, Python can also have *floating point* values.  These are values with a decimal point — so numbers that do not have to be integers, but can be any value between the integers.  These floating points values are of type `float`:

```{python opts.label="py_ed"}
f = 10.1
type(f)
```

### Numpy arrays

You have also seen that Numpy contains another type, the *array*.  An
array is a value that contains a sequence of values.  For example, here
is an array of integers:

```{python opts.label="py_ed"}
arr = np.array([0, 10, 99, 4])
arr
```

Notice that this value `arr` is of type `np.ndarray`:

```{python opts.label="py_ed"}
type(arr)
```

The array has its own internal record of what type of values it holds.  This is
called the array `dtype`:

```{python opts.label="py_ed"}
arr.dtype
```

The array `dtype` records the type of value stored in the array.  All values in the array must be of this type, and all values in the array are therefore of the same type.

The array above contains integers, but we can also make arrays containing floating point values:

```{python opts.label="py_ed"}
float_arr = np.array([0.1, 10.1, 99.0, 4.3])
float_arr
```

```{python opts.label="py_ed"}
float_arr.dtype
```

### Lists {#sec-lists}

We have elided past another Python type, the *list*.  In fact we have already used lists in making arrays.  For example, here we make an array with four values:

```{python opts.label="py_ed"}
np.array([0, 10, 99, 4])
```

We could also write the statement above in two steps:

```{python opts.label="py_ed"}
my_list = [0, 10, 99, 4]
np.array(my_list)
```

In the first statement — `my_list = [0, 10, 99, 4]` — we construct a *list* — a
container for the four values.  Let's look at the `my_list` value:

```{python, opts.label="py_ed"}
my_list
```

Notice that we do not see `array` in the display — this is not an array but a list:

```{python, opts.label="py_ed"}
type(my_list)
```

A list is a basic Python type.  We can construct it by using the square
brackets notation that you see above; we start with `[`, then we put the values
we want to go in the list, separated by commas, followed by `]`.  Here is another list:

```{python, opts.label="py_ed"}
# Creating another list.
list_2 = [5, 10, 20]
```

As you saw, we have been building arrays by building lists, and then passing
the list to the `np.array` function, to create an array.

```{python, opts.label="py_ed"}
list_again = [100, 10, 0]
np.array(list_again)
```

Of course, we can do this one line, as we have been doing up till now, by
constructing the list inside the parentheses of the function.  So, the
following cell has just the same output as the cell above:

```{python, opts.label="py_ed"}
# Constructing the list inside the function brackets.
np.array([100, 10, 0])
```

Lists are like arrays in that they are values that contain values, but they are
unlike arrays in various ways — that we will not go into now.  We often use
lists to construct sequences into lists to turn them into arrays.  For our
purposes, and particularly for our calculations, arrays are much more useful
and efficient than lists.

<!---
End of Python section
-->
:::

## String values {#sec-intro-to-strings}

So far, all the values you have seen in {{< var lang >}} {{< var array >}}s
have been numbers.  Now we get on to values that are bits of text.  These are
called *strings*.

Here is a single {{< var lang >}} string value:

```{python opts.label="py_ed"}
s = "Resampling"
s
```

```{r opts.label="r_ed"}
s <- "Resampling"
s
```

::: r
We can see what type of value `v` holds by using the `class` function.

For example, for a number value, you will usually find the `class` is
`numeric`:

```{r opts.label="r_ed"}
v <- 10
class(v)
```
:::

What is the
[`class`]{.r}[`type`]{.python}
of the new bit-of-text value `s`?

```{python opts.label="py_ed"}
type(s)
```

```{r opts.label="r_ed"}
class(s)
```

The {{< var lang >}} [`str`]{.python}[`character`]{.r} value is a bit of text,
and therefore consists of a sequence of characters.

As {{< var array >}}s are containers for other things, such as numbers, strings
are containers for characters.

::: python
As we can find the number of elements in an {{< var array >}}
(@sec-array-length), we can find the number of characters in a string with the
`len` function:

```{python opts.label="py_ed"}
# Number of characters in s
len(s)
```
:::

::: r
To get the length of a string, use the `nchar` function (Number of Characters):


```{r opts.label="r_ed"}
# Number of characters in s
nchar(s)
```
:::

::: python
As we can *index* into {{< var array >}} values to get individual elements
(@sec-array-indexing), we can index into string values to get individual
characters:

```{python opts.label="py_ed"}
# Get the second character of the string
# Remember, Python's index positions start at 0.
second_char = s[1]
second_char
```
:::

::: r
R has a `substring` function that allows you to select individual characters or
sequences of characters from a string. The arguments to `substring` are: first
— the string; second — the index of the first character you want to select; and
third — the index of the last character you want to select.  For example to
select the second character in the string you would specify 2 as the starting
index, and 2 as the ending index, like this:

```{r opts.label="r_ed"}
# Get the second character of the string
second_char <- substring(s, 2, 2)
second_char
```
:::

::: python

## Strings in {{< var array >}}s

As we can store numbers as elements in {{< var array >}}s, we can also store
strings as {{< var array >}} elements.

```{python opts.label="py_ed"}
# Just for clarity, make the list first.
# Lists can also contain strings.
list_of_strings = ['Julian', 'Lincoln', 'Simon']
# Then pass the list to np.array to make the array.
arr_of_strings = np.array(list_of_strings)
arr_of_strings
```

```{python opts.label="py_ed"}
# We can also create the list and the array in one line,
# as we have been doing up til now.
arr_of_strings = np.array(['Julian', 'Lincoln', 'Simon'])
arr_of_strings
```

```{r opts.label="r_ed"}
vector_of_strings = c('Julian', 'Lincoln', 'Simon')
vector_of_strings
```

::: python
Notice the array `dtype`:

```{python opts.label="py_ed"}
arr_of_strings.dtype
```

The `U` in the `dtype` tells you that the elements in the array are
[Unicode](https://en.wikipedia.org/wiki/Unicode) strings (Unicode is a computer
representation of text characters).  The number after the `U` gives the maximum
number of characters for any string in the array, here set to the length of the
longest string when we created the array.

::: {.callout-warning}
## Take care with Numpy string arrays

It is easy to run into trouble with Numpy string arrays where the elements have
a maximum length, as here.  Remember, the `dtype` of the array tells you what
type of element the array can hold.  Here the `dtype` is telling you that the
array can hold strings of maximum length 7 characters.  Now imagine trying to
put a longer string into the array — what do you think would happen?

This happens:

```{python opts.label="py_ed"}
# An array of small strings.
small_strings = np.array(['six', 'one', 'two'])
small_strings.dtype
```

```{python opts.label="py_ed"}
# Set a new value for the first element (first string).
small_strings[0] = 'seven'
small_strings
```

Numpy truncates the new string to match the original maximum length.

For that reason, it is often useful to instruct Numpy that you want to use
effectively infinite length strings, by specifying the array `dtype` as
`object` *when you make the array*, like this:

```{python opts.label="py_ed"}
# An array of small strings, but this time, tell Numpy
# that the strings should be of effectively infinite length.
small_strings_better = np.array(['six', 'one', 'two'], dtype=object)
small_strings_better
```

Notice that the code uses a *named function argument* (@sec-named-arguments),
to specify to `np.array` that the array elements should be of type `object`.
This type can store any Python value, and so, when the array is storing
strings, it will use Python's own string values as elements, rather than the
more efficient but more fragile Unicode strings that Numpy uses by default.

```{python opts.label="py_ed"}
# Set a new value for the first element in the new array.
small_strings_better[0] = 'seven'
small_strings_better
```
<!---
End of warning
-->
:::
<!---
End of Python section
-->
:::

As for any {{< var array >}}, you can select elements with *indexing*.  When
you select an element with a given position (index), you get the *string* at
at that position:

```{python opts.label="py_ed"}
# Julian Lincoln Simon's second name.
# (Remember, Python's positions start at 0).
middle_name = arr_of_strings[1]
middle_name
```

```{r opts.label="r_ed"}
# Julian Lincoln Simon's second name
middle_name <- vector_of_strings[2]
middle_name
```

As for numbers, we can compare strings with, for example, the `==` operator,
that asks whether the two strings are equal:

```{python opts.label="py_ed"}
middle_name == 'Lincoln'
```

```{r opts.label="r_ed"}
middle_name == 'Lincoln'
```

## Repeating elements

Now let us go back to the problem of selecting black and white jurors.

We started with the strategy of using numbers 0 through 25 to mean "black" jurors, and 26 through 99 to mean "white" jurors.  We selected values at random from 0 through 99, and then worked out whether the number meant a "black" juror (was less than 26) or a "white" juror (was greater than 25).

It would be good to use strings instead of numbers to identify the potential
jurors.  Then we would not have to remember our coding of 0 through 25 and 26
through 99.

If only there was a way to make {{< var an_array >}} of 100 strings, where 26 of the strings were "black" and 74 were "white".  Then we could select randomly from that array, and it would be immediately obvious that we had a "black" or "white" juror.

Luckily, of course, we can do that, by using the
[`rep`]{.r}[`np.repeat`]{.python} function to construct the {{< var array >}}.

Here is how that works:

```{python opts.label="py_ed"}
# The values that we will repeat to fill up the larger array.
# Use a list to store the sequence of values.
juror_types = ['black', 'white']
# The number of times we want to repeat "black" and "white".
# Use a list to store the sequence of values.
repeat_nos = [26, 74]
# Repeat "black" 26 times and "white" 74 times.
# We have passed two lists here, but we could also have passed
# arrays - the Numpy repeat function converts the lists to arrays
# before it builds the repeats.
jury_pool = np.repeat(juror_types, repeat_nos)
# Show the result
jury_pool
```

```{r opts.label="r_ed"}
# The values that we will repeat to fill up the larger array.
juror_types <- c('black', 'white')
# The number of times we want to repeat "black" and "white".
repeat_nos <- c(26, 74)
# Repeat "black" 26 times and "white" 74 times.
jury_pool <- rep(juror_types, repeat_nos)
# Show the result
jury_pool
```

We can use this {{< var array >}} of repeats of strings, to sample from. The
result is easier to grasp, because we are using the string labels, instead of
numbers:

```{python opts.label="py_ed"}
# Select one juror at random from the black / white pool.
one_juror = rnd.choice(jury_pool)
one_juror
```

```{r opts.label="r_ed"}
# Select one juror at random from the black / white pool.
# replace=TRUE is redundant here, but we leave it for consistency.
one_juror <- sample(jury_pool, 1, replace=TRUE)
one_juror
```

We can select our full jury of 12 jurors, and see the results in a more obvious
form:

```{python opts.label="py_ed"}
# Select 12 jurors at random from the black / white pool.
one_jury = rnd.choice(jury_pool, 12)
one_jury
```

```{r opts.label="r_ed"}
# Select one juror at random from the black / white pool.
one_jury <- sample(jury_pool, 12, replace=TRUE)
one_jury
```

::: {.callout-note}
## Using the `size` argument to {{< var sample >}}

In the code above, we have specified the *size* of the sample we want (12)
with the second argument to {{< var sample >}}.  As you saw in
@sec-named-arguments, we can also give names to the function arguments, in
this case, to make it clearer what we mean by "12" in the code above.  In
fact, from now on, that is what we will do; we will specify the *size* of our
sample by using the *name* for the function argument to {{< var sample >}} —
`size` — like this:

```{python opts.label="py_ed"}
# Select 12 jurors at random from the black / white pool.
# Specify the sample size using the "size" named argument.
one_jury = rnd.choice(jury_pool, size=12)
one_jury
```

```{r opts.label="r_ed"}
# Select one juror at random from the black / white pool.
# Specify the sample size using the "size" named argument.
one_jury <- sample(jury_pool, size=12, replace=TRUE)
one_jury
```
:::

We can use `==` on the {{< var array >}} to get {{< var true >}} values where
the juror was "black" and {{< var false >}} values otherwise:

```{python opts.label="py_ed"}
are_black = one_jury == 'black'
are_black
```

```{r opts.label="r_ed"}
are_black <- one_jury == 'black'
are_black
```

Finally, we can [`sum`]{.r}[`np.sum`]{.python} to find the number of black
jurors (@sec-count-with-sum):

```{python opts.label="py_ed"}
# Number of black jurors in this simulated jury.
n_black = np.sum(are_black)
n_black
```

```{r opts.label="r_ed"}
# Number of black jurors in this simulated jury.
n_black <- sum(are_black)
n_black
```

Putting that all together, this is our new procedure to select one jury and
count the number of black jurors:

```{python opts.label="py_ed"}
one_jury = rnd.choice(jury_pool, size=12)
are_black = one_jury == 'black'
n_black = np.sum(are_black)
n_black
```

```{r opts.label="r_ed"}
one_jury <- sample(jury_pool, size=12, replace=TRUE)
are_black <- one_jury == 'black'
n_black <- sum(are_black)
n_black
```

Or we can be even more compact by putting several statements together into one
line:

```{python opts.label="py_ed"}
# The same as above, but on one line.
n_black = np.sum(rnd.choice(jury_pool, size=12) == 'black')
n_black
```

```{r opts.label="r_ed"}
# The same as above, but on one line.
n_black = sum(sample(jury_pool, size=12, replace=TRUE) == 'black')
n_black
```

## Resampling with and without replacement

Now let us return to the details of Robert Swain's case, that you first saw in
@sec-resampling-two.

We looked at the composition of Robert Swain's 12-person jury — but in fact, by
law, that does not have to be representative of the eligible jurors.  The
12-person jury is drawn from a jury *panel*, of 100 people, and this should, in
turn, be drawn from the population of all eligible jurors in the county,
consisting, at the time, of "all male citizens in the community over 21 who
are reputed to be honest, intelligent men and are esteemed for their integrity,
good character and sound judgment."  So, unless there was some bias against
black jurors, we might expect the 100-person jury panel to be a plausibly
random sample of the eligible jurors, of whom 26% were black. See [the Supreme
Court case judgement](https://supreme.justia.com/cases/federal/us/380/202) for
details.

In fact, in Robert Swain's trial, there were 8 black members in the 100-person
jury panel. We will leave it to you to adapt the simulation from
@sec-resampling-two to ask the question — is 8% surprising as a random sample
from a population with 26% black people?

But we have a different question: given that 8 out of 100 of the jury panel
were black, is it surprising that none of the 12-person jury were black?  As
usual, we can answer that question with simulation.

Let's think about what a single simulated jury selection would look like.

First we compile a representation of the actual jury panel, using the tools we
have used above.

```{python opts.label="py_ed"}
juror_types = ['black', 'white']
# in fact there were 8 black jurors and 92 white jurors.
panel_nos = [8, 92]
jury_panel = np.repeat(juror_types, panel_nos)
# Show the result
jury_panel
```

```{r opts.label="r_ed"}
juror_types <- c('black', 'white')
# in fact there were 8 black jurors and 92 white jurors.
panel_nos <- c(8, 92)
jury_panel <- rep(juror_types, panel_nos)
# Show the result
jury_panel
```

Now consider taking a 12-person jury at random from this panel.  We select the
first juror at random, so that juror has an 8 out of 100 chance of being black.
But when we select the second jury member, the situation has changed slightly.
We can't select the first juror again, so our panel is now 99 people.   If our
first juror was black, then the chances of selecting another black juror next
are not 8 out of 100, but 7 out of 99 — a smaller chance.  The problem is, as
we shall see in more detail later, the chances of getting a black juror as the
second, and third and fourth members of the jury depend on whether we selected
a black juror as the first and second and third jury members.  At its most
extreme, imagine we had already selected eight jurors, and by some strange
chance, all eight were black.   Now our chances of selecting a black juror as
the ninth juror are zero — there are no black jurors left to select from the
panel.

In this case we are selecting jurors from the panel *without replacement*,
meaning, that once we have selected a particular juror, we cannot select them
again, and we do not put them back into the panel when we select our next
juror.

This is the probability equivalent of the situation when you are dealing a hand
of cards.  Let's say someone is dealing you, and you only, a hand of five
cards.  You get an ace as your first card.   Your chances of getting an ace as
your first card were just the number of aces in the deck divided by the number
of cards — four in 52 -- $\frac{4}{52}$.  But for your second card, the
probability has changed, because there is one less ace remaining in the pack,
and one less card, so your chances of getting an ace as your second card are
now $\frac{3}{51}$.  This is sampling without replacement — in a normal game,
you can't get the same card twice.  Of course, you could imagine getting a hand
where you sampled *with replacement*.  In that case, you'd get a card, you'd
write down what it was, and you'd give the card back to the dealer, who would
*replace* the card in the deck, shuffle again, and give you another card.

As you can see, the chances change if you are sampling with or without
replacement, and the kind of sampling you do, will dictate how you model your
chances in your simulations.

Because this distinction is so common, and so important, the machinery you have
already seen in [`sample`]{.r}[`rnd.choice`]{.python} has simple ways for you
to select your sampling type.   You have already seen sampling *with
replacement*, and it looks like this:

```{python opts.label="py_ed"}
# Take a sample of 12 jurors from the panel *with replacement*
# With replacement is the default for `rnd.choice`.
strange_jury = rnd.choice(jury_panel, size=12)
strange_jury
```

```{r opts.label="r_ed"}
# Take a sample of 12 jurors from the panel *with replacement*
strange_jury <- sample(jury_panel, size=12, replace=TRUE)
strange_jury
```

This is a strange jury, because it can select any member of the jury pool more
than once.  Perhaps that juror would have to fill two (or more!) seats, or run
quickly between them.   But of course, that is not how juries are selected.
They are selected *without replacement*:

```{python opts.label="py_ed"}
# Take a sample of 12 jurors from the panel *without replacement*
ok_jury = rnd.choice(jury_panel, 12, replace=False)
ok_jury
```

::: r
Thus far, we have always done sampling *with replacement*, and, in order to do
that with `sample`, we pass the argument `replace=TRUE`. We do that because the
default for `sample` is `replace=FALSE`, that is, by default, `sample` does
sampling without replacement.  If you want to do sampling without replacement,
you can just omit the `replace=TRUE` argument to `sample`, or you can specify
`replace=FALSE` explicitly, perhaps to remind yourself that this is sampling
without replacement. Whether you omit the `replace` argument, or specify
`replace=FALSE`, the behavior is the same.
::::

```{r opts.label="r_ed"}
# Take a sample of 12 jurors from the panel *with replacement*
ok_jury <- sample(jury_panel, size=12)  # replace=FALSE is the default.
ok_jury
```

To finish the procedure for simulating a single jury selection, we count the
number of black jurors:

```{python opts.label="py_ed"}
n_black = np.sum(ok_jury == 'black')
n_black
```

```{r opts.label="r_ed"}
n_black <- sum(ok_jury == 'black')
n_black
```

Now we have the procedure for one simulated trial, here is the procedure for
10000 simulated trials.

```{python opts.label="py_ed"}
counts = np.zeros(10000)
for i in np.arange(10000):
    # Single trial procedure
    jury = rnd.choice(jury_panel, size=12, replace=False)
    n_black = np.sum(jury == 'black')
    # Store the result
    counts[i] = n_black

# Number of juries with 0 black jurors.
zero_black = np.sum(counts == 0)
# Proportion
p_zero_black = zero_black / 10000
print(p_zero_black)
```

```{r opts.label="r_ed"}
counts <- numeric(10000)
for (i in 1:10000) {
    # Single trial procedure
    jury <- sample(jury_panel, size=12)
    n_black <- sum(jury == 'black')
    # Store the result
    counts[i] <- n_black
}
# Number of juries with 0 black jurors.
zero_black <- sum(counts == 0)
# Proportion
p_zero_black <- zero_black / 10000
message(p_zero_black)
```

We have found that,  when there are only 8% black jurors in the jury panel,
having no black jurors in the final jury happens about 
`r round(get_var('p_zero_black') * 100)`%
of the time, even in this case, where the jury is selected completely at random
from the jury panel.

We should look for the main source of bias in the initial selection of the jury
panel, not in the selection of the jury from the panel.

<!---
End of notebook
-->

:::

::: {.callout-note}
## With or without replacement for the original jury selection

You may have noticed in @sec-resampling-two that we were sampling Robert
Swain's jury from the eligible pool of jurors, *with replacement*.  You might
reasonably ask whether we should have selected from the eligible jurors
*without replacement*, given that the same juror cannot serve more than once in
the same jury, and therefore, the same argument applies there as here.

The trick there was that we were selecting from a very large pool of many
thousand eligible jurors, of whom 26% were black.  Let's say there were 10,000
eligible jurors, of whom 2,600 were black.  When selecting the first juror,
there is exactly a 2,600 in 10,000 chance of getting a black juror — 26%.  And
yes, it's true, that the chance will have changed very slightly for the second
juror — if the first juror was black, the chance that the second juror will be
black is 2,599 in 9,999.  But these changes are very small; even if we select
eleven black jurors out of eleven, when we come to the twelfth juror, we still
have a 2,589 out of 9,989 chance of getting another black juror, and that works
out at a 25.92% chance — hardly changed from the original 26%.  So yes, you'd
be right, we really should have compiled our population of 2,600 black jurors
and 7,400 white jurors, and then sampled without replacement from that
population, but as the resulting sample probabilities will be very similar to
the simpler sampling with replacement, we're afraid we chose to slide that one
quietly past you, in the hope you would forgive us when you realized.
:::

## Conclusion

This chapter introduced you to the idea of strings — values in {{< var lang >}}
that store bits of text.  Strings are very useful as labels for the entities we
are sampling from, when we do our simulations.  Strings are particularly useful
when we use them with {{< var array >}}s, and one way we often do that is to
build up {{< var array >}}s of strings to sample from, using the
[`rep`]{.r}[`np.repeat`]{.python} function.

There is a fundamental distinction between two different types of sampling —
sampling *with replacement*, where we draw an element from a larger pool, then
put that element back before drawing again, and sampling *without replacement*,
where we remove the element from the remaining pool when we draw it into the
sample.   As we will see later, it is often a judgment call which of these two
types of sampling is a more reasonable model of the world you are trying to
simulate.
