---
jupyter:
  jupytext:
    metadata_filter:
      notebook:
        additional: all
        excluded:
        - language_info
    text_representation:
      extension: .Rmd
      format_name: rmarkdown
      format_version: '1.0'
      jupytext_version: 0.8.6
  kernelspec:
    display_name: Python 3
    language: python
    name: python3
resampling_with:
    ed2_fname: null
---

```{r setup, include=FALSE}
source("_common.R")
```

# Tools for samples and sampling {#sec-sampling-tools}

## Introduction

Now you have some experience with {{< var lang >}}, probabilities and
resampling, it is time to introduce some useful tools for our experiments and
programs.

## Samples and labels

Thus far we have used numbers such as 1 and 0 and 10 to represent the elements
we are sampling from.  For example, in @sec-resampling-code, we were simulating the chance of a particular patient being cured, given the overall standard cure rate of 90%.   We used *numbers* for that task, where 0 represented not-cured, and numbers 1 through 9 represented cured.  For example, this is the procedure for getting one simulated patient in @sec-sampling-arrays:

```{python, opts.label="py_ed"}
# Get a single value at random from
# numbers 0 through 9
a = rnd.choice(range(10))
# Show the result.
a
```

```{r, opts.label="r_ed"}
# Get a single value at random from
# numbers 0 through 9.
a <- sample(0:9)
# Show the result.
a
```

We considered the patient in `a` *not cured* if `a` was equal to 0, and cured otherwise.
