---
resampling_with:
    ed2_fname: 19-Chap-15
jupyter:
  jupytext:
    notebook_metadata_filter: all,-language_info
    split_at_heading: true
    text_representation:
      extension: .Rmd
      format_name: rmarkdown
      format_version: '1.2'
      jupytext_version: 1.15.2
  kernelspec:
    display_name: Python 3 (ipykernel)
    language: python
    name: python3
---

```{r setup, include=FALSE}
source("_common.R")
```

# Hypothesis-Testing with Counted Data, Part 1 {#sec-testing-counts-one}

## Introduction

The first task in inferential statistics is to make one or more *point
estimates* — that is, to make one or more statements about *how much*
there is of something we are interested in — including especially the
mean and the dispersion. (That work goes under the label "estimation"
and is discussed in @sec-point-estimation.) Frequently the next step, after
making such quantitative estimation of the universe from which a sample has
been drawn, is to consider whether two or more samples are different from each
other, or whether the single sample is different from a specified value; this
work goes under the label "hypothesis testing." We ask: Did something happen?
Or: Is there a difference between two universes? These are yes-no questions.

In other cases, the next step is to inquire into the reliability of the
estimates; this goes under the label "confidence intervals." (Some
writers include assessing reliability under the rubric of estimation,
but I judge it better not to do so).

So: Having reviewed how to convert hypothesis-testing problems into
statistically testable questions in @sec-framing-questions, we now must ask:
How does one employ resampling methods to make the statistical test? As is
always the case when using resampling techniques, there is no unique series of
steps by which to proceed. The crucial criterion in assessing the model is
whether it accurately simulates the actual event. With hypothesis-testing
problems, any number of models may be correct. Generally speaking, though, the
model that makes fullest use of the quantitative information available from
the data is the best model.

When attempting to deduce the characteristics of a universe from sample
data, or when asking whether a sample was drawn from a particular
universe, a crucial issue is whether a "one-tailed test" or a
"two-tailed test" should be applied. That is, in examining the results
of our resampling experiment based on the benchmark universe, do we
examine both ends of the frequency distribution, or just one? If there
is strong reason to believe *a priori* that the difference between the
benchmark (null) universe and the sample will be in a given
direction — for example if you hypothesize that the sample mean will be
*smaller* than the mean of the benchmark universe — you should then
employ a *one-tailed test* . If you do *not* have strong basis for such
a prediction, use the *two-tailed* test. As an example, when a scientist
tests a new medication, his/her hypothesis would be that the number of
patients who get well will be higher in the treated group than in the control
group. Thus, s/he applies the one-tailed test.  See the text below for more
detail on one- and two-tailed tests.

Some language first:

**Hypothesis:** In inferential statistics, a statement or claim about a
universe that can be tested and that you wish to investigate.

**Testing:** The process of investigating the validity of a hypothesis.

**Benchmark (or null) hypothesis:** A particular hypothesis chosen for
convenience when testing hypotheses in inferential statistics. For
example, we could test the hypothesis that there is *no difference*
between a sample and a given universe, or between two samples, or that a
parameter is less than or greater than a certain value. The benchmark
universe refers to this hypothesis. (The concept of the benchmark or
null hypothesis was discussed in @sec-prob-theory-one-b and
@sec-framing-questions.)

Now let us begin the actual statistical testing of various sorts of
hypotheses about samples and populations.

## Should a single sample of counted data be considered different from a benchmark universe?

#### Example: Does Irradiation Affect the Sex Ratio in Fruit Flies? {#sec-fruitfly}

**Where the Benchmark Universe Mean (in this case, the Proportion) is Known, is
the Mean (Proportion) of the Population Affected by the Treatment?)**

You think you have developed a technique for irradiating the genes of
fruit flies so that the sex ratio of the offspring will *not* be half
males and half females. In the first twenty cases you treat, there are
fourteen males and six females. Does this experimental result confirm
that the irradiation does work?

First convert the scientific question — whether or not the treatment
affects the sex distribution — into a probability-statistical question:
Is the observed sample likely to have come from a benchmark universe in
which the sex ratio is one male to one female? The benchmark (null)
hypothesis, then, is that the treatment makes no difference and the
sample comes from the one-male-to-one-female universe. Therefore, we
investigate *how likely a one-to-one universe is to produce a
distribution of fourteen or more of just one sex*.

A coin has a one-to-one (one out of two) chance of coming up tails. Therefore,
we might flip a coin in groups of twenty flips, and count the number of tails
in each twenty flips. Or we can use a random number table. The following steps
will produce a sound estimate:

* **Step 1.** Let tails = male, heads = female.
* **Step 2.** Flip twenty coins and count the number of males. If 14 or more
  males occur, record "yes." Also, if 6 or fewer males occur, record "yes"
  because this means we have gotten 14 or more females. Otherwise, record "no."
* **Step 3.** Repeat step 2 perhaps 100 times.
* **Step 4.** Calculate the proportion "yes" in the 100 trials. This
  proportion estimates the probability that a fruit-fly population with
  a propensity to produce 50 percent males will by chance produce as many as 14
  or as few as 6 males in a sample of 20 flies.


```{r echo=FALSE, eval=TRUE}
# Calculate some values for the text below.
flies <- read.csv('data/fruitfly_trials.csv', check.names=FALSE)
n <- 25
counts <- head(flies[['# of tails']], n)
n_25_gte_14 <- sum(counts >= 14)
n_25_lte_6 <- sum(counts <= 6)
n_25_both <- n_25_gte_14 + n_25_lte_6
# Show the table.
ketable(head(flies, n),
        caption = "Results from 25 random trials for Fruitfly problem {#tbl-fruitfly-trials}")
```

@tbl-fruitfly-trials shows the results obtained in `r int2tw(n)` trials of
twenty flips each. In `r int2tw(n_25_gte_14)` of the twenty-five trials
(`r round(n_25_gte_14 / n * 100)` percent) there were fourteen or more tails,
which we call "males," and in `r int2tw(n_25_lte_6)` of the `r int2tw(n)`
trials (`r round(n_25_lte_6 / n * 100)` percent) there six or fewer tails,
meaning there were fourteen or more heads ("females"). We can therefore
estimate that, even if the treatment does *not* affect the sex and the births
over a long period really are one to one, `r int2tw(n_25_both)` out of
twenty-five times (`r round(n_25_both / n * 100)` percent) we would get
fourteen or more of one sex or the other. Therefore, finding fourteen males
out of 20 births is not overwhelming evidence that the treatment has any
effect, even though the result is suggestive.

```{r, eval=TRUE, echo=FALSE}
counts <- flies[['# of tails']]
n <- length(counts)
n_gte_14 <- sum(counts >= 14)
n_lte_6 <- sum(counts <= 6)
n_both <- n_gte_14 + n_lte_6
```

How accurate is the estimate? Seventy-five more trials were made, and of
the 100 trials `r int2tw(n_gte_14)` contained fourteen or more "males"
(`r round(n_gte_14 / n * 100)` percent), and  `r round(n_lte_6 / n * 100)`
trials contained fourteen or more "females" (`r round(n_lte_6 / n * 100)`
percent), a total of `r round(n_both / n * 100)` percent. So the first
twenty-five trials gave a fairly reliable indication. As a matter of fact,
analytically-based computation (not explained here) shows that the probability
of getting fourteen or more females out of twenty births is .057 and, of
course, the same for fourteen or more males from a one-to-one universe,
implying a total probability of .114 of getting fourteen or more males *or*
females.

Now let us obtain larger and more accurate simulation samples with the
computer. The key step in the {{< var lang >}} notebook below represents
male fruit flies with the string `'male'` and female fruit flies with the
string `'female'`. The [`sample`]{.r}[`rnd.choice`]{.python} function is then
used to generate 20 of these strings with an equal probability that either
string is selected. This simulates randomly choosing 20 fruit flies on the
benchmark assumption — the "null hypothesis" — that each fruit fly has an equal
chance of being a male or female. Now we want to discover the chances of
getting more than 13 (i.e., 14 or more) males or more than 13 females under
these conditions. So we use [`sum`]{.r}[`np.sum`]{.python} to count the number
of males in each random sample and then store this value in the `scores` {{<
var array >}} of this number for each sample. We repeat these steps 10,000
times.

After ten thousand samples have been drawn, we count (`sum`) how often there
were more than 13 males and then count the number of times there were fewer
than 7 males (because if there were fewer than 7 males there must have been
more than 13 females). When we add the two results together we have the
probability that the results obtained from the sample of irradiated fruit flies
would be obtained from a random sample of fruit flies.

::: {.notebook name="fruit_fly" title="Fruit fly simulation"}

::: nb-only
This notebook uses simulation to test the null hypothesis that it is equally
likely that new fruit files are male or female.
:::

```{python}
import numpy as np
import matplotlib.pyplot as plt

# set up the random number generator
rnd = np.random.default_rng()
```

```{python}
# Set the number of trials
n_trials = 10000

# set the sample size for each trial
sample_size = 20

# An empty array to store the trials
scores = np.zeros(n_trials)

# Do 10000 trials
for i in range(n_trials):

    # Generate 20 simulated fruit flies, where each has an equal chance of
    # being male or female
    a = rnd.choice(['male', 'female'],
                   size=sample_size,
                   p=[0.5, 0.5],
                   replace = True)

    # count the number of males in the sample
    b = np.sum(a == 'male')

    # store the result of this trial
    scores[i] = b

# Produce a histogram of the trial results
plt.title(f"Number of males in {n_trials} samples of \n{sample_size} simulated fruit flies")
plt.hist(scores)
plt.xlabel('Number of Males')
plt.ylabel('Frequency')
```

```{r}
# Set the number of trials
n_trials <- 10000

# set the sample size for each trial
sample_size <- 20

# An empty array to store the trials
scores <- numeric(n_trials)

# Do 10000 trials
for (i in 1:n_trials) {
    # Generate 20 simulated fruit flies, where each has an equal chance of being
    # male or female
    a <- sample(c('male', 'female'),
                size=sample_size,
                prob=c(0.5, 0.5),
                replace=TRUE)

    # count the number of males in the sample
    b <- sum(a == 'male')

    # store the result of this trial
    scores[i] <- b
}

# Produce a histogram of the trial results
title_of_plot <- paste0("Number of males in", n_trials, " samples of \n", sample_size, " simulated fruit flies")
hist(scores, xlab='Number of Males', main=title_of_plot)
```

In the histogram above, we see that in 16 percent of the trials, the number of
males was 14 or more, or 6 or fewer. Or instead of reading the results from the
histogram, we can calculate the result by tacking on the following commands to
the above program:


```{python}
# Determine the number of trials in which we had 14 or more males.
j = np.sum(scores >= 14)

# Determine the number of trials in which we had 6 or fewer males.
k = np.sum(scores <= 6)

# Add the two results together.
m = j + k

# Convert to a proportion.
mm = m / n_trials

# Print the results.
print(mm)
```

```{r}
# Determine the number of trials in which we had 14 or more males.
j <- sum(scores >= 14)

# Determine the number of trials in which we had 6 or fewer males.
k <- sum(scores <= 6)

# Add the two results together.
m <- j + k

# Convert to a proportion.
mm <- m/n_trials

# Print the results.
print(mm)
```

:::
<!---
End of notebook
-->

Notice that the strength of the evidence for the effectiveness of the
radiation treatment depends upon the original question: whether or not
the treatment had *any* effect on the sex of the fruit fly, which is a
two-tailed question. If there were reason to believe at the start that
the treatment could increase *only* the number of *males* , then we
would focus our attention on the result that in only
`r int2tw(n_25_gte_14)` of the twenty-five trials were fourteen or more males.
There would then be only a `r n_25_gte_14`/25 =
`r round(n_25_gte_14 / 25, 2)`  probability of getting the observed
results by chance if the treatment really has no effect, rather than the weaker
odds against obtaining fourteen or more of *either* males or females.

Therefore, whether you decide to figure the odds of just fourteen or
more males (what is called a "one-tail test") or the odds for fourteen
or more males *plus* fourteen or more females (a "two-tail test"),
depends upon your advance knowledge of the subject. If you have no
reason to believe that the treatment will have an effect *only* in the
direction of creating more males and if you figure the odds for the
one-tail test anyway, then you will be kidding yourself. Theory comes to
bear here. If you have a strong hypothesis, deduced from a strong
theory, that there will be more males, then you should figure one-tail
odds, but if you have no such theory you should figure the weaker
two-tail odds.[^maybe-bayes]

[^maybe-bayes]: If you are very knowledgeable, you may do some in-between
    figuring (with what is known as "Bayesian analysis"), but leave this alone
    unless you know well what you are doing.

In the case of the next problem concerning calves, we shall see that a
one-tail test is appropriate because we have no interest in producing
more male calves. Before leaving this example, let us review our
intellectual strategy in handling the problem. First we observe a result
(14 males in 20 flies) which differs from the proportion of the
benchmark population (50 percent males). Because we have treated this
sample with irradiation and observed a result that differs from the
untreated benchmark-population's mean, we speculate that the irradiation
caused the sample to differ from the untreated population. We wish to
check on whether this speculation is correct.

When asking whether this speculation is correct, we are implicitly
asking whether future irradiation would also produce a proportion of
males higher than 50 percent. That is, we are implicitly asking whether
irradiated flies would produce more samples with male proportions as
high as 14/20 than would occur by chance in the absence of irradiation.

If samples as far away as 14/20 from the benchmark population mean of
10/20 would occur frequently by chance, then we would not be impressed
with that experimental evidence as proof that irradiation does affect
the sex ratio. Hence we set up a model that will tell us the frequency
with which samples of 14 or more males out of 20 births would be
observed by chance. Carrying out the resampling procedure tells us that
perhaps a tenth of the time such samples would be observed by chance.
That is not extremely frequent, but it is not infrequent either. Hence
we would probably conclude that the evidence is provocative enough to
justify further experimentation, but not so strong that we should
immediately believe in the truth of this speculation.

The logic of attaching meaning to the probabilistic outcome of a test of
a hypothesis is discussed in @sec-significance-concept. There also is more
about the concept of the level of significance in @sec-significance-concept.

Because of the great importance of this sort of case, which brings out
the basic principles particularly clearly, let us consider another
example:

### Example: Does a treatment increase the female calf rate? {#sec-female-calves}

**What is the probability that among 10 calves born, 9 or more will be
female?**

Let's consider this question in the context of a set of queries for
performing statistical inference that will be discussed further in
@sec-testing-procedures.

***The question***: (From @hodges1970basic): Female calves are more valuable
than males. A bio-engineer claims to be able to cause more females to be born
than the expected 50 percent rate. He conducts his procedure, and nine females
are born out of the next 10 pregnancies among the treated cows. Should you
believe his claim? That is, what is the probability of a result this (or more)
surprising occurring by chance if his procedure has no effect? In this problem,
we assume that on average 100 of 206 births are female, in contrast to the
50-50 benchmark universe in the previous problem.

***What is the purpose of the work?***: Female calves are more valuable than
male calves.

***Statistical inference?***: Yes.

***Confidence interval or Test of hypothesis?***: Test of hypothesis.

***Will you state the costs and benefits of various outcomes, or a loss
function?***: Yes. One need only say that the benefits are very large, and if
the results are promising, it is worth gathering more data to confirm results.

***How many samples of data are part of the hypothesis test?***: 
One.

***What is the size of the first sample about which you wish to make
significance statements?***: Ten.

***What comparison(s) to make?***: Compare the sample to the benchmark
universe.

***What is the benchmark universe***: **that embodies the null hypothesis?**
100/206 female.

***Which symbols for the observed entities?***: Balls in bucket, or numbers.

***What values or ranges of values?***: We could write numbers 1 through 206 on
pieces of paper, and take numbers 1-100 as "male" and 101-206 as "female".  Or
we could use some other mechanism to give us a 100/206 chance of any one calf
being female.

***Finite or infinite universe?***: Infinite.

***Which sample(s) do you wish to compare to which, or to the null
universe (and perhaps to the alternative universe)?***: Ten calves.

***What procedure to produce the sample entities?***: Sampling with
replacement.

***Simple (single step) or complex (multiple "if" drawings)?***: Can
think of it either way.

***What to record as the outcome of each resample trial?***: The proportion (or
number) of females.

***What is the criterion to be used in the test?***: The probability that in
a sample of ten calves, nine (or more) females would be drawn by chance from
the benchmark universe of 100/206 females.

***"One tail" or "two tail" test?***: One tail, because the farmer is only
interested in females. Finding a large proportion of males would not be of
interest; it would not cause rejecting the null hypothesis.

The actual computation of probability may be done in several ways, as discussed
earlier for four children and for ten cows. Conventional methods are discussed
for comparison in @sec-testing-procedures. Here is the resampling solution in
{{< var lang >}}.

::: {.notebook name="female_calves" title="Female calf numbers simulation"}

::: nb-only
This notebook uses simulation to test the null hypothesis that the chances of
any one calf being female is 100 / 206.
:::

```{python}
import numpy as np
import matplotlib.pyplot as plt

rnd = np.random.default_rng()

# Set the number of trials.
n_trials = 10000

# Set the size of each sample.
sample_size = 10

# Probability of any one calf being female.
p_female = 100 / 206

# An array to store the results.
scores = np.zeros(n_trials)

# For 10000 repeats.
for i in range(n_trials):

    a = rnd.choice(['female', 'male'],
                   p=[p_female, 1 - p_female],
                   size=sample_size)
    b = np.sum(a == 'female')

    # Store the result of the current trial.
    scores[i] = b

# Plot a histogram of the scores.
plt.title(f"Number of females in {n_trials} samples of \n{sample_size} simulated calves")
plt.hist(scores)
plt.xlabel('Number of Females')
plt.ylabel('Frequency')

# Count the number of scores that were greater than or equal to 9.
k = np.sum(scores >= 9)

# Express as a proportion.
kk = k / n_trials

# Show the proportion.
print("Probability of 9 or 10 females occurring by chance:", kk)
```

```{r}
# Set the number of trials.
n_trials <- 10000

# Set the size of each sample.
sample_size <- 10

# Probability of any one calf being female.
p_female <- 100 / 206

# A vector to store the results.
scores <- numeric(n_trials)

# for 10000 repeats.
for (i in 1:n_trials) {

    a <- sample(c('female', 'male'),
                size=sample_size,
                prob=c(p_female, 1 - p_female),
                replace=TRUE)
    b <- sum(a == 'female')

    # Store the result of the current trial.
    scores[i] <- b
}

# Plot a histogram of the scores.
title_of_plot <- paste0("Number of females in", n_trials, " samples of \n", sample_size, " simulated calves")
hist(scores, xlab = 'Number of Females', main=title_of_plot)

# Count the number of scores that were greater than or equal to 9.
k <- sum(scores >= 9)

# Express as a proportion.
kk <- k / n_trials

# Show the proportion.
message("Probability of 9 or 10 females occurring by chance: ", kk)
```

We read from the result in variable `kk` that the probability of 9 or 10
females occurring by chance is a bit more than one percent.

:::
<!---
End of calves notebook.
-->

### Example: A Public-Opinion Poll {#sec-contract-poll}

**Is the Proportion of a Population Greater Than a Given Value?**

A municipal official wants to determine whether a majority of the town's
residents are for or against the awarding of a high-speed internet
contract, and she asks you to take a poll. You judge that the voter
registration records are a fair representation of the universe in which the
politician was interested, and you therefore decided to interview a random
selection of registered voters. Of a sample of fifty people who expressed
opinions, thirty said "yes" they were for the plan and twenty said "no," they
were against it. How conclusively do the results show that the people in town
want this internet contract?

Now comes some necessary subtle thinking in the interpretation of what
seems like a simple problem. Notice that our aim in the analysis is to
avoid the mistake of saying that the town favors the plan when in fact
it does not favor the plan. Our chance of making this mistake is
greatest when the voters are evenly split, so we choose as the benchmark
(null) hypothesis that 50 percent of the town does not want the plan.
This statement really means that "50 percent or more do not want the
plan." We could assess the probability of obtaining our result from a
population that is split (say) 52-48 against, but such a probability
would necessarily be even smaller, and we are primarily interested in
assessing the maximum probability of being wrong. If the maximum
probability of error turns out to be inconsequential, then we need not
worry about less likely errors.

This problem is very much like the one-group fruit fly irradiation problem
above. The only difference is that now we are comparing the observed sample
against an arbitrary value of 50 percent (because that is the break-point in
a situation where the majority decides) whereas in @sec-fruitfly we compared
the observed sample against the normal population proportion (also 50 percent,
because that is the normal proportion of males). But it really does not matter
*why* we are comparing the observed sample to the figure of 50 percent; the
procedure is the same in both cases. (Please notice that there is nothing
special about the 50 percent figure; the same procedure would be followed for
20 percent or 85 percent.)

In brief, we a) take two pieces of paper, write "Yes" on one and "No" on the
other, put them in a bucket b) draw a piece of paper from the bucket, record
whether it was "Yes" or "No", replace, and repeat 50 times c) count the number
of "yeses" and "noes" in the fifty draws, c) repeat for perhaps a hundred
trials, then d) count the proportion of the trials in which a 50-50 universe
would produce thirty or more "yes" answers.

In operational steps, the procedure is as follows:

* **Step 1.** "1-5" = no, "6-0" = yes.
* **Step 2.** In 50 random numbers, count the "yeses," and record "false
  positive" if 30 or more "yeses."
* **Step 3.** Repeat step 2 perhaps 100 times.
* **Step 4.** Calculate the proportion of experimental trials showing
  "false positive." This estimates the probability that as many as 30 "yeses"
  would be observed by chance in a sample of 50 people *if* half (or more) are
  really against the plan.

```{python, eval=TRUE, echo=FALSE}
import numpy as np
import pandas as pd

# We need the seed to make the results match the text below, and to match the
# how_big_sample discussion.
rnd = np.random.default_rng(1914)

n = 20
yeses = np.zeros(n, dtype=int)
for i in range(n):
    answers = rnd.choice(['No', 'Yes'], size=50)
    yeses[i] = np.sum(answers == 'Yes')
contract_trials = pd.DataFrame(
    {'Trial no': np.arange(1, n + 1),
     '# of "Noes"': 50 - yeses,
     '# of "Yeses"': yeses,
     '>= 30 "Yeses"': np.where(yeses >= 30, '+', '')
     })
```

```{r echo=FALSE, eval=TRUE}
yeses <- py$yeses
n <- length(yeses)
n_gte_30 <- sum(yeses >= 30)
gte_30_pct <- round(n_gte_30 / n * 100)
# Show the table.
ketable(py$contract_trials,
        caption = "Results from 20 random trials for contract poll problem {#tbl-contract-trials}")
```

In @tbl-contract-trials, we see the results of twenty trials; `r n_gte_30` of
`r n` times (`r gte_30_pct` percent), 30 or more "yeses" were observed by
chance. So our "significance level" or "probability value" is `r gte_30_pct`
percent, which is normally too high to feel confident that our poll results are
reliable. This is the probability that as many as thirty of fifty people would
say "yes" by chance if the population were "really" split evenly. (If the
population were split so that *more* than 50 percent were against the plan, the
probability would be even *less* that the observed results would occur by
chance. In this sense, the benchmark hypothesis is conservative). On the other
hand, if we had been counting the number of times there are 30 or more "No"
votes that, in our setup, have the same odds as to 30 or more "Yes" votes,
there would have been `r int2tw(sum(yeses <= 20))`. This indicates how samples
can vary just by chance.

Taken together, the evidence suggests that the official would be wise not to
place very much confidence in the poll results, but rather ought to act with
caution or else take a larger sample of voters.

::: {.notebook name="contract_poll" title="Contract poll simulation"}

This {{< var lang >}} notebook generates samples of 50 simulated voters on the
assumption that only 50 percent are in favor of the contract. Then it counts
(`sum`s) the number of samples where over 29 (30 or more) of the 50 respondents
said they were in favor of the contract. (That is, we use a "one-tailed test.")
The result in the `kk` variable is the chance of a "false positive," that is,
30 or more people saying they favor a contract when support for the proposal is
actually split evenly down the middle.

```{python}
import numpy as np
import matplotlib.pyplot as plt

rnd = np.random.default_rng()

# We will do 10,000 iterations.
n = 10_000

# Make an array of integers to store the "Yes" counts.
yeses = np.zeros(n, dtype=int)

for i in range(n):
    answers = rnd.choice(['No', 'Yes'], size=50)
    yeses[i] = np.sum(answers == 'Yes')

# Produce a histogram of the trial results.
# Use integer bins for histogram, from 10 through 40.
plt.hist(yeses, bins=range(10, 41))
plt.title('Number of yes votes out of 50, in null universe')
```

```{r}
# We will do 10,000 iterations.
n <- 10000

# Make an array of integers to store the "Yes" counts.
yeses <- numeric(n)

for (i in 1:n) {
    answers <- sample(c('No', 'Yes'), size=50, replace=TRUE)
    yeses[i] <- sum(answers == 'Yes')
}

# Produce a histogram of the trial results.
# Use integer bins for histogram, from 10 through 40.
hist(yeses, breaks=10:40,
     main='Number of yes votes out of 50, in null universe')
```

In the histogram above, we see that about 11 percent of our trials had 30 or
more voters in favor, despite the fact that they were drawn from a population
that was split 50-50. {{< var lang >}} will calculate this proportion directly
if we add the following commands to the above:

```{python}
k = np.sum(yeses >= 30)
kk = k / n
print('Proportion >= 30:', np.round(kk, 2))
```

```{r}
k <- sum(yeses >= 30)
kk <- k / n
message('Proportion >= 30: ', round(kk, 2))
```

:::
<!---
End of contract_poll notebook.
-->

The section above discusses testing hypotheses about a single sample of counted
data relative to a benchmark universe. This section discusses the issue of
whether *two* samples with counted data should be considered the same or
different.

### Example: Did the Trump-Clinton Poll Indicate that Trump Would Win?

::: {.notebook name="trump_clinton" title="Trump/Clinton poll simulation"}

What is the probability that a sample outcome such as actually observed (840
Trump, 660 Clinton) would occur by chance if Clinton is "really" ahead — that
is, if Clinton has 50 percent (or more) of the support? To restate in sharper
statistical language: What is the probability that the observed sample or one
even more favorable to Trump would occur if the universe has a mean of 50
percent or below?

Here is a procedure that responds to that question:

1.  Create a benchmark universe with one ball marked "Trump" and another
    marked "Clinton"
2.  Draw a ball, record its marking, and replace. (We sample with replacement
    to simulate the practically-infinite population of U. S. voters.)
3.  Repeat step 2 1500 times and count the number of "Trump"s. If 840 or
    greater, record "Y"; otherwise, record "N."
4.  Repeat steps 3 and 4 perhaps 1000 or 10,000 times, and count the
    number of "Y"s. The outcome estimates the probability that 840 or
    more Trump choices would occur if the universe is "really" half or
    more in favor of Clinton.

This procedure may be done as follows with {{< var lang >}}.

```{python}
import numpy as np
import matplotlib.pyplot as plt

rnd = np.random.default_rng()

# Number of repeats we will run.
n = 10_000

# Make an integer array to store the counts.
trumps = np.zeros(n, dtype=int)

for i in range(n):
    votes = rnd.choice(['Trump', 'Clinton'], size=1500)
    trumps[i] = np.sum(votes == 'Trump')

# Integer bins from 670 through 830 in steps of 5.
plt.hist(trumps, bins=range(670, 831, 5))
plt.title('Number of Trump voters of 1500 in null-world simulation')

# How often >= 840 Trump votes in random draw?
k = np.sum(trumps >= 840)
# As a proportion of simulated resamples.
kk = k / n

print('Proportion voting for Trump:', kk)
```

```{r}
# Number of repeats we will run.
n <- 10000

# Make an array to store the counts.
trumps <- numeric(n)

for (i in 1:n) {
    votes <- sample(c('Trump', 'Clinton'), size=1500, replace=TRUE)
    trumps[i] <- sum(votes == 'Trump')
}

# Integer bins from 670 through 830 in steps of 5.
hist(trumps, breaks=seq(670, 830, by=5),
     main='Number of Trump voters of 1500 in null-world simulation')

# How often >= 840 Trump votes in random draw?
k <- sum(trumps >= 840)
# As a proportion of simulated resamples.
kk <- k / n

message('Proportion voting for Trump: ', kk)
```

The value for `kk` is our estimate of the probability that Trump's "victory" in
the sample would occur by chance if he really were behind.  In this case, our
probability estimate is less than 1 in 10,000 (< 0.0001).

:::
<!---
End of Trump / Clinton notebook.
-->

<!---
The results of the various runs may be seen in the histogram and printout
following.

**Bin Center Freq Pct Cum Pct**

  ------- ----- ------ ------
  0.465   4     0.4    0.4
  0.47    14    1.4    1.8
  0.475   28    2.8    4.6
  0.48    46    4.6    9.2
  0.485   97    9.7    18.9
  0.49    106   10.6   29.5
  0.495   159   15.9   45.4
  0.5     121   12.1   57.5
  0.505   157   15.7   73.2
  0.51    102   10.2   83.4
  0.515   75    7.5    90.9
  0.52    39    3.9    94.8
  0.525   35    3.5    98.3
  0.53    11    1.1    99.4
  0.535   5     0.5    99.9
  0.54    1     0.1    100
  ------- ----- ------ ------
-->

### Example: Comparison of Possible Cancer Cure to Placebo {#sec-cancer-cures}

**Do Two Binomial Populations Differ in Their Proportions**.

@sec-fruitfly used an observed sample of male and female fruitflies to test the
benchmark (null) hypothesis that the flies came from a universe with
a one-to-one sex ratio, and the poll data problem also compared results to
a 50-50 hypothesis. The calves problem also compared the results to a single
benchmark universe — a proportion of 100/206 females. Now we want to compare
*two samples with each other* , rather than comparing one sample with
a hypothesized universe. That is, in this example we are not comparing one
sample to a benchmark universe, but rather asking whether *both* samples come
from the *same* universe. The universe from which both samples come, *if* both
belong to the same universe, may be thought of as the benchmark universe, in
this case.

The scientific question is whether pill P cures a rare cancer. A researcher
gave pill P to six patients selected randomly from a group of twelve cancer
patients; of the six, five got well. He gave an inactive placebo to the other
six patients, and two of them got well. Does the evidence justify a conclusion
that the pill has a curative effect?

(An identical statistical example would serve for an experiment on methods of
teaching reading to children. In such a situation the researcher would respond
to inconclusive results by running the experiment on more subjects, but in
cases like the cancer-pill example the researcher often cannot obtain more
subjects.)

We can answer the stated question by *combining* the two samples and testing
both samples against the resulting combined universe. In this case, the
universe is twelve subjects, seven (5 + 2) of whom got well. How likely would
such a universe produce two samples as far apart as five of six, and two of
six, patients who get well? In other words, how often will two samples of six
subjects, each drawn from a universe in which 7/12 of the patients get well, be
as far apart as 5 - 2 = 3 patients in favor of the sample designated "pill"?
This is obviously a one-tail test, for we have no reason to believe that the
pill group might do *less* well than the placebo group.

We might construct a twelve-sided die, seven of whose sides are marked "get
well." Or put 12 pieces of paper in a bucket, seven with "get well" and five
with "not well".  Or we would use pairs of numbers from the random-number
table, with numbers "01-07" corresponding to get well, numbers "08-12"
corresponding to "not get well," and all other numbers omitted. (If you wish to
save time, you can work out a system that uses more numbers and skips fewer,
but that is up to you.) Designate the first six subjects "pill" and the next
six subjects "placebo."

The specific procedure might be as follows:

* **Step 1.** Write "get well" on seven pieces of paper, "not well" on another
  five.  Put the 12 pieces of paper into a bucket.
* **Step 2.** Select two groups, "pill" and "placebo", each with six random
  draws (with replacement) from the 12 pieces of paper.
* **Step 3.** Record how many "get well" in each group.
* **Step 4.** Subtract the result in group "placebo" from that in group "pill"
  (the difference may be negative).
* **Step 5.** Repeat steps 1-4 perhaps 100 times.
* **Step 6.** Compute the proportion of trials in which the pill does
  better by three or more cases.

```{python, eval=TRUE, echo=FALSE}
options = np.repeat(['get well', 'not well'], [7, 5])

n = 25
pill_cures = np.zeros(n, dtype=int)
placebo_cures = np.zeros(n, dtype=int)
pill_betters = np.zeros(n, dtype=int)

for i in range(n):
    pill = rnd.choice(options, size=6)
    pill_cures[i] = np.sum(pill == 'get well')
    placebo = rnd.choice(options, size=6)
    placebo_cures[i] = np.sum(placebo == 'get well')
    pill_betters[i] = pill_cures[i] - placebo_cures[i]

pill_trials = pd.DataFrame(
    {'Trial no': np.arange(1, n + 1),
     '# of pill cures': pill_cures,
     '# of placebo cures': placebo_cures,
     'Difference': pill_betters})
```

```{r echo=FALSE, eval=TRUE}
pill_betters <- py$pill_betters
n <- length(pill_betters)
n_gte_3 <- sum(pill_betters >= 3)
gte_3_pct <- round(n_gte_3 / n * 100)
# Show the table.
ketable(py$pill_trials,
        caption = "Results from 25 random trials for pill/placebo {#tbl-pill-trials}")
```

In the trials shown in @tbl-pill-trials, in `r int2tw(n_gte_3)` cases
(`r gte_3_pct` percent) the difference between the randomly-drawn groups is
three cases or greater. Apparently it is *somewhat* unusual — it happens 
`r gte_3_pct` percent of the time — for this universe to generate "pill"
samples in which the number of recoveries exceeds the number in the "placebo"
samples by three or more. Therefore the answer to the scientific question,
based on these samples, is that there is *some* reason to think that the
medicine does have a favorable effect. But the investigator might sensibly
await more data before reaching a firm conclusion about the pill's efficiency,
given the `r gte_3_pct` percent probability.

::: {.notebook name="pill_placebo" title="Cures for pill vs placebo"}

Now for a {{< var lang >}} solution. Again, the benchmark hypothesis is that
pill P has no effect, and we ask how often, on this assumption, the results
that were obtained from the actual test of the pill would occur by chance.

Given that in the test 7 of 12 patients overall got well, the
benchmark hypothesis assumes 7/12 to be the chances of any random
patient being cured. We generate two similar samples of 6 patients,
both taken from the same universe composed of the combined
samples — the bootstrap procedure. We count (`sum`) the number who are "get
well" in each sample. Then we subtract the number who got well in the "pill"
sample from the number who got well in the "no-pill" sample. We record the
resulting difference for each trial in the variable `pill_betters`.

In the actual test, 3 more patients got well in the sample given the pill than
in the sample given the placebo. We therefore count how many of the trials
yield results where the difference between the sample given the pill and the
sample not given the pill was greater than 2 (equal to or greater than 3). This
result is the probability that the results derived from the actual test would
be obtained from random samples drawn from a population which has a constant
cure rate, pill or no pill.

```{python}
import numpy as np
import matplotlib.pyplot as plt

rnd = np.random.default_rng()

# The bucket with the pieces of paper.
options = np.repeat(['get well', 'not well'], [7, 5])

n = 10_000

pill_betters = np.zeros(n, dtype=int)

for i in range(n):
    pill = rnd.choice(options, size=6)
    pill_cures = np.sum(pill == 'get well')
    placebo = rnd.choice(options, size=6)
    placebo_cures = np.sum(placebo == 'get well')
    pill_betters[i] = pill_cures - placebo_cures

plt.hist(pill_betters, bins=range(-6, 7))
plt.title('Number of extra cures pill vs placebo in null universe')
```

```{r}
# The bucket with the pieces of paper.
options <- rep(c('get well', 'not well'), c(7, 5))

n <- 10000

pill_betters <- numeric(n)

for (i in 1:n) {
    pill <- sample(options, size=6, replace=TRUE)
    pill_cures <- sum(pill == 'get well')
    placebo <- sample(options, size=6, replace=TRUE)
    placebo_cures <- sum(placebo == 'get well')
    pill_betters[i] <- pill_cures - placebo_cures
}

hist(pill_betters, breaks=-6:6,
     main='Number of extra cures pill vs placebo in null universe')
```

Recall our actual observed results: In the medicine group, three
more patients were cured than in the placebo group. From the
histogram, we see that in only about 8 percent of the simulated trials did the
"medicine" group do as well or better. The results seem to suggest — but by no
means conclusively — that the medicine's performance is not due to chance.
Further study would probably be warranted. The following commands added to the
above program will calculate this proportion directly:

```{python}
# How many trials gave an advantage of 3 or greater to the pill?
k = np.sum(pill_betters >= 3)
# Convert to a proportion.
kk = k / n
# Print the result.
print('Proportion with advantage of 3 or more for pill:',
      np.round(kk, 2))
```

:::
<!---
End of pill vs placebo notebook.
-->

As I (JLS) wrote when I first proposed this bootstrap method in 1969, this
method is not the standard way of handling the problem; it is not even
analogous to the standard analytic difference-of-proportions method (though
since then it has become widely accepted). Though the method shown is quite
direct and satisfactory, there are also *many other* resampling methods that
one might construct to solve the same problem. By all means, invent your own
statistics rather than simply trying to copy the methods described here; the
examples given here only illustrate the process of inventing statistics rather
than offering solutions for all classes of problems.

### Example: Did Attitudes About Marijuana Change?

<!---
Example 8.18 in W+W 1990.  Looks like fake data.
-->

Consider two polls, each asking 1500 Americans about marijuana legalization.
One poll, taken in 1980, found 52 percent of respondents in favor of
decriminalization; the other, taken in 1985, found 46 percent in favor of
decriminalization [@wonnacott1990introductory page 275]. Our null (benchmark)
hypothesis is that both samples came from the same universe (the universe made
up of the total of the two sets of observations). If so, let us then ask how
likely would be two polls to produce results as different as were observed?
Hence we construct a universe with a mean of 49 percent (the mean of the two
polls of 52 percent and 46 percent), and repeatedly draw pairs of samples of
size 1500 from it.

To see how the construction of the appropriate question is much more
challenging intellectually than is the actual mathematics, let us consider
another possibility suggested by a student: What about considering the universe
to be the earlier poll with a mean of 52 percent, and then asking the
probability that the later poll of 1500 people with a mean of 46 percent would
come from it? Indeed, on first thought that procedure seems reasonable.

Upon reflection — and it takes considerable thought on these matters to get
them right — that would *not* be an appropriate procedure. The student's
suggested procedure would be the same as assuming that we had long-run solid
knowledge of the universe, as if based on millions of observations, and then
asking about the probability of a particular sample drawn from it. That does
not correspond to the facts.

The only way to find the approach you eventually consider best — and there is
no guarantee that it is indeed *correct* — is by close reference to the
particular facts of the case.

### Example: Infarction and Cholesterol: Framingham Study {#sec-framingham-example}

It is so important to understand the logic of hypothesis tests, and
of the resampling method of doing them, that we will now tackle
another problem similar to the preceding one.

This will be the first of several problems that use data from the
famous Framingham study (drawn from Kahn and Sempos [-@kahn1989statistical])
concerning the development of myocardial infarction 16 years after the
Framingham study began, for men ages 35- 44 with serum cholesterol above 250,
compared to those with serum cholesterol below 250. The raw data are shown in
@tbl-framingham-data.  The data are from [@shurtleff1970framingham], cited in
[@kahn1989statistical, page 61, Table 3-8]. Kahn and Sempos divided the cases
into "high" and "low" cholesterol.

| Serum Cholesterol | Developed MI | Didn't Develop MI | Total |
|-------------------|--------------|-------------------|-------|
| > 250             | 10           | 125               | 135   |
| <= 250            | 21           | 449               | 470   |

: Development of Myocardial Infarction in Men Aged 35-44 After 16 Years {#tbl-framingham-data}

The statistical logic properly begins by asking: How likely is that
the two observed groups "really" came from the same "population"
with respect to infarction rates? That is, we start with this
question: How sure should one be that there is a difference in
myocardial infarction rates between the high and low-cholesterol
groups? Operationally, we address this issue by asking how likely it
is that two groups as different in disease rates as the observed
groups would be produced by the same "statistical universe."

Key step: We assume that the relevant "benchmark" or
"null hypothesis" population (universe) is the composite of the two
observed groups. That is, if there were no "true" difference in
infarction rates between the two serum-cholesterol groups, and the
observed disease differences occurred just because of sampling
variation, the most reasonable representation of the population from
which they came is the composite of the two observed groups.

Therefore, we compose a hypothetical "benchmark" universe containing
(135 + 470 =) 605 men at risk, and designate (10 + 21 =) 31 of them
as infarction cases. We want to determine how likely it is that a
universe like this one would produce — just by chance — two groups
that differ as much as do the actually observed groups. That is, how
often would random sampling from this universe produce one
sub-sample of 135 men containing a large enough number of
infarctions, and the other sub-sample of 470 men producing few
enough infarctions, that the difference in occurrence rates would be
as high as the observed difference of .029? (10/135 = .074, and
21/470 = .045, and .074 - .045 = .029).

So far, everything that has been said applies both to the conventional
formulaic method and to the "new statistics" resampling method. But the logic
is seldom explained to the reader of a piece of research — if indeed the
researcher her/ himself grasps what the formula is doing. And if one just grabs
for a formula with a prayer that it is the right one, one need never analyze
the statistical logic of the problem at hand.

Now we tackle this problem with a method that you would think of yourself if
you began with the following mind-set: How can I simulate the mechanism whose
operation I wish to understand? These steps will do the job:

* **Step 1:** Fill a bucket with 605 balls, 31 red (infarction) and the rest
  (605 — 31 = 574) green (no infarction).
* **Step 2:** Draw a sample of 135 (simulating the high serum-cholesterol
group), one ball at a time and throwing it back after it is drawn to keep the
simulated probability of an infarction the same throughout the sample; record
the number of reds. Then do the same with another sample of 470 (the low
serum-cholesterol group).
* **Step 3:** Calculate the difference in infarction rates for the two
  simulated groups, and compare it to the actual difference of .029; if the
  simulated difference is that large, record "Yes" for this trial; if not,
  record "No."
* **Step 4:** Repeat steps 2 and 3 until a total of (say) 400 or 1000
  trials have been completed. Compute the frequency with which the simulated
  groups produce a difference as great as actually observed. This frequency is
  an estimate of the probability that a difference as great as actually
  observed in Framingham would occur even if serum cholesterol has no effect
  upon myocardial infarction.

The procedure above can be carried out with balls in a bucket in a few hours.
Yet it is natural to seek the added convenience of the computer to draw the
samples. Here is a {{< var lang >}} program:

::: {.notebook name="framingham_hearts" title="Framingham heart data"}

::: nb-only
We use simulation to investigate the relationship between serum cholesterol and heart attacks in the Framingham data.
:::

```{python}
import numpy as np
import matplotlib.pyplot as plt

rnd = np.random.default_rng()

n = 10_0005
men = np.repeat(['infarction', 'no infarction'], [31, 574])

n_high = 135  # Number of men with high cholesterol
n_low = 470  # Number of men with low cholesterol

infarct_differences = np.zeros(n)

for i in range(n):
    highs = rnd.choice(men, size=n_high)
    lows = rnd.choice(men, size=n_low)
    high_infarcts = np.sum(highs == 'infarction')
    low_infarcts = np.sum(lows == 'infarction')
    high_prop = high_infarcts / n_high
    low_prop = low_infarcts / n_low
    infarct_differences[i] = high_prop - low_prop

plt.hist(infarct_differences, bins=np.arange(-0.1, 0.1, 0.005))
plt.title('Infarct proportion differences in null universe')

# How often was the resampled difference >= the observed difference?
k = np.sum(infarct_differences >= 0.029)
# Convert this result to a proportion
kk = k / n

print('Proportion of trials with difference >= observed:',
      np.round(kk, 2))
```

```{r}
n <- 10000

men <- rep(c('infarction', 'no infarction'), c(31, 574))

n_high <- 135  # Number of men with high cholesterol
n_low <- 470  # Number of men with low cholesterol

infarct_differences <- numeric(n)

for (i in 1:n) {
    highs <- sample(men, size=n_high, replace=TRUE)
    lows <- sample(men, size=n_low, replace=TRUE)
    high_infarcts <- sum(highs == 'infarction')
    low_infarcts <- sum(lows == 'infarction')
    high_prop <- high_infarcts / n_high
    low_prop <- low_infarcts / n_low
    infarct_differences[i] <- high_prop - low_prop
}

hist(infarct_differences, breaks=seq(-0.1, 0.1, by=0.005),
     main='Infarct proportion differences in null universe')

# How often was the resampled difference >= the observed difference?
k <- sum(infarct_differences >= 0.029)
# Convert this result to a proportion
kk <- k / n

message('Proportion of trials with difference >= observed: ',
        round(kk, 2))
```

The results of the test using this program may be seen in the
histogram. We find — perhaps surprisingly — that a difference as
large as observed would occur by chance around 10 percent of the
time. (If we were not guided by the theoretical expectation that
high serum cholesterol produces heart disease, we might include the
10 percent difference going in the other direction, giving a 20
percent chance). Even a ten percent chance is sufficient to call
into question the conclusion that high serum cholesterol is
dangerous. At a minimum, this statistical result should call for
more research before taking any strong action clinically or
otherwise.

:::
<!---
End of Framingham notebook.
-->

Where should one look to determine which procedures should be used to deal with
a problem such as set forth above? Unlike the formulaic approach, the basic
source is not a manual which sets forth a menu of formulas together with sets
of rules about when they are appropriate. Rather, you consult your own
understanding about what is happening in (say) the Framingham situation, and
the question that needs to be answered, and then you construct a "model" that
is as faithful to the facts as is possible. The bucket-sampling described above
is such a model for the case at hand.

To connect up what we have done with the conventional approach, one could apply
a z test (conceptually similar to the t test, but applicable to yes-no data; it
is the Normal-distribution approximation to the large binomial distribution).
Do so, we find that the results are much the same as the resampling result — an
eleven percent probability.

Someone may ask: Why do a resampling test when you can use a standard device
such as a z or t test? The great advantage of resampling is that it avoids
using the wrong method. The researcher is more likely to arrive at sound
conclusions with resampling because s/he can understand what s/he is doing,
instead of blindly grabbing a formula which may be in error.

The textbook from which the problem is drawn is an excellent one; the
difficulty of its presentation is an inescapable consequence of the formulaic
approach to probability and statistics. The body of complex algebra and tables
that only a rare expert understands down to the foundations constitutes an
impenetrable wall to understanding. Yet without such understanding, there can
be only rote practice, which leads to frustration and error.

### Example: Is One Pig Ration More Effective Than the Other? {#sec-pig-rations}

**Testing For a Difference in Means With a Two-by-Two Classification**.

Each of two new types of ration is fed to twelve pigs. A farmer wants to know
whether ration A or ration B is better.[^pig-example] The weight gains in
pounds for pigs fed on rations A and B are:

[^pig-example]: The data for this example are based on Dixon and Massey
    [-@dixon1983introduction, p. 124], who offer an orthodox method of
    handling the problem with a t-test.

A: 31, 34, 29, 26, 32, 35, 38, 34, 31, 29, 32, 31

B: 26, 24, 28, 29, 30, 29, 31, 29, 32, 26, 28, 32

The statistical question may be framed as follows: should one
consider that the pigs fed on the different rations come from the
same universe with respect to weight gains?

In the actual experiment, 9 of the 12 pigs who were fed ration A
also were in the top half of weight gains. How likely is it that one
group of 12 randomly-chosen pigs would contain 9 of the 12 top
weight gainers?

One approach to the problem is to divide the pigs into two
groups — the twelve with the highest weight gains, and the twelve
with the lowest weight gains — and examine whether an unusually
large number of high-weight-gain pigs were fed on one or the other
of the rations.

We can make this test by ordering and grouping the twenty four pigs:

* **High-weight group:** 38 (ration A), 35 (A), 34 (A), 34 (A), 32 (B), 32 (A),
  32 (A), 32 (B), 31 (A), 31 (B), 31 (A), 31 (A)
* **Low-weight group:** 30 (B), 29 (A), 29 (A), 29 (B), 29 (B), 29 (B), 28 (B),
  28 (B), 26 (A), 26 (B), 26 (B), 24 (B).

Among the twelve high-weight-gain pigs, nine were fed on ration A.
We ask: Is this further from an even split than we are likely to get
by chance? Let us take twelve red and twelve black cards, shuffle
them, and deal out twelve cards (the other twelve need not be dealt
out). Count the proportion of the hands in which one ration comes up
nine or more times in the first twelve cards, to reflect ration A's
appearance nine times among the highest twelve weight gains. More
specifically:

* **Step 1.** Constitute a deck of twelve red and twelve black cards,
  and shuffle.
* **Step 2.** Deal out twelve cards, count the number red, and record
  "yes" if there are nine or more of *either* red or black.
* **Step 3.** Repeat step 2 perhaps fifty times.
* **Step 4.** Compute the proportion "yes." This proportion estimates the
  probability sought.

```{python, eval=TRUE, echo=FALSE}
cards = np.repeat(['red', 'black'], [12, 12])
n = 25

black_nos = np.zeros(n, dtype=int)
red_nos = np.zeros(n, dtype=int)

for i in range(n):
    group_a = rnd.choice(cards, size=12)
    red_nos[i] = np.sum(group_a == 'red')
    black_nos[i] = np.sum(group_a == 'black')

rb_gte_9 = (red_nos >= 9) | (black_nos >= 9)

pig_trials = pd.DataFrame(
    {'Trial no': np.arange(1, n + 1),
     '# red': red_nos,
     '# black': black_nos,
     '>=9 red or black': np.where(rb_gte_9, '+', '')})
```

```{r echo=FALSE, eval=TRUE}
rb_gte_9 <- py$rb_gte_9
n <- length(rb_gte_9)
n_rb_gte_9 <- sum(rb_gte_9)
rb_gte_9_pct <- round(n_rb_gte_9 / n * 100)
# Show the table.
ketable(py$pig_trials,
        caption = "Results from 25 random trials for pig rations {#tbl-pig-trials}")
```

@tbl-pig-trials shows the results of 25 trials. In
`r int2tw(n_rb_gte_9)` (marked by + signs) of the 25 (that is,
`r rb_gte_9_pct` percent of the trials) there were nine or more either red or
black cards in the first twelve cards. Again the results suggest that it would
be *slightly unusual* for the results to favor one ration or the other so
strongly just by chance if they come from the same universe.

Now the {{< var lang >}} procedure to answer the question:

::: {.notebook name="pig_rations" title="Weight gain on pig rations"}

::: nb-only
We do a simulation of weight gain ranks for two different pig rations.
:::

The [`ranks <- 1:24`]{.r}[`ranks = np.arange(1, 25)`]{.python}
statement creates {{< var an_array >}} of numbers 1 through 24, which will
represent the rankings of weight gains for each of the 24 pigs. We repeat the
following procedure for 10000 trials. First we shuffle the elements of {{< var
array >}} `ranks` so that the rank numbers for weight gains are randomized and
placed in {{< var array >}} `shuffled`. We then select the first 12 elements of
`shuffled` and place them in `first_12`; this represents the rankings of
a randomly-selected group of 12 pigs. We next count (`sum`) in `n_top` the
number of pigs whose rankings for weight gain were in the top half — that is,
a rank of less than 13. We record that number in `top_ranks`, and then continue
the loop, until we finish our `n` trials.

Since we did not know beforehand the direction of the effect of ration A on
weight gain, we want to count the times that *either more than 8* of the random
selection of 12 pigs were in the top half of the rankings, *or that fewer than
4* of these pigs were in the top half of the weight gain rankings — (The latter
is the same as counting the number of times that more than 8 of the 12
*non-selected* random pigs were in the top half in weight gain.)

We do so with the final two `sum` statements. By adding the two results `n_gte_9` and
`n_lte_3` together, we have the number of times out of 10,000 that differences
in weight gains in two groups as dramatic as those obtained in the actual
experiment would occur by chance.

```{python}
import numpy as np
import matplotlib.pyplot as plt

rnd = np.random.default_rng()

# Constitute the set of the weight gain rank orders. ranks is now a vector
# consisting of the numbers 1 — 24, in that order.
ranks = np.arange(1, 25)

n = 10_000

top_ranks = np.zeros(n, dtype=int)

for i in range(n):
    # Shuffle the ranks of the weight gains.
    shuffled = rnd.permuted(ranks)
    # Take the first 12 ranks.
    first_12 = shuffled[:12]
    # Determine how many of these randomly selected 12 ranks are less than
    # 12 (i.e. 1-12), put that result in n_top.
    n_top = np.sum(first_12 <= 12)
    # Keep track of each trial result in top_ranks
    top_ranks[i] = n_top

plt.hist(top_ranks, bins=np.arange(1, 12))
plt.title('Number of top 12 ranks in pig-ration trials')
```

```{r}
# Constitute the set of the weight gain rank orders. ranks is now a vector
# consisting of the numbers 1 — 24, in that order.
ranks <- 1:24

n <- 10000

top_ranks <- numeric(n)

for (i in 1:n) {
    # Shuffle the ranks of the weight gains.
    shuffled <- sample(ranks)
    # Take the first 12 ranks.
    first_12 <- shuffled[1:12]
    # Determine how many of these randomly selected 12 ranks are less than
    # 12 (i.e. 1-12), put that result in n_top.
    n_top <- sum(first_12 <= 12)
    # Keep track of each trial result in top_ranks
    top_ranks[i] <- n_top
}

hist(top_ranks, breaks=1:11,
     main='Number of top 12 ranks in pig-ration trials')
```

We see from the histogram that, in about 3 percent of the trials, either more
than 8 or fewer than 4 top half ranks (1-12) made it into the random group of
twelve that we selected. {{< var lang >}} will calculate this for us as
follows:

```{python}
# Determine how many of the trials yielded 9 or more top ranks.
n_gte_9 = np.sum(top_ranks >= 9)
# Determine how many trials yielded 3 or fewer of the top ranks.
# If there were 3 or fewer, then 9 or more of the top ranks must
# have been in the other group (not selected).
n_lte_3 = np.sum(top_ranks <= 3)
# Add the two together.
n_both = n_gte_9 + n_lte_3
# Convert to a proportion.
prop_both = n_both / n

print('Trial proportion >=9 top ranks in either group:',
      np.round(prop_both, 2))
```

```{r}
# Determine how many of the trials yielded 9 or more top ranks.
n_gte_9 <- sum(top_ranks >= 9)
# Determine how many trials yielded 3 or fewer of the top ranks.
# If there were 3 or fewer, then 9 or more of the top ranks must
# have been in the other group (not selected).
n_lte_3 <- sum(top_ranks <= 3)
# Add the two together.
n_both <- n_gte_9 + n_lte_3
# Convert to a proportion.
prop_both <- n_both / n

message('Trial proportion >=9 top ranks in either group: ',
        round(prop_both, 2))
```

The decisions that are warranted on the basis of the results depend
upon one's purpose. If writing a scientific paper on the merits of
ration A is the ultimate purpose, it would be sensible to test
another batch of pigs to get further evidence. (Or you could proceed
to employ another sort of test for a slightly more precise
evaluation.) But if the goal is a decision on which type of ration
to buy for a small farm and they are the same price, just go ahead
and buy ration A because, even if it is no better than ration B, you
have strong evidence that it is *no worse* .

:::
<!---
End of pig-ration notebook.
-->

### Example: Do Planet Densities Differ?

**Consider the five planets known to the ancient world.**

Mosteller and Rourke [-@mosteller1973sturdy, pages 17--19] ask us to compare
the densities of the three planets *farther* from the sun than is the earth
(Mars, density 0.71; Jupiter, 0.24; and Saturn, 0.12) against the densities of
the planets *closer* to the sun than is the earth (Mercury, 0.68; Venus, 0.94).

The average density of the distant planets is .357, of the closer planets is
.81. Is this difference (.353) statistically surprising, or is it likely to
occur in a chance ordering of these planets?

We can answer this question with a permutation test; such sampling without
replacement makes sense here because we are considering the entire set of
planets, rather than a sample drawn from a larger population of planets (the
word "population" is used here, rather than "universe," to avoid confusion.)
And because the number of objects is so small, one could examine all possible
arrangements (permutations), and see how many have (say) differences in mean
densities between the two groups as large as observed.

Another method that Mosteller and Rourke suggest is by a comparison of the
density *ranks* of the two sets, where Saturn has rank 1 and Venus has rank 5.
This might have a scientific advantage if the sample data are dominated by
a single "outlier," whose domination is removed when we rank the data.

We see that the sum of the ranks for the "closer" set is 3+5=8. We can then
ask: If the ranks were assigned at random, how likely is it that a set of two
planets would have a sum as large as 8? Again, because the sample is small, we
can examine all the possible permutations, as Mosteller and Rourke do in Table
3-1 [@mosteller1973sturdy, page 56] (Substitute "Closer" for "B," "Further" for
"A"). In two of the ten permutations, a sum of ranks as great as 8 is observed,
so the probability of a result as great as observed happening by chance is 20
percent, using these data. (We could just as well consider the difference in
mean ranks between the two groups: (8/2 - 7/3 = 10 / 6 = 1.67).

<!---
Perhaps easier to reproduce M+R table here - no-one will have that reference to
hand.
-->

To illuminate the logic of this test, consider comparing the heights of two
samples of trees. If sample A has the five tallest trees, and sample B has the
five shortest trees, the difference in mean ranks will be (6+7+8+9+10=) 40
— (1+2+3+4+5=) 15, the largest possible difference. If the groups are less
sharply differentiated — for example, if sample A has #3 and sample B has #8
— the difference in ranks will be less than the maximum of 40, as you can
quickly verify.

The method we have just used is called a Mann-Whitney test, though that label
is usually applied when the data are too many to examine all the possible
permutations; in that case one conventionally uses a table prepared by formula.
In the case where there are too many for a complete permutation test, our
resampling algorithm is as follows (though we'll continue with the planets
example):

1.  Compute the mean ranks of the two groups.
2.  Calculate the difference between the means computed in step 1.
3.  Create a bucket containing the ranks from 1 to the number of
    observations (5, in the case of the planets)
4.  Shuffle the ranks.
5.  Since we are working with the ranked data, we must draw without
    replacement, because there can only be one #3, one #7, and so on. So draw
    the number of observations corresponding to the number of observations
    — 2 "Closer" and 3 "Further."
6.  Compute the mean ranks of the two simulated groups of planets.
7.  Calculate the difference between the means computed in step 5
    and record.
8.  Repeat steps 4 through 7 perhaps 1000 times.
9.  Count how often the shuffled difference in ranks exceeds the
    observed difference from step 2 (1.67).

::: {.notebook name="planet_densities" title="Planet densities and distance"}

::: nb-only
We apply the logic of resampling to the problem of close and distant planets
and their densities.
:::

```{python}
import numpy as np

rnd = np.random.default_rng()

# Steps 1 and 2.
actual_mean_diff = 8 / 2 - 7 / 3

# Step 3
ranks = np.arange(1, 6)

n = 10_000

mean_differences = np.zeros(n)

for i in range(n):
    # Step 4
    shuffled = rnd.permuted(ranks)
    # Step 5
    closer = shuffled[:2]  # First 2
    further = shuffled[2:] # Last 3
    # Step 6
    mean_close = np.mean(closer)
    mean_far = np.mean(further)
    # Step 7
    mean_differences[i] = mean_close - mean_far

# Step 9
k = np.sum(mean_differences >= actual_mean_diff)
prob = k / n

print('Proportion of trials with mean difference >= 1.67:',
      np.round(prob, 2))
```

```{r}
# Steps 1 and 2.
actual_mean_diff <- 8 / 2 - 7 / 3

# Step 3
ranks <- 1:5

n <- 10000

mean_differences <- numeric(n)

for (i in 1:n) {
    # Step 4
    shuffled <- sample(ranks)
    # Step 5
    closer <- shuffled[1:2]  # First 2
    further <- shuffled[3:5] # Last 3
    # Step 6
    mean_close <- mean(closer)
    mean_far <- mean(further)
    # Step 7
    mean_differences[i] <- mean_close - mean_far
}

# Step 9
k <- sum(mean_differences >= actual_mean_diff)
prob <- k / n

message('Proportion of trials with mean difference >= 1.67: ',
        round(prob, 2))
```

Interpretation: `r round(get_var('prob') * 100)` percent of the time, random
shufflings produced a difference in ranks as great as or greater than observed.
Hence, on the strength of this evidence, we should *not* conclude that there is
a statistically surprising difference in densities between the further planets
and the closer planets.

:::
<!---
End of planets notebook.
-->

## Conclusion

This chapter has begun the actual work of testing hypotheses. The next chapter
continues with discussion of somewhat more complex problems with counted data
— more complex to think about, but no more difficult to actually treat
mathematically with resampling simulation. If you have understood the general
logic of the procedures used up until this point, you are in command of all the
necessary conceptual knowledge to construct your own tests to answer any
statistical question. A lot more practice, working on a variety of problems,
obviously would help. But the key elements are simple: 1) Model the real
situation accurately, 2) experiment with the model, and 3) compare the results
of the model with the observed results.
