---
jupyter:
  jupytext:
    metadata_filter:
      notebook:
        additional: all
        excluded:
        - language_info
    text_representation:
      extension: .Rmd
      format_name: rmarkdown
      format_version: '1.0'
      jupytext_version: 0.8.6
  kernelspec:
    display_name: Python 3
    language: python
    name: python3
resampling_with:
    ed2_fname: 23-Chap-19
---

```{r setup, include=FALSE}
source("_common.R")
```

# General Procedures for Testing Hypotheses {#sec-testing-procedures}

:::{.callout-warning}
## Draft page partially ported from original PDF

This page is an automated and partial import from the [original second-edition
PDF](https://resample.com/content/text/23-Chap-19.pdf).

We are in the process of updating this page for formatting, and porting any
code from the original [RESAMPLING-STATS
language](http://www.statistics101.net) to Python and R.

Feel free to read this version for the sense, but expect there to be multiple
issues with formatting.

We will remove this warning when the page has adequate formatting, and we have
ported the code.
:::


## Introduction

The previous chapters have presented procedures for making statistical
inferences that apply to both testing hypotheses and constructing
confidence intervals: This chapter focuses on specific procedures for
testing hypotheses.

\`The general idea in testing hypotheses is to ask: Is there some
*other* universe which might well have produced the observed sample? So
we consider alternative hypotheses. This is a straightforward exercise
in probability, asking about behavior of one or more universes. The
choice of another universe(s) to examine depends upon purposes and other
considerations.

## Canonical question-and-answer procedure for testing hypotheses

## Skeleton procedure for testing hypotheses

**Akin to skeleton procedure for questions in probability and confidence
intervals shown elsewhere**

The following series of questions will be repeated below in the context
of a specific inference.

*What is the question? What is the purpose to be served by answering the
question?*

*Is this a "probability" or a "statistics" question?*

**Assuming the Question is a Statistical Inference Question**

*What is the form of the statistics question?*

Hypothesis test, or confidence interval, or other inference? One must
first decide whether the conceptual-scientific question is of the form
a) a test about the probability that some sample is likely to happen by
chance rather than being very surprising (a test of a hypothesis), or b)
a question about the accuracy of the estimate of a parameter of the
population based upon sample evidence (a confidence interval):

**Assuming the Question Concerns Testing Hypotheses**

*Will you state the costs and benefits of various outcomes, perhaps in
the form of a "loss function"? If "yes," what are they?*

*How many samples of data have been observed?*

One, two, more than two?

*What is the description of the observed sample(s)?*

Raw data?

*Which characteristic(s) (parameters) of the population are of interest
to you?*

*What are the statistics of the sample(s) that refer to this (these)
characteristics(s) in which you are interested?*

*What comparison(s) to make?*

Samples to each other?

Sample to particular universe(s)? If so, which?

*What is the benchmark (null) universe?*

This may include presenting the raw data and/or such summary statistics
as the computed mean, median, standard deviation, range, interquartile
range, other:

*If there is to be a Neyman-Pearson-type alternative universe, what is
it?* (In most cases the answer to this technical question is "no.")

*Which symbols for the observed entities?*

Discrete or continuous?

What values or ranges of values?

*Which sample(s) do you wish to compare to which, or to the null
universe (and perhaps to the alternative universe)?* (Answer: samples
the same size as has been observed)

\[Here one may continue with the conventional method, using perhaps a
*t* or *f* or chi-square test or whatever: Everything up to now is the
same whether continuing with resampling or with standard parametric
test.\]

*What procedure will be used to produce the resampled entities?*

Randomly drawn?

Simple (single step) or complex (multiple "if" drawings)?

*What procedure to produce resample?*

Which universe will you draw them from? With or without replacement?

What size resamples? Number of resample trials?

*What to record as outcome of each resample trial?*

Mean, median, or whatever of resample?

**Classifying the outcomes**

*What is the criterion of significance to be used in evaluating the
results of the test?*

**Stating the distribution of results**

Graph of each statistic recorded — occurrences for each value.

Count the outcomes that exceed criterion and divide by number of trials.

## An example: can the bio-engineer increase the female calf rate?

*The question.* (from [@hodges1970basic, page 310]: Female calves are more
valuable than male calves. A bio-engineer claims to have a method that
can produce more females. He tests the procedure on ten of your pregnant
cows, and the result is nine females. Should you believe that his method
has some effect? That is, what is the probability of a result this
surprising occurring by chance?

*The purpose:* Female calves are more valuable than male.

*Inference?* Yes.

*Test of hypothesis?* Yes.

*Will you state the costs and benefits* of various outcomes (or a loss
function)? We need only say that the benefits of a method that works are
very large, and if the results are promising, it is worth gathering more
data to confirm results.

*How many samples of data* are part of the significance test? One

*What is the size of the first sample* about which you wish to make
significance statements? Ten.

*What comparison(s) to make?* Compare sample to benchmark universe.

*What is the benchmark universe* that embodies the null hypothesis?
50-50 female, or 100/206 female.

*If there is to be a Neyman-Pearson alternative universe* , what is it?
None.

*Which symbols for the observed entities?* Balls in bucket, or numbers.

*What values or ranges of values?* 0-1, (1-100), or 101-206.

*Finite or infinite?* Infinite.

*Which sample(s) do you wish to compare* to which, or to the null
universe (and perhaps to the alternative universe)? Ten calves compared
to universe.

*What procedure to produce entities?* Sampling with replacement,

*Simple (single step) or complex (multiple "if" drawings)?* One can
think of it either way.

*What to record as outcome of each resample trial?* The proportion (or
number) of females.

*What is the criterion* to be used in the test? The probability that in
a sample of ten calves, nine (or more) females would be drawn by chance
from the benchmark universe of half females. (Or frame in terms of a
significance level.)

*"One-tail" or "two-tail" test?* One tail, because the farmer is only
interested in females: Finding a large proportion of males would not be
of interest, and would not cause one to reject the null hypothesis.

*Computation of the probability sought.* The actual computation of
probability may be done with several formulaic or sample-space methods,
and with several resampling methods: I will first show a resampling
method and then several conventional methods. The following material, which
allows one to compare resampling and conventional methods, is more germane to
the earlier explication of resampling taken altogether in earlier chapters
than it is to the theory of hypothesis tests discussed in this chapter, but it
is more expedient to present it here.

## Computation of Probabilities with Resampling

We can do the problem by hand as follows:

1.  Constitute a bucket with either one blue and one pink ball, or 106
    blue and 100 pink balls.
2.  Draw ten balls with replacement, count pinks, and record.
3.  Repeat step (2) say 400 times.
4.  Calculate proportion of results with 9 or 10 pinks.

Or, we can take advantage of the speed and efficiency of the computer as
follows:

```{python}
import numpy as np
import matplotlib.pyplot as plt

rnd = np.random.default_rng()

n = 10000

females = np.zeros(n)

for i in range(n):
    samp = rnd.choice(['female', 'male'], size=10, replace=True)
    females[i] = np.sum(samp == 'female')

plt.hist(females, bins='auto')

k = np.sum(females >= 9)
kk = k / n
print('Proportion with >= 9 females:', kk)
```

```{r}
n <- 10000

females <- numeric(n)

for (i in 1:n) {
    samp <- sample(c('female', 'male'), size=10, replace=TRUE)
    females[i] <- sum(samp == 'female')
}

hist(females)

k <- sum(females >= 9)
kk <- k / n
message('Proportion with >= 9 females: ', kk)
```

This outcome implies that there is roughly a one percent chance that one
would observe 9 or 10 female births in a single sample of 10 calves if
the probability of a female on each birth is .5. This outcome should
help the decision-maker decide about the plausibility of the
bio-engineer's claim to be able to increase the probability of female
calves being born.

## Conventional methods

### The Sample Space and First Principles

Assume for a moment that our problem is a smaller one and therefore much
easier — the probability of getting two females in two calves if the
probability of a female is .5. One could then map out what
mathematicians call the "sample space," a technique that (in its
simplest form) assigns to each outcome a single point, and find the
proportion of points that correspond to a "success." We list all four
possible combinations — FF, FM, MF, MM. Now we look at the ratio of the
number of combinations that have 2 females to the total, which is 1/4.
We may then interpret this probability.

We might also use this method for (say) *five* female calves *in a row*. We can
make a list of possibilities such as FFFFF, MFFFF, MMFFF, MMMFFF ... MFMFM ...
MMMMM. There will be 2\*2\*2\*2\*2 = 32 possibilities, and 64 and 128
possibilities for six and seven calves respectively. But when we get as high as
ten calves, this method would become very troublesome.

### Sample Space Calculations

For two females in a row, we could use the well known, and very simple,
multiplication rule; we could do so even for ten females in a row. But
calculating the probability of nine females in ten is a bit more
complex.

### Pascal's Triangle

One can use Pascal's Triangle to obtain binomial coefficients for p = .5
and a sample size of 10, focusing on those for 9 or 10 successes. Then
calculate the *proportion of the total cases* with 9 or 10 "successes"
in one direction, to find the proportion of cases that pass beyond the
criterion of 9 females. The method of Pascal's Triangle requires more
complete understanding of the probabilistic system than does the
resampling simulation described above because Pascal's Triangle requires
that one understand the entire structure; simulation requires only that
you follow the rules of the model.

### The Quincunx

The quincunx — a device that filters tiny balls through a set of bumper points
not unlike a pinball machine, mentioned here simply for completeness — is more
a simulation method than theoretical, but it may be considered "conventional."
Hence, it is included here.

### Table of Binomial Coefficients

Pascal's Triangle becomes cumbersome or impractical with large
numbers — say, 17 females of 20 births — or with probabilities other
than .5. One might produce the binomial coefficients by algebraic
multiplication, but that, too, becomes tedious even with small sample
sizes. One can also use the pre-computed table of binomial coefficients
found in any standard text. But the probabilities for n = 10 and 9 or 10
females are too small to be shown.

### Binomial Formula

For larger sample sizes, one can use the binomial formula. The binomial
formula gives no deeper understanding of the statistical structure than
does the Triangle (but it does yield a deeper understanding of the pure
mathematics). With very large numbers, even the binomial formula is
cumbersome.

### The Normal Approximation

When the sample size becomes too large for any of the above methods, one can
then use the Normal approximation, which yields results close to the binomial
(as seen very nicely in the output of the quincunx). But use of the Normal
distribution requires an estimate of the standard deviation, which can be
derived either by formula or by resampling. (See a more extended parallel
discussion in @sec-confidence-two-approaches on confidence intervals for the
Bush-Dukakis comparison.)

The desired probability can be obtained from the Z formula and a standard table
of the Normal distribution found in every elementary text.

The Z table can be made less mysterious if we generate it with
simulation, or with graph paper or Archimedes' method, using as raw
material (say) five "continuous" (that is, non-binomial) distributions,
many of which are skewed: 1) Draw samples of (say) 50 or 100. 2) Plot
the means to see that the Normal shape is the outcome. Then 3)
standardize with the standard deviation by marking the standard
deviations onto the histograms.

The aim of the above exercise and the heart of the conventional
parametric method is to compare the sample result — the mean — to a
standardized plot of the means of samples drawn from the universe of
interest to see how likely it is that that universe produces means
deviating as much from the universe mean as does our observed sample
mean. The steps are:

1.  Establish the Normal shape — from the exercise above, or from the
    quincunx or Pascal's Triangle or the binomial formula or the formula
    for the Normal approximation or some other device.
2.  Standardize that shape in standard deviations.
3.  Compute the Z score for the sample mean — that is, its deviation
    from the universe mean in standard deviations.
4.  Examine the Normal (or really, tables computed from graph paper,
    etc.) to find the probability of a mean deviating that far by
    chance.

This is the canon of the procedure for most parametric work in
statistics. (For some small samples, accuracy is improved with an
adjustment.)

## Choice of the benchmark universe[^bruce]

[^bruce]: This is one of many issues that Peter Bruce first raised, and whose
    treatment here reflects back-and-forth discussion between us.

In the example of the ten calves, the choice of a benchmark universe
— a universe that (on average) produces equal proportions of males and females
— seems rather straightforward and even automatic, requiring no difficult
judgments. But in other cases the process requires more judgments.

Let's consider another case where the choice of a benchmark universe
requires no difficult judgments. Assume the U.S. Department of Labor's
Bureau of Labor Statistics (BLS) takes a very large sample — say, 20,000
persons — and finds a 10 percent unemployment rate. At some later time another
but smaller sample is drawn — 2,000 persons — showing an 11 percent
unemployment rate. Should BLS conclude that unemployment has risen, or is there
a large chance that the difference between 10 percent and 11 percent is due to
sample variability? In this case, it makes rather obvious sense to ask how
often a sample of 2,000 drawn from a universe of 10 percent unemployment
(ignoring the variability in the larger sample) will be as different as 11
percent due solely to sample variability? This problem differs from that of the
calves only in the proportions and the sizes of the samples.

Let's change the facts and assume that a very large sample had not been
drawn and only a sample of 2,000 had been taken, indicating 11 percent
unemployment. A policy-maker asks the probability that unemployment is
above ten percent. It would still seem rather straightforward to ask how
often a universe of 10 percent unemployment would produce a sample of
2000 with a proportion of 11 percent unemployed.

Still another problem where the choice of benchmark hypothesis is
relatively straightforward: Say that BLS takes two samples of 2000
persons a month apart, and asks whether there is a difference in the
results. Pooling the two samples and examining how often two samples
drawn from the pooled universe would be as different as observed seems
obvious.

One of the reasons that the above cases — especially the two-sample
case — seem so clear-cut is that the variance of the benchmark hypothesis is
not an issue, being implied by the fact that the samples deal with proportions.
If the data were continuous, however, this issue would quickly arise. Consider,
for example, that the BLS might take the same sorts of samples and ask
unemployed persons the *lengths of time* they had been unemployed. Comparing
a small sample to a very large one would be easy to decide about. And even
comparing two small samples might be straightforward — simply pooling them as
is.

But what about if you have a sample of 2,000 with data on lengths of
unemployment spells with a mean of 30 days, and you are asked the
probability that it comes from a universe with a mean of 25 days? Now
there arises the question about the amount of variability to assume for
that benchmark universe. Should it be the variability observed in the
sample? That is probably an overestimate, because a universe with a
smaller mean would probably have a smaller variance, too. So some
judgment is required; there cannot be an automatic "objective" process
here, whether one proceeds with the conventional or the resampling
method.

The example of the comparison of liquor retailing systems in
@sec-liquor-public-measured provides more material on this subject.

## Why is statistics — and hypothesis testing — so difficult?

Why is statistics such a difficult subject? The aforegoing procedural
outline provides a window to the explanation. Hypothesis testing — as is
also true of the construction of confidence intervals (but unlike simple
probability problems) — involves a very long chain of reasoning, perhaps
longer than in any other realm of systematic thinking. Furthermore, many
decisions in the process require judgment that goes beyond technical
analysis. All this emerges as one proceeds through the skeleton
procedure above with any specific example.

(Bayes' rule also is very difficult intuitively, but that probably is a
result of the twists and turns required in all complex problems in
conditional probability. Decision-tree analysis is counter-intuitive,
too, probably because it starts at the end instead of the beginning of
the story, as we are usually accustomed to doing.)
